---
title: "Differential Abundance Analysis with Speckle"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

DISCLAMIR:
I am summarizing my code here, so there might be typos or copy paste errors. This is more like a detailed index page.


```{r eval=FALSE,}
r3dcol <- list()
r3dcol$cols_94 <- c(
  "#F0E442","#1CE6FF","#FF34FF","#FF4A46","#008941","#006FA6","#A30059","#FFDBE5","#7A4900","#0089A3",
  "#63FFAC","#B79762",
  "#004D43","#8FB0FF","#997D87","#809693","#6A3A4C","#b4b701","#4FC601","#3B5DFF","#FF2F80","#61615A",
  "#BA0900","#6B7900",
  "#00C2A0","#FFAA92","#FF90C9","#B903AA","#D16100","#922329","#7B4F4B","#A1C299","#0AA6D8","#00846F",
  "#FFB500","#C2FFED",
  "#A079BF","#CC0744","#C0B9B2","#C2FF99","#00489C","#6F0062","#0CBD66","#EEC3FF","#456D75","#B77B68",
  "#7A87A1","#788D66",
  "#885578","#FAD09F","#FF8A9A","#D157A0","#BEC459","#456648","#0086ED","#886F4C","#34362D","#B4A8BD",
  "#00A6AA","#452C2C",
  "#636375","#A3C8C9","#FF913F","#938A81","#575329","#00FECF","#B05B6F","#8CD0FF","#3B9700","#04F757",
  "#C8A1A1","#1E6E00",
  "#7900D7","#A77500","#6367A9","#A05837","#6B002C","#772600","#D790FF","#9B9700","#549E79","#FFF69F",
  "#201625","#72418F",
  "#BC23FF","#99ADC0","#3A2465","#DDEFFF","#5B4534","#FDE8DC","#404E55","#CB7E98","#A4E804","#324E72"
)
r3dcol$cols_46 <- c(
  "#FFB500","#00C2A0","#D157A0","#8CD0FF","#FF4A46","#FDE8DC","#63FFAC","#B79762","#8FB0FF","#997D87",
  "#809693","#b4b701",
  "#4FC601","#FFAA92","#FF90C9","#A1C299","#F0E442","#C2FFED","#A079BF","#C0B9B2","#C2FF99","#EEC3FF",
  "#B77B68","#7A87A1",
  "#788D66","#FAD09F","#FF8A9A","#BEC459","#B4A8BD","#A3C8C9","#FF913F","#938A81","#00FECF","#FF34FF",
  "#1CE6FF","#04F757",
  "#C8A1A1","#D790FF","#9B9700","#549E79","#FFF69F","#99ADC0","#DDEFFF","#FFDBE5","#CB7E98","#A4E804"
)
r3dcol$cols_pjsala <- c(
  "#005579","#0F4A9C","#139992","#1A1A1A","#354E23","#3F84AA","#532C8A","#635547","#647a4f","#65A83E",
  "#7F6874","#8870ad",
  "#8DB5CE","#8EC792","#989898","#9e6762","#B51D8D","#c19f70","#C3C388","#C594BF","#C72228","#c9a997",
  "#C9EBFB","#cc7818",
  "#CDE088","#DABE99","#DFCDE4","#EF4E22","#EF5A9D","#F397C0","#F6BFCB","#f79083","#f7f79e","#f9decf",
  "#FACB12","#FBBE92",
  "#ff891c"
)
r3dcol$cols_4x3 <- c(
  "#551153","#970098","#9c24cd","#412c00","#997600","#c6a000","#1e3604","#355e07","#4f890e","#002f64",
  "#0050ae","#1077f3"
)
r3dcol$cols_12x1 <- c(
  '#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a',
  '#ffff99','#b15928'
)
r3dcol$cols_4x3b <- c(
  "#551153","#970098","#cc34cd","#c85b00","#f98517","#fdbf6f","#00603d","#008c5c","#33b983","#002f64",
  "#0050ae","#1077f3"
)
r3dcol$cols_4x1 <- c(
  "#970098","#997600","#4f890e","#0050ae"
)
r3dcol$cols_4x1b <- c(
  "#970098","#f98517","#008c5c","#1077f3"
)
r3dcol$cols_3a <- c(
  "#D157A0","#FFB500","#00C2A0"
)
r3dcol$cols_3b <- c(
  "#D454B5","#c85b00","#008c5c"
)
```

```{r eval=FALSE}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(miloR)
  library(dplyr)
  library(scater)
  library(speckle)
  library(ggplot2)
})
```

# Introduction

Here we will summarize the steps of the Compositional Analysis / Differential Abundance Analysis between KO clones and
NT-Control (NonTargetting Control). We will do it in two different analysis 1) cluster levels with *Speckle* and 2)
k-nearstneighborhood with smaller groups of cells.

# Data Preprocessing

## Loading Data


```{r eval=FALSE}
cmo <- readRDS("~/cmo_chapter/introduction/rds/cmoy2.RDS")
Idents(cmo) <- "Conditions_lv"

# subsetting with each KO and Control
cmo_CA <- subset(cmo, idents =c("Control","Atf3KO"))
cmo_BA <- subset(cmo, idents =c("Control","Bcl6bKO"))
cmo_ZA <- subset(cmo, idents =c("Control","Zfp711KO"))

# we have to clean the levels for with following code
clean_metadata <- function(seurat_obj) {
  metadata <- seurat_obj@meta.data
  # Apply droplevels to each factor column
  metadata[] <- lapply(metadata, function(column) {
    if (is.factor(column)) {
      return(droplevels(column))
    } else {
      return(column)
    }
  })
  seurat_obj@meta.data <- metadata
  return(seurat_obj)
}

cmo_CA <- clean_metadata(cmo_CA)
cmo_BA <- clean_metadata(cmo_BA)
cmo_ZA <- clean_metadata(cmo_ZA)
```

I will show only one condition (Atf3-KO versus NT-Control).
We will first follow the classical "speckle" pipeline during the analysis. Later I will subset the samples to equal cell numbers and use the plotCellTypeProps function to compare KO and NT conditions at annotation level. 


# Atf3

## Loading the RDS file

```{r}
cmo_CA <- readRDS("../../../RDS/CMO/cmo_CA.RDS")
```

## just to be sure
if you are subsetting the data and loosing some of the levels use droplevels() before running speckle and miloR
e.g.
```{r}
cmo_CA$Conditions_lv <- droplevels(cmo_CA$Conditions_lv)
```



```{r}
c_ca_4 <- levels(cmo_CA$cl_4) # ordered subcluster-cell cycle annotation (jus t for fun)
c_ca_3 <- levels(cmo_CA$cl_3) # ordered subcluster annotation
c_ca_2 <- levels(cmo_CA$cl_2) # ordered cluster annotation
c_ca_1 <- levels(cmo_CA$cl_1) #  ordered group annotation
c_ca_cc <- levels(cmo_CA$Phasel) # ordered cell cycle annotation
c_ca_conditions <- levels(cmo_CA$Conditions_lv)  # Conditions (Atf3-KO and NT-Control)
c_ca_samples <- levels(cmo_CA$Samples_lv) # Replicates
```

```{r}
Idents(cmo_CA) <- "cl_3"
```

# Propeller

we will do DAA the test now in 5 levels (cl_1, ...) in both "asin" and "logit" tranformation. You can find the results in the supplementary table.


### Asin&logit



```{r}
prop_CA <- list()
prop_CA$S1a <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S2a <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S3a <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S4a <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S5a<- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")

prop_CA$S1l <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S2l <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S3l <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S4l <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S5l<- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")

```


```{r}
ca1_asin <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
ca2_asin <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
ca3_asin <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
cacc_asin <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
```

```{r}
ca1_logit <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
ca2_logit <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
ca3_logit <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
cacc_logit <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
```


```{r}
write.table(ca1_asin, "./Atf3/Atf3_propeller_groups_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_asin, "./Atf3/Atf3_propeller_clusters_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_asin, "./Atf3/Atf3_propeller_subclusters_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_asin, "./Atf3/Atf3_propeller_cellcycle_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
```

```{r}
write.table(ca1_logit, "./Atf3/Atf3_propeller_groups_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_logit, "./Atf3/Atf3_propeller_clusters_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_logit, "./Atf3/Atf3_propeller_subclusters_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_logit, "./Atf3/Atf3_propeller_cellcycle_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
```


```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_1 <-
  plotCellTypeProps(clusters = cmo_CA$cl_1, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Groups", 
    values = c("#66c2a5","#fc8d62","#8da0cb"),
    guide = guide_legend(ncol = 1)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_1)
```

```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_2 <-
  plotCellTypeProps(clusters = cmo_CA$cl_2, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Clusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 1)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_2)
```

```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_3 <-
  plotCellTypeProps(clusters = cmo_CA$cl_3, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Subclusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 2)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_3)
```

```{r,fig.width=14, fig.height=5.5}
plot_cell_type_normal_4 <-
  plotCellTypeProps(clusters = cmo_CA$cl_4, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "subclusters-cc (A-Z)", 
    values = c(r3dcol$cols_94,r3dcol$cols_46),
    guide = guide_legend(ncol = 6)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_4)
```


# equal cell numbers & filipped axis on plotCellTypeProps

In order to have each sample has equal cell number, we will take equal number of cells from each sample.

```{r}
Idents(cmo_CA) <- "Samples_lv"
```

```{r}
table(Idents(cmo_CA))
min(table(Idents(cmo_CA)))
```

```{r}
 # Downsample the number of cells per identity class
subset_cmo_CA <-subset(x = cmo_CA, downsample = 4954)
table(Idents(subset_cmo_CA))
```

```{r}
levels(subset_cmo_CA)
Idents(subset_cmo_CA) <- "Conditions"
levels(subset_cmo_CA)
```
```{r}
new.cluster.ids <- c("b_Atf3KO","a_Control")
names(new.cluster.ids) <- levels(subset_cmo_CA)
subset_cmo_CA <- RenameIdents(subset_cmo_CA, new.cluster.ids)
subset_cmo_CA$conditionsxx <- Idents(subset_cmo_CA)
```

```{r,fig.width=4.5, fig.height=7}
# Load necessary libraries
pp1 <-
  plotCellTypeProps(clusters = subset_cmo_CA$conditionsxx, sample = subset_cmo_CA$cl_1) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 10, margin = margin(l = 1)),
    legend.title = element_text(size = 0, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l = -10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_1) +  
  scale_fill_manual(values = c( "#009E73","#E69F00"), breaks = c("a_Control", "b_Atf3KO"))+geom_hline(yintercept = 0.5,linetype=2,color='red')+NoLegend()
pp1
```


```{r,fig.width=9, fig.height=7}
# Load necessary libraries
pp2 <-
  plotCellTypeProps(clusters = subset_cmo_CA$conditionsxx, sample = subset_cmo_CA$cl_2) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 10, margin = margin(l = 1)),
    legend.title = element_text(size = 0, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l = -10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_2) +  
  scale_fill_manual(values = c( "#009E73","#E69F00"))+geom_hline(yintercept = 0.5,linetype=2,color='red')+NoLegend()
pp2
```


```{r,fig.width=9, fig.height=7}
# Load necessary libraries
pp3 <-
  plotCellTypeProps(clusters = subset_cmo_CA$conditionsxx, sample = subset_cmo_CA$cl_3) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 10, margin = margin(l = 1)),
    legend.title = element_text(size = 0, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l = -10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_3) +  
  scale_fill_manual(values = c( "#009E73","#E69F00"))+geom_hline(yintercept = 0.5,linetype=2,color='red')+NoLegend()
pp3
```



```{r,fig.width=9, fig.height=7}
# Load necessary libraries
pp4 <-
  plotCellTypeProps(clusters = subset_cmo_CA$conditionsxx, sample = subset_cmo_CA$Phase) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16),
    axis.text.y = element_text(size = 7),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 10, margin = margin(l = 1)),
    legend.title = element_text(size = 0, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l = -10, unit = "pt")
  ) + 
  scale_x_discrete(limits=c("G1","S","G2M")) +  
  scale_fill_manual(values = c( "#009E73","#E69F00"))+geom_hline(yintercept = 0.5,linetype=2,color='red')+NoLegend()
pp4
```



```{r,fig.width=20,fig.height=10}
plot_grid(
 pp1,
 pp2,
 pp3,
 pp4,
 ncol = 4,rel_widths = c(4,15,33,4),align = "h")
```

```{r}
ggsave("./Atf3/plotCellTypeProps_flipped_new_1.png",width = 20,height = 4.8, units = "in",dpi = 900)
ggsave("./Atf3/plotCellTypeProps_flipped_new_1.pdf",width = 20,height = 4.8, units = "in",dpi = 900)
ggsave("./Atf3/plotCellTypeProps_flipped_new_1.svg",width = 20,height = 4.8, units = "in",dpi = 900)
```



```{r}
sessionInfo()
```













