---
title: "Differential Gene Expression Analysis with DESeq2"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---


Pseudobulked differential gene expression analysis with DESeq2. Process will be repeated for each annotations in groups, clusters and sub-clusters. And number of significant genes will be stored in a table as a summary table.

gene_info can be found in either in github or in the gse.
```{r eval=FALSE}
suppressPackageStartupMessages({
library(Seurat)
library(DESeq2)
library(dplyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(ggvenn)
library(venn)
library(tidyverse)
library(VennDiagram)
library(cowplot)
library(UpSetR)  
})
```


# Data Preprocessing

Example code for analysis. It will be repeated for conditions which desired.

```{r  eval=FALSE}
Atf3NT <- readRDS("~/cmo_chapter_v1/Analysis/RDS/CMO/cmo_CA.RDS")
gene_info <- read.csv("~/cmo_chapter/figs/DEG/gene_info.csv")
# Gene_info is extracted from cellranger's annotation file contains gene symbol and ensmbl name. 
```


```{r  eval=FALSE}
## Setting the Idents
Idents(Atf3NT) <- "cl_3"
## gene of interest
gene_of_interest <- "Atf3"
## Define the output folder
output_folder <- "DESeq2_results_Atf3_cl_3"
dir.create(output_folder, showWarnings = FALSE)
## cleaning up the remaning levels(from subsetting etc)
Atf3NT$Samples <- droplevels(Atf3NT$Samples_lv)
Atf3NT$c_groups <- droplevels(Atf3NT$cl_3)
## Getting the cluster names     
cluster_names <- unique(Atf3NT$c_groups)
## Average Expression
exp <- AverageExpression(Atf3NT, features = "Atf3")
exp <- exp$RNA
exp <- as.data.frame(exp)
## psudobulk
cts <- AggregateExpression(Atf3NT,group.by = c("c_groups","Samples"),
                                    assays = "RNA",
                                    slot = "counts",
                                    return.seurat = F)
# Just follow
#------------------------------------------------------------------------------------------------------------------
cts <- cts$RNA
# transpose
cts.t <- t(cts)
# cnvert to data.frame
cts.t <- as.data.frame(cts.t)
# get values where to split
splitRows <- gsub('_.*', '', rownames(cts.t))
# split data.frame
cts.split <- split.data.frame(cts.t,
               f = factor(splitRows))
#fix colnames ANd trANspose
cts.split.modified <- lapply(cts.split, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})
```



```{r  eval=FALSE}
# Adapt summary_dataframe structure for the new requirements including average expression
summary_dataframe <- data.frame(
  cluster=character(),
  average_expression=integer(),
  total_genes=integer(),
  included_genes=integer(),
  excluded_genes=integer(),
  padj_0_01=integer(),
  upregulated_genes_padj_0_01=integer(),
  downregulated_genes_padj_0_01=integer(),
  padj_0_05=integer(),
  upregulated_genes_padj_0_05=integer(),
  downregulated_genes_padj_0_05=integer(),
  padj_0_1=integer(),
  upregulated_genes_padj_0_1=integer(),
  downregulated_genes_padj_0_1=integer(),
  pvalue_0_05=integer(),
  upregulated_genes_pvalue_0_05=integer(),
  downregulated_genes_pvalue_0_05=integer(),
  stringsAsFactors=FALSE
)

# Loop through each cluster
for (cluster_name in cluster_names) {
  if (cluster_name %in% names(cts.split.modified)) {
    counts_cluster <- cts.split.modified[[cluster_name]]

    # Calculate average expression for the gene of interest
     avg_exp <- exp[[cluster_name]]
     
    colData <- data.frame(samples = colnames(counts_cluster))
    colData <- colData %>%

mutate(condition = ifelse(grepl(gene_of_interest, samples), 'KO', 'NT')) %>%
      column_to_rownames(var = 'samples')
    
    # Create DESeq2 object
    dds <- DESeqDataSetFromMatrix(countData = counts_cluster,
                                  colData = colData,
                                  design = ~ condition)

    # Pre-filtering
    smallestGroupSize <- 3
    total_genes_before <- nrow(dds)
    keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
    dds <- dds[keep,]
    total_genes_after <- nrow(dds)
    
    # DESeq2 analysis
    dds$condition <- relevel(dds$condition, ref = "NT")
    dds <- DESeq(dds)
    res <- results(dds)
    resOrdered <- res[order(res$pvalue),]
    
    resOrdered <- resOrdered %>%
        data.frame() %>%
        rownames_to_column(var="gene") %>%
        as_tibble()
    
    # Enhancing res and resOrdered with gene_info
    resOrdered_enhanced <- merge(resOrdered, gene_info, by.x = "gene", by.y = "gene_symbol", all.x = TRUE)

    # Save the enhanced results for each cluster
    save(res, file=paste0(output_folder, "/", cluster_name, "_res.RData"))
    save(resOrdered_enhanced, file=paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.RData"))
    write.table(resOrdered_enhanced, file=paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.txt"), sep = '\t',quote = F, row.names = F)
    save(dds, file=paste0(output_folder, "/", cluster_name, "_dds.RData"))
    write.table(counts_cluster, file=paste0(output_folder, "/", cluster_name, "_counts_cluster_table.txt"), sep = '\t',quote = F, row.names = F)

    # Handling NA values for summary statistics
    complete_cases <- complete.cases(resOrdered_enhanced$padj, resOrdered_enhanced$log2FoldChange)

    # Update summary data frame with the calculated average expression
    summary_dataframe <- rbind(summary_dataframe, 
                               data.frame(
                                 cluster=cluster_name,
                                 average_expression=avg_exp,
                                 total_genes=total_genes_before,
                                 included_genes=total_genes_after,
                                 excluded_genes=total_genes_before - total_genes_after,
                                 padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01, na.rm = TRUE),
                                 upregulated_genes_padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05, na.rm = TRUE),
                                 upregulated_genes_padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1, na.rm = TRUE),
                                 upregulated_genes_padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05, na.rm = TRUE),
                                 upregulated_genes_pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE)
                               ))
  }
}
```

```{r}
# Save the final summary data frame
write.csv(summary_dataframe, file=paste0(output_folder, "/summary_dataframe.csv"), row.names = FALSE)
```

```{r  eval=FALSE}
Idents(Atf3NT) <- "cl_2"
levels(Atf3NT)
```

```{r eval=FALSE}
# Adapt summary_dataframe structure for the new requirements including average expression
summary_dataframe <- data.frame(
  cluster=character(),
  average_expression=integer(),
  total_genes=integer(),
  included_genes=integer(),
  excluded_genes=integer(),
  padj_0_01=integer(),
  upregulated_genes_padj_0_01=integer(),
  downregulated_genes_padj_0_01=integer(),
  padj_0_05=integer(),
  upregulated_genes_padj_0_05=integer(),
  downregulated_genes_padj_0_05=integer(),
  padj_0_1=integer(),
  upregulated_genes_padj_0_1=integer(),
  downregulated_genes_padj_0_1=integer(),
  pvalue_0_05=integer(),
  upregulated_genes_pvalue_0_05=integer(),
  downregulated_genes_pvalue_0_05=integer(),
  stringsAsFactors=FALSE
)
```

```{r eval=FALSE}
## Setting the Idents
Idents(Atf3NT) <- "cl_2"
Atf3NT$ident <- Atf3NT$cl_2
## gene of interest
gene_of_interest <- "Atf3"
## Define the output folder
output_folder <- "DESeq2_results_Atf3_cl_2"
dir.create(output_folder, showWarnings = FALSE)
## cleaning up the remaning levels(from subsetting etc)
Atf3NT$Samples <- droplevels(Atf3NT$Samples_lv)
Atf3NT$c_groups <- droplevels(Atf3NT$cl_2)

## Average Expression
exp <- AverageExpression(Atf3NT, features = "Atf3")
exp <- exp$RNA
exp <- as.data.frame(exp)
## psudobulk
cts <- AggregateExpression(Atf3NT,group.by = c("c_groups","Samples"),
                                    assays = "RNA",
                                    slot = "counts",
                                    return.seurat = F)
# Just follow
#------------------------------------------------------------------------------------------------------------------
cts <- cts$RNA
# transpose
cts.t <- t(cts)
# cnvert to data.frame
cts.t <- as.data.frame(cts.t)
# get values where to split
splitRows <- gsub('_.*', '', rownames(cts.t))
# split data.frame
cts.split <- split.data.frame(cts.t,
               f = factor(splitRows))
#fix colnames ANd trANspose
cts.split.modified <- lapply(cts.split, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})


## Getting the cluster names     
cluster_names <- names(cts.split.modified)
```

```{r  eval=FALSE}
# Adapt summary_dataframe structure for the new requirements including average expression
summary_dataframe <- data.frame(
  cluster=character(),
  average_expression=integer(),
  total_genes=integer(),
  included_genes=integer(),
  excluded_genes=integer(),
  padj_0_01=integer(),
  upregulated_genes_padj_0_01=integer(),
  downregulated_genes_padj_0_01=integer(),
  padj_0_05=integer(),
  upregulated_genes_padj_0_05=integer(),
  downregulated_genes_padj_0_05=integer(),
  padj_0_1=integer(),
  upregulated_genes_padj_0_1=integer(),
  downregulated_genes_padj_0_1=integer(),
  pvalue_0_05=integer(),
  upregulated_genes_pvalue_0_05=integer(),
  downregulated_genes_pvalue_0_05=integer(),
  stringsAsFactors=FALSE
)

# Loop through each cluster
for (cluster_name in cluster_names) {
  if (cluster_name %in% names(cts.split.modified)) {
    counts_cluster <- cts.split.modified[[cluster_name]]

    # Calculate average expression for the gene of interest
     avg_exp <- exp[[cluster_name]]

     
    colData <- data.frame(samples = colnames(counts_cluster))
    colData <- colData %>%

mutate(condition = ifelse(grepl(gene_of_interest, samples), 'KO', 'NT')) %>%
      column_to_rownames(var = 'samples')
    
    # Create DESeq2 object
    dds <- DESeqDataSetFromMatrix(countData = counts_cluster,
                                  colData = colData,
                                  design = ~ condition)

    # Pre-filtering
    smallestGroupSize <- 3
    total_genes_before <- nrow(dds)
    keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
    dds <- dds[keep,]
    total_genes_after <- nrow(dds)
    
    # DESeq2 analysis
    dds$condition <- relevel(dds$condition, ref = "NT")
    dds <- DESeq(dds)
    res <- results(dds)
    resOrdered <- res[order(res$pvalue),]
    
    resOrdered <- resOrdered %>%
        data.frame() %>%
        rownames_to_column(var="gene") %>%
        as_tibble()
    
    # Enhancing res and resOrdered with gene_info
    resOrdered_enhanced <- merge(resOrdered, gene_info, by.x = "gene", by.y = "gene_symbol", all.x = TRUE)

    # Save the enhanced results for each cluster
    save(res, file=paste0(output_folder, "/", cluster_name, "_res.RData"))
    save(resOrdered_enhanced, file=paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.RData"))
    write.table(resOrdered_enhanced, file=paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.txt"), sep = '\t',quote = F, row.names = F)
    save(dds, file=paste0(output_folder, "/", cluster_name, "_dds.RData"))
    write.table(counts_cluster, file=paste0(output_folder, "/", cluster_name, "_counts_cluster_table.txt"), sep = '\t',quote = F, row.names = F)

    # Handling NA values for summary statistics
    complete_cases <- complete.cases(resOrdered_enhanced$padj, resOrdered_enhanced$log2FoldChange)

    # Update summary data frame with the calculated average expression
    summary_dataframe <- rbind(summary_dataframe, 
                               data.frame(
                                 cluster=cluster_name,
                                 average_expression=avg_exp,
                                 total_genes=total_genes_before,
                                 included_genes=total_genes_after,
                                 excluded_genes=total_genes_before - total_genes_after,
                                 padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01, na.rm = TRUE),
                                 upregulated_genes_padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_01=sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05, na.rm = TRUE),
                                 upregulated_genes_padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_05=sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1, na.rm = TRUE),
                                 upregulated_genes_padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_1=sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05, na.rm = TRUE),
                                 upregulated_genes_pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_pvalue_0_05=sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE)
                               ))
  }
}
```

```{r eval=FALSE}
# Save the final summary data frame
write.csv(summary_dataframe, file=paste0(output_folder, "/summary_dataframe.csv"), row.names = FALSE)
```

```{r}
sessionInfo()
```










