---
title: "Unique Clusters Marker Search"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r}
library(Seurat)
```


We observed there are three small clusters/populations. Due to limitations of the cell numbers, we cannot get proper DESeq2 results from pseudbulked data. For this reason we will use seurat for finding markers.

# baIfit1

```{r}
s_cmo$ident <- s_cmo$cl_2
Idents(s_cmo) <- "cl_2"
baIfit1_cmo_CvsN_3 <- FindMarkers(s_cmo, ident.1 = "Mesoderm_Ifit1" ,ident.2 = c("Mesoderm_Hoxb","Mesoderm_Hoxa"),min.pct=0,logfc.threshold=0  )
```
```{r}
# Replace p_val_adj values of 0 with a very small number
baIfit1_cmo_CvsN_3$p_val_adj[baIfit1_cmo_CvsN_3$p_val_adj == 0] <- 1e-300
# Replace p_ values of 0 with a very small number
baIfit1_cmo_CvsN_3$p_val[baIfit1_cmo_CvsN_3$p_val == 0] <- 1e-300
```

## adding ranking column (which partially tries to mimic the stat values from DESeq2)

```{r}
baIfit1_cmo_CvsN_3$statlike <-  sign(baIfit1_cmo_CvsN_3$avg_log2FC)*(-log10(baIfit1_cmo_CvsN_3$p_val))
```

# EndRsad2
```{r}
s_cmo$ident <- s_cmo$cl_3
Idents(s_cmo) <- "cl_3"
EndRsad2_cmo_CvsN_3 <- FindMarkers(s_cmo, ident.1 = "EndRsad2" ,ident.2 = c("EndGadd45g","EndCdh5","EndCyp26b1"),min.pct=0,logfc.threshold=0  )
```

```{r}
# Replace p_val_adj values of 0 with a very small number
EndRsad2_cmo_CvsN_3$p_val_adj[EndRsad2_cmo_CvsN_3$p_val_adj == 0] <- 1e-300
# Replace p_ values of 0 with a very small number
EndRsad2_cmo_CvsN_3$p_val[EndRsad2_cmo_CvsN_3$p_val == 0] <- 1e-300
```

## adding ranking column (which partially tries to mimic the stat values from DESeq2)
```{r}
EndRsad2_cmo_CvsN_3$statlike <-  sign(EndRsad2_cmo_CvsN_3$avg_log2FC)*(-log10(EndRsad2_cmo_CvsN_3$p_val))
```

# EndCyp26b1
```{r}
s_cmo$ident <- s_cmo$cl_3
Idents(s_cmo) <- "cl_3"
EndCyp26b1_cmo_CvsN_3 <- FindMarkers(s_cmo, ident.1 = "EndCyp26b1" ,ident.2 = c("EndGadd45g","EndCdh5","EndRsad2" ),min.pct=0,logfc.threshold=0  )
```

```{r}
# Replace p_val_adj values of 0 with a very small number
EndCyp26b1_cmo_CvsN_3$p_val_adj[EndCyp26b1_cmo_CvsN_3$p_val_adj == 0] <- 1e-300
# Replace p_ values of 0 with a very small number
EndCyp26b1_cmo_CvsN_3$p_val[EndCyp26b1_cmo_CvsN_3$p_val == 0] <- 1e-300
```

```{r}
EndCyp26b1_cmo_CvsN_3$statlike <-  sign(EndCyp26b1_cmo_CvsN_3$avg_log2FC)*(-log10(EndCyp26b1_cmo_CvsN_3$p_val))
```


```{r}
sessionInfo()
```



