---
title: "Analysis of Flow Cytometric Data"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

# Introduction

In this notebook, we analyze flow cytometry data to compare the effects of **Atf3 Knockout (KO)** and **Non-Targeting Wild Type (NT WT)** conditions on various cellular measurements. We perform mixed-effects linear modeling to assess the statistical significance of differences between these two conditions while accounting for differentiation stages as random effects.

**Note:** The data files `FCA_2024_3.csv` and `FCA_2024_2.csv` should be available in the `../../FCA_DATA/` directory.

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(lme4)
library(ggplot2)
library(reshape2)
library(ggpubr)
```

# Data Preparation

## Loading the Data

First, we load the main dataset containing the flow cytometry measurements.

```{r}
# Load the flow cytometry data
data <- read.csv("../../FCA_DATA/FCA_2024_3.csv", check.names = FALSE)

# Convert certain columns to factors
data$Samples0 <- as.factor(data$Samples0)
data$Samples1 <- as.factor(data$Samples1)
data$Samples2 <- as.factor(data$Samples2)
data$Samples3 <- as.factor(data$Samples3)
data$Conditions <- as.factor(data$Conditions)
data$DifferentiationA <- as.factor(data$DifferentiationA)
data$DifferentiationB <- as.factor(data$DifferentiationB)
data$DifferentiationC <- as.factor(data$DifferentiationC)

# Save a copy of the original data
data0 <- data
```

**Explanation:**

-   **`read.csv`**: Reads the CSV file containing the data.

-   **`as.factor`**: Converts specified columns to factor type for categorical variables.

-   **`data0`**: A backup of the original data before any modifications.

## Standardizing Column Names

We standardize column names for consistency and ease of use.

```{r}
# Standardize column names
data$Conditions <- data$Conditions
data$Differentiation <- data$DifferentiationA
data$Replicate <- data$Samples0

# Create another version of the data
data1 <- data

# Keep only necessary columns (remove columns 1,2,3,4,6,7,8)
data <- data[ , -c(1, 2, 3, 4, 6, 7, 8)]
data2 <- data

```

**Explanation:**

-   We create a new column **`Differentiation`** based on **`DifferentiationA`**.

-   **`Replicate`** is set to **`Samples0`** to indicate replicate samples.

-   We remove unnecessary columns to focus on relevant data.

## Loading Human-Readable Headers

We load another CSV file that contains human-readable titles or headers for the measurements.

```{r}
# Load human-readable titles/headers
humanreadble <- read.csv("../../FCA_DATA/FCA_2024_2.csv", check.names = FALSE)

# Keep only the first two rows (assumed to contain headers)
humanreadble <- humanreadble[1:2, ]
```

**Explanation:**

-   The **`humanreadble`** dataframe contains the mapping from code names to human-readable measurement names.

# Data Filtering

We focus our analysis on the **Atf3 KO** and **NT WT** conditions.

```{r}
# Filter data for Atf3 KO and NT WT conditions
data_Atf3_control <- data %>% filter(Conditions %in% c("Atf3 KO", "NT WT"))

# Remove unused factor levels
data_Atf3_control <- droplevels(data_Atf3_control)

```

**Explanation:**

-   **`filter`**: Selects rows where **`Conditions`** is either "Atf3 KO" or "NT WT".

-   **`droplevels`**: Removes unused levels from factor variables.

We also prepare the data with human-readable headers.

```{r}
# Create a copy of the original data with human-readable headers
data_hr <- data0

# Replace column names with human-readable names
colnames(data_hr) <- humanreadble[1, ]

# Verify the changes
head(data_hr)
```

**Explanation:**

-   We replace the column names in **`data_hr`** with the first row of **`humanreadble`**, which contains the human-readable names.

Filter the human-readable data for Atf3 KO and NT WT conditions.

```{r}
# Filter human-readable data for Atf3 KO and NT WT conditions
data_Atf3_control_hr <- data_hr %>% filter(Conditions %in% c("Atf3 KO", "NT WT"))

# Remove unused factor levels
data_Atf3_control_hr <- droplevels(data_Atf3_control_hr)
```

# Statistical Analysis

We perform mixed-effects linear modeling to assess the differences between conditions.

## Defining the Mixed-Effects Model Function

We define a function **`run_mixed_effects`** that runs a mixed-effects model for a given measurement.

```{r}
# Define the mixed-effects linear model function
run_mixed_effects <- function(data, measurement) {
  tryCatch({
    # Ensure the response variable is numeric
    if (!is.numeric(data[[measurement]])) {
      stop("The response variable must be numeric.")
    }
    
    # Fit the mixed-effects model
    mixed_model <- lmer(
      as.formula(paste(measurement, "~ Conditions + (1 | Differentiation)")), 
      data = data, 
      na.action = na.exclude
    )
    
    # Check for singular fit
    if (isSingular(mixed_model)) {
      warning("Singular fit for measurement ", measurement)
      return(list(
        result = data.frame(
          Measurement = measurement, 
          Mixed_Effects = NA,
          Mean_Atf3 = NA,
          Mean_Control = NA,
          N_Atf3 = NA,
          N_Control = NA,
          Overall_Mean = NA,
          SD = NA,
          Median = NA,
          IQR = NA,
          AIC = NA,
          BIC = NA,
          Effect_Size = NA,
          CI_Lower = NA,
          CI_Upper = NA
        ), 
        singular = TRUE
      ))
    }
    
    # Get summary statistics
    mixed_summary <- summary(mixed_model)
    
    # Calculate means and counts for each condition
    condition_means <- data %>%
      group_by(Conditions) %>%
      summarize(Mean = mean(!!sym(measurement), na.rm = TRUE))
      
    condition_n <- data %>%
      group_by(Conditions) %>%
      summarize(N = sum(!is.na(!!sym(measurement))))
    
    # Compile results
    result <- data.frame(
      Measurement = measurement, 
      Mixed_Effects = mixed_summary$coefficients[2, "Pr(>|t|)"],
      Mean_Atf3 = condition_means$Mean[condition_means$Conditions == "Atf3 KO"],
      Mean_Control = condition_means$Mean[condition_means$Conditions == "NT WT"],
      N_Atf3 = condition_n$N[condition_n$Conditions == "Atf3 KO"],
      N_Control = condition_n$N[condition_n$Conditions == "NT WT"],
      Overall_Mean = mean(data[[measurement]], na.rm = TRUE),
      SD = sd(data[[measurement]], na.rm = TRUE),
      Median = median(data[[measurement]], na.rm = TRUE),
      IQR = IQR(data[[measurement]], na.rm = TRUE),
      AIC = AIC(mixed_model),
      BIC = BIC(mixed_model),
      Effect_Size = mixed_summary$coefficients[2, "Estimate"],
      CI_Lower = confint(mixed_model, oldNames = FALSE)[2, 1],
      CI_Upper = confint(mixed_model, oldNames = FALSE)[2, 2]
    )
    
    return(list(result = result, singular = FALSE))
    
  }, error = function(e) {
    message("Error for measurement ", measurement, ": ", e$message)
    return(list(
      result = data.frame(
        Measurement = measurement, 
        Mixed_Effects = NA,
        Mean_Atf3 = NA,
        Mean_Control = NA,
        N_Atf3 = NA,
        N_Control = NA,
        Overall_Mean = NA,
        SD = NA,
        Median = NA,
        IQR = NA,
        AIC = NA,
        BIC = NA,
        Effect_Size = NA,
        CI_Lower = NA,
        CI_Upper = NA
      ), 
      singular = NA
    ))
  })
}

```

**Explanation:**

-   The function takes **`data`** and a **`measurement`** column name.

-   Fits a mixed-effects model with **`Conditions`** as a fixed effect and **`Differentiation`** as a random effect.

-   Handles singular fits and errors gracefully.

-   Returns a list containing the result and a flag indicating singularity.

## Running the Mixed-Effects Models

We apply the function to all numeric measurement columns.

```{r}
# Initialize empty data frames to store results
mixed_effects_results <- data.frame()
significant_mixed_effects <- data.frame()
singular_fit_measurements <- c()

# Significance threshold for p-values
significance_threshold <- 0.05

# Extract column names for numeric measurements
measurement_cols <- colnames(data_Atf3_control)[sapply(data_Atf3_control, is.numeric)]

# Loop over each measurement
for (measurement in measurement_cols) {
  if (nrow(data_Atf3_control) > 0) {
    # Run the mixed-effects model
    result <- run_mixed_effects(data_Atf3_control, measurement)
    
    # Store the results
    mixed_effects_results <- rbind(mixed_effects_results, result$result)
    
    # Check for significance
    if (!is.na(result$result$Mixed_Effects) && result$result$Mixed_Effects < significance_threshold) {
      significant_mixed_effects <- rbind(significant_mixed_effects, result$result)
    }
    
    # Record measurements with singular fits
    if (!is.na(result$singular) && result$singular) {
      singular_fit_measurements <- c(singular_fit_measurements, measurement)
    }
  }
}

# Print all results
print(mixed_effects_results)

# Print significant results
print(significant_mixed_effects)

# List measurements with singular fits
cat("Measurements with singular fits:\n")
print(singular_fit_measurements)

```

**Explanation:**

-   Iterate over each numeric measurement and run the mixed-effects model.

-   Store all results in **`mixed_effects_results`**.

-   Collect significant results in **`significant_mixed_effects`** based on the p-value threshold.

-   Record any measurements that resulted in singular fits.

## Mapping to Human-Readable Measurement Names

We replace the measurement codes with human-readable names for clarity.

```{r}
# Convert the humanreadable dataframe to a named vector for lookup
human_readable_names <- as.list(humanreadble[1, ])

# Function to replace measurement names with human-readable names
replace_with_human_readable <- function(measurements, lookup) {
  sapply(measurements, function(measurement) {
    if (measurement %in% names(lookup)) {
      return(lookup[[measurement]])
    } else {
      return(measurement)
    }
  })
}

# Replace Measurement names in mixed_effects_results with human-readable names
mixed_effects_results$Measurement_hr <- replace_with_human_readable(mixed_effects_results$Measurement, human_readable_names)

# Print the updated dataframe to verify the changes
print(mixed_effects_results)


# Replace Measurement names in mixed_effects_results with human-readable names
significant_mixed_effects$Measurement_hr <- replace_with_human_readable(significant_mixed_effects$Measurement, human_readable_names)

# Print the updated dataframe to verify the changes
print(significant_mixed_effects)
```

# **Explanation:**

-   Create a lookup list from the **`humanreadble`** dataframe.

-   **`replace_with_human_readable`** function replaces measurement codes with human-readable equivalents.

-   Apply this function to both **`mixed_effects_results`** and **`significant_mixed_effects`**.

# Visualization

We create plots to visualize the results of the mixed-effects models.

## Preparing Data for Plotting

```{r}
# Add a column for significance labels (GraphPad-like notation)
mixed_effects_results$Significance <- cut(
  mixed_effects_results$Mixed_Effects,
  breaks = c(-Inf, 0.0001, 0.001, 0.01, 0.05, Inf),
  labels = c("****", "***", "**", "*", "n.s.")
)

# Extract measurements with human-readable names
mixed_effects_results_mixed <- mixed_effects_results$Measurement_hr

# Melt the data for ggplot2
data_melted_mixed <- data_Atf3_control_hr %>%
  select(
    Samples0, Samples1, Samples2, Samples3, Conditions, 
    DifferentiationA, DifferentiationB, DifferentiationC, 
    all_of(mixed_effects_results_mixed)
  ) %>%
  melt(
    id.vars = c("Samples0", "Samples1", "Samples2", "Samples3", "Conditions", "DifferentiationA", "DifferentiationB", "DifferentiationC"), 
    variable.name = "Measurement_hr", 
    value.name = "Value"
  )

# Merge with the mixed-effects results to get significance labels
data_melted_mixed <- merge(data_melted_mixed, mixed_effects_results, by = "Measurement_hr")

```

**Explanation:**

-   Categorize p-values into significance labels (e.g., "\*\*\*\*" for p \< 0.0001).

-   Prepare the data for plotting by melting it into long format suitable for **`ggplot2`**.

-   Merge the data with the results to include significance labels.

## Creating the Faceted Plot

```{r,fig.width=15, fig.height=8}
# Create a data frame for custom annotations (significance labels)
annotations_df <- data_melted_mixed %>%
  group_by(Measurement_hr) %>%
  summarize(
    y_position = min(max(Value, na.rm = TRUE) * 1.05, 98),  # Position the annotation above the max value but not exceeding 100
    Significance = first(Significance)
  )

# Create the faceted boxplot
plot <- ggplot(data_melted_mixed, aes(x = Conditions, y = Value)) +
  geom_boxplot(alpha = 1, size = 0.2, outlier.shape = NA) +  # Thin boxplot lines
  geom_jitter(aes(color = Samples3, shape = DifferentiationC), size = 1.5, width = 0.2, alpha = 1) +
  # Define your color palette (replace r3dcol$l10z with your color vector)
  scale_color_manual(values = c("blue", "red", "green", "purple", "orange")) +
  facet_wrap(~ Measurement_hr, scales = "free", ncol = 11) +
  labs(
    title = "Mixed-Effects Results: Atf3-KO vs NT-WT",
    y = "",
    x = ""
  ) +
  theme_classic() +
  theme(
    plot.title = element_blank(),
    axis.title.y = element_blank(),
    strip.text = element_text(size = 8),
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 6),
    axis.text.y = element_text(size = 6),
    axis.ticks = element_line(size = 0.2),
    axis.line = element_line(size = 0.2),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 8, margin = margin(l = 2, unit = "pt")),
    legend.margin = margin(t = -5),
    legend.box.margin = margin(t = -10),
    legend.spacing.x = unit(0.3, 'cm'),
    legend.spacing.y = unit(0.05, 'cm'),
    legend.key.size = unit(0, "cm"),
    legend.key.spacing.x = unit(0.1, "cm"),
    legend.key.width = unit(0.0, "cm"),
    legend.key.height = unit(0.0, "cm")
  ) +
  guides(
    color = guide_legend(title = "Samples", nrow = 1), 
    shape = guide_legend(title = "Differentiation", nrow = 1)
  ) +
  geom_text(
    data = annotations_df, 
    aes(x = 1.5, y = y_position, label = Significance), 
    vjust = 0.2, hjust = 0.5, size = 2.5, color = "black"
  )

# Display the plot
print(plot)

```

**Explanation:**

-   Create a faceted boxplot for each measurement, comparing the two conditions.

-   **`geom_boxplot`**: Plots the boxplots with thin lines.

-   **`geom_jitter`**: Adds individual data points.

-   **`facet_wrap`**: Creates a grid of plots for each measurement.

-   Annotations are added to display the significance level above each plot.

-   **`scale_color_manual`**: Define custom colors for samples.

## Saving the Plot

```{r}
# Save the plot in different formats
ggsave("./plot.png", width = 15, height = 8, units = "in", dpi = 600)
ggsave("./plot.pdf", width = 15, height = 8, units = "in", dpi = 600)
ggsave("./plot.svg", width = 15, height = 8, units = "in", dpi = 600)

```

**Explanation:**

-   **`ggsave`**: Saves the last plot to a file in the specified format.

# Exporting the Results

We save the results of the mixed-effects models to text files for further analysis.

```{r}
# Write the results to text files
write.table(
  mixed_effects_results,
  "./mixed_effects_results_Atf3.txt",
  col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t'
)

write.table(
  significant_mixed_effects,
  "./significant_mixed_effects_results_Atf3.txt",
  col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t'
)

```

**Explanation:**

-   **`write.table`**: Exports data frames to text files in tab-delimited format.

# Session Information

Finally, we include the session information to document the R environment used for the analysis.

```{r}
sessionInfo()
```

**Explanation:**

-   **`sessionInfo`**: Displays information about the R session, including loaded packages and their versions.
