---
title: "Demultiplexing"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Glossary

**Lane**: Output of single-cell RNA-seq as a unit. In the context of
scRNA-seq, we use chips, each with 8 lanes. These chips can be loaded
with 8 lanes each time. To simplify, I refer to whatever we obtain from
each lane as a 'Lane'.

**Library**: Essentially the same as a 'Lane'. The cells were
captured using 10X Chips, and the resulting material was prepared for
NGS (Next-Generation Sequencing). The cells were transformed, and an NGS
library was created for sequencing.

**NT**: In this case, Non-Targeting Control. Similar to Wild Type
(WT). Keeping all the experimental conditions the same, we treat these
samples the same way as the Knockouts (KOs).

**Negative**: Based on the result of the demultiplexing, the cells
weren't labeled with a multiplexing reagent.

**Doublets**: Droplets containing more than one cell from different
samples (this is not the doublets that occur within the same sample).

**Singlets**: Based on the results of the demultiplexing, droplets
contain only one sample barcode.

**Unassigned**: Cells which cannot be assigned with multiplexing.

**CMO**: Cell Multiplexing Oligos used in the demultiplexing method
for this experiment (CMO301 - CMO312).

**CellRanger**: The 10X Genomics pipeline that processes the NGS
data and provides the transcriptome/gene expression matrix and other
files. In this case, it also includes demultiplexing data.

**HTODemux**: Seurat's demultiplexing function/package.

**Droplet**: Single-cell beads plus captured cells in a droplet. It
can be one cell and bead or multiple cells in a bead.

# Introduction

In this notebook, I will summarize the codes, packages, and parameters
used for the pre-processing and quality control (QC) steps of the CMO
project. The CMO project involves multiplexed Atf3-KO, Bcl6b-KO,
Zfp711-KO, and NT samples, resulting in 5 libraries of scRNA-seq
datasets. I will outline the steps taken to prepare the data for
downstream analysis.

**Note:** I am not including the original exploratory code here to
maintain clarity and focus on reproducibility.

# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP AND PACKAGE LOADING
# ==============================================================================

```{r}
# ==============================================================================
# Load Required Packages
# ==============================================================================

# Suppress package loading messages for cleaner output
suppressMessages({
  
  # dplyr: Data manipulation and transformation (filter, mutate, group_by, etc.)
  library(dplyr)
  
  # Seurat: Primary package for single-cell RNA-seq analysis
  # Provides functions for QC, normalization, clustering, and visualization
  library(Seurat)
  
  # ggplot2: Grammar of graphics for creating publication-quality plots
  library(ggplot2)
  
  # cowplot: Extension of ggplot2 for combining multiple plots
  library(cowplot)
  
  # ggvenn: Create Venn diagrams using ggplot2 syntax
  library(ggvenn)
  
  # data.table: Fast data reading and manipulation (fread function)
  library(data.table)
  
  # gprofiler2: Gene list functional enrichment and ortholog conversion
  library(gprofiler2)
})

# ==============================================================================
# Set Working Directory
# ==============================================================================

# Set the working directory to the location of your data files
# Replace "your/working/directory" with the actual path
setwd("your/working/directory")

# Verify the working directory was set correctly
getwd()
```

# Data Preprocessing

We have 5 single-cell libraries containing 12 samples (Atf3KO x3,
Bcl6b-KO x3, Zfp711-KO x3, NT-Control x3). To reach enough cell numbers,
we created 5 libraries. Due to multiplexing of the single-cell data, we
also have multiplexing data (CMO) alongside the transcriptomic data. The
goals of demultiplexing are:

1.  **Demultiplexing the samples**: Identifying the origin of each
    single cell.

2.  **Separation of cross-sample multiplets**: Detecting droplets
    containing cells from different samples.

3.  **Indirect assessment of cell types with higher multiplet rates**.

4.  **Observing unlabelled or low-labelling cell types**.

## Loading Data

We will start by loading the CellRanger outputs. Note that 'Library' and
'Lane' will be used interchangeably.

CellRanger provides separate sample files after demultiplexing data, but
we will use the unseparated files here to retain all available
information.

# ==============================================================================
# SECTION 2: DATA LOADING - TRANSCRIPTOME DATA
# ==============================================================================

### Transcriptome Data

```{r eval=FALSE}
# ==============================================================================
# Load 10X Genomics Data for Lane 1 (Example)
# ==============================================================================
# This process should be repeated for all 5 lanes

# Define the path to the raw feature barcode matrix directory
# This directory contains: barcodes.tsv.gz, features.tsv.gz, matrix.mtx.gz
data_dir_l1 <- "path/to/Lane_1/raw_feature_bc_matrix/"

# Read10X: Reads the 10X Genomics output files
# Returns a list with 'Gene Expression' and 'Multiplexing Capture' matrices
# - 'Gene Expression': Contains the transcriptome data (genes x cells)
# - 'Multiplexing Capture': Contains the CMO barcode counts (CMOs x cells)
data_l1 <- Read10X(data.dir = data_dir_l1)

# CreateSeuratObject: Initialize a Seurat object with gene expression counts
# - counts: The raw count matrix (genes as rows, cells as columns)
# - Uses the 'Gene Expression' slot from the Read10X output
seurat_Lane1 <- CreateSeuratObject(counts = data_l1$`Gene Expression`)

# Add the CMO (Cell Multiplexing Oligos) data as a separate assay
# CreateAssayObject: Creates a new assay from a count matrix
# This stores the multiplexing capture data alongside the RNA data
seurat_Lane1[["CMO"]] <- CreateAssayObject(counts = data_l1$`Multiplexing Capture`)

# ==============================================================================
# REPEAT FOR LANES 2-5
# ==============================================================================
# The same process should be applied to Lane 2, Lane 3, Lane 4, and Lane 5
# Each lane will have its own Seurat object: seurat_Lane2, seurat_Lane3, etc.
```

**Explanation:**

-   **Read10X** reads the output files from CellRanger.

-   **CreateSeuratObject** initializes a Seurat object with the gene
    expression data.

-   The **'CMO'** assay is added to the Seurat object to store
    multiplexing capture data.

# ==============================================================================
# SECTION 3: INITIAL CELL COUNT SUMMARY
# ==============================================================================

### Display the Number of Cells Loaded from Each Lane

```{r,eval=FALSE}
# ==============================================================================
# Create Summary Table of Initial Cell Counts
# ==============================================================================

# Create a data frame to summarize cell counts across all lanes
# ncol() returns the number of columns (cells) in each Seurat object
cell_counts <- data.frame(
  Sample = c("Lane 1", "Lane 2", "Lane 3", "Lane 4", "Lane 5"),
  Cells = c(ncol(seurat_Lane1), 
            ncol(seurat_Lane2), 
            ncol(seurat_Lane3), 
            ncol(seurat_Lane4), 
            ncol(seurat_Lane5))
)

# Display the cell count table
print(cell_counts)
```

| Sample Name  | Amount of Cells |
|--------------|-----------------|
| seurat_Lane1 | 3065453         |
| seurat_Lane2 | 3029894         |
| seurat_Lane3 | 3216261         |
| seurat_Lane4 | 3305265         |
| seurat_Lane5 | 3010848         |

: **Explanation:**

-   Initially, we have over 3 million cells per lane. However, these
    include empty droplets and low-quality cells.

# ==============================================================================
# SECTION 4: LOAD CELLRANGER DEMULTIPLEXING RESULTS
# ==============================================================================

### CellRanger Demultiplexing Results

We will load the **Confidence Assignment Table** from CellRanger, which
contains demultiplexing information based on the CellRanger software.

```{r  eval=FALSE}
# ==============================================================================
# Load CellRanger Assignment Confidence Tables
# ==============================================================================

# Initialize an empty list to store confidence tables from all lanes
# Using a list allows easy access to each lane's data
confidence_tb_cellranger <- list()

# fread: Fast file reading function from data.table package
# Much faster than read.csv for large files
# The assignment_confidence_table.csv contains:
# - Barcode: Cell barcode
# - Assignment: Sample assignment (CMO ID or 'Multiplet'/'Blanks')
# - Assignment Probability: Confidence score for the assignment

# Load demultiplexing results for Lane 1
confidence_tb_cellranger$CellRanger_L1 <- fread(
  "path/to/Lane_1/outs/multi/multiplexing_analysis/assignment_confidence_table.csv"
)

# Load demultiplexing results for Lane 2
confidence_tb_cellranger$CellRanger_L2 <- fread(
  "path/to/Lane_2/outs/multi/multiplexing_analysis/assignment_confidence_table.csv"
)

# Load demultiplexing results for Lane 3
confidence_tb_cellranger$CellRanger_L3 <- fread(
  "path/to/Lane_3/outs/multi/multiplexing_analysis/assignment_confidence_table.csv"
)

# Load demultiplexing results for Lane 4
confidence_tb_cellranger$CellRanger_L4 <- fread(
  "path/to/Lane_4/outs/multi/multiplexing_analysis/assignment_confidence_table.csv"
)

# Load demultiplexing results for Lane 5
confidence_tb_cellranger$CellRanger_L5 <- fread(
  "path/to/Lane_5/outs/multi/multiplexing_analysis/assignment_confidence_table.csv"
)
```

**Explanation:**

-   **fread** is used to read CSV files efficiently.

-   The **assignment_confidence_table.csv** contains the cell barcodes
    and their assignment confidence from CellRanger's demultiplexing.

## Cleaning Data

We will start cleaning the data in multiple steps.

# ==============================================================================
# SECTION 5: FILTER EMPTY DROPLETS
# ==============================================================================

### Subsetting with CellRanger's Confidence Table

We will remove the empty droplets by subsetting the Seurat objects with
the valid cell barcodes present in the CellRanger confidence tables.

```{r eval=FALSE}
# ==============================================================================
# Subset Seurat Objects to Keep Only Valid Cells
# ==============================================================================
# CellRanger identifies valid cell barcodes vs empty droplets
# We use these barcodes to filter our Seurat objects

# subset(): Filters the Seurat object to keep only specified cells
# cells parameter: Vector of cell barcodes to retain
# confidence_tb_cellranger$CellRanger_L1$Barcode: Column containing valid barcodes

# Filter Lane 1 to keep only cells identified by CellRanger
seurat_Lane1 <- subset(seurat_Lane1, 
                       cells = confidence_tb_cellranger$CellRanger_L1$Barcode)

# Filter Lane 2 to keep only cells identified by CellRanger
seurat_Lane2 <- subset(seurat_Lane2, 
                       cells = confidence_tb_cellranger$CellRanger_L2$Barcode)

# Filter Lane 3 to keep only cells identified by CellRanger
seurat_Lane3 <- subset(seurat_Lane3, 
                       cells = confidence_tb_cellranger$CellRanger_L3$Barcode)

# Filter Lane 4 to keep only cells identified by CellRanger
seurat_Lane4 <- subset(seurat_Lane4, 
                       cells = confidence_tb_cellranger$CellRanger_L4$Barcode)

# Filter Lane 5 to keep only cells identified by CellRanger
seurat_Lane5 <- subset(seurat_Lane5, 
                       cells = confidence_tb_cellranger$CellRanger_L5$Barcode)
```

**Explanation:**

-   This step reduces the dataset to include only barcodes that
    CellRanger has identified as valid cells.

# ==============================================================================
# SECTION 6: POST-FILTERING CELL COUNT SUMMARY
# ==============================================================================

### Display the Number of Cells Remaining After Subsetting

```{r,eval=FALSE}
# ==============================================================================
# Create Summary Table of Cell Counts After Filtering
# ==============================================================================

# Create a data frame showing remaining cells after empty droplet removal
cleaned_cell_counts <- data.frame(
  Sample = c("Lane 1", "Lane 2", "Lane 3", "Lane 4", "Lane 5"),
  Cells = c(ncol(seurat_Lane1), 
            ncol(seurat_Lane2), 
            ncol(seurat_Lane3), 
            ncol(seurat_Lane4), 
            ncol(seurat_Lane5))
)

# Display the filtered cell count table
print(cleaned_cell_counts)
```

| Sample Name  | Amount of Cells |
|--------------|-----------------|
| seurat_Lane1 | 31833           |
| seurat_Lane2 | 26969           |
| seurat_Lane3 | 26951           |
| seurat_Lane4 | 28537           |
| seurat_Lane5 | 30214           |

: **Explanation:**

-   After subsetting, the number of cells per lane is significantly
    reduced, as we have filtered out empty droplets and low-quality
    cells.

# Analysis

We will perform demultiplexing using another method called
**HTODemux()** from the Seurat package. Before that, we need to prepare
the data by normalizing and identifying variable features, which are
essential steps for the demultiplexing process.

## Demultiplexing

### Preparations

# ==============================================================================
# SECTION 7: RNA DATA PREPROCESSING FOR DEMULTIPLEXING
# ==============================================================================

#### RNA Data

We need to process the RNA data to prepare it for demultiplexing.

```{r, warning=FALSE, eval=FALSE}
# ==============================================================================
# Step 7.1: Set Default Assay to RNA
# ==============================================================================
# DefaultAssay(): Sets which assay (RNA, CMO, etc.) is used for operations
# We set RNA as default to process transcriptome data first

DefaultAssay(seurat_Lane1) <- "RNA"
DefaultAssay(seurat_Lane2) <- "RNA"
DefaultAssay(seurat_Lane3) <- "RNA"
DefaultAssay(seurat_Lane4) <- "RNA"
DefaultAssay(seurat_Lane5) <- "RNA"

# ==============================================================================
# Step 7.2: Normalize RNA Data
# ==============================================================================
# NormalizeData(): Applies log-normalization to the count data
# Default method: LogNormalize with scale.factor = 10000
# Formula: log1p(count / total_count * scale.factor)
# This corrects for differences in sequencing depth between cells

seurat_Lane1 <- NormalizeData(seurat_Lane1)
seurat_Lane2 <- NormalizeData(seurat_Lane2)
seurat_Lane3 <- NormalizeData(seurat_Lane3)
seurat_Lane4 <- NormalizeData(seurat_Lane4)
seurat_Lane5 <- NormalizeData(seurat_Lane5)

# ==============================================================================
# Step 7.3: Identify Highly Variable Features
# ==============================================================================
# FindVariableFeatures(): Identifies genes with high cell-to-cell variation
# selection.method = "mean.var.plot": Uses mean-variance relationship
# These genes are most informative for distinguishing cell populations

seurat_Lane1 <- FindVariableFeatures(seurat_Lane1, selection.method = "mean.var.plot")
seurat_Lane2 <- FindVariableFeatures(seurat_Lane2, selection.method = "mean.var.plot")
seurat_Lane3 <- FindVariableFeatures(seurat_Lane3, selection.method = "mean.var.plot")
seurat_Lane4 <- FindVariableFeatures(seurat_Lane4, selection.method = "mean.var.plot")
seurat_Lane5 <- FindVariableFeatures(seurat_Lane5, selection.method = "mean.var.plot")

# ==============================================================================
# Step 7.4: Scale the Data
# ==============================================================================
# ScaleData(): Centers and scales expression values
# - Shifts mean expression to 0
# - Scales variance to 1
# features parameter: Only scale the variable features (saves computation)
# Required before PCA

seurat_Lane1 <- ScaleData(seurat_Lane1, features = VariableFeatures(seurat_Lane1))
seurat_Lane2 <- ScaleData(seurat_Lane2, features = VariableFeatures(seurat_Lane2))
seurat_Lane3 <- ScaleData(seurat_Lane3, features = VariableFeatures(seurat_Lane3))
seurat_Lane4 <- ScaleData(seurat_Lane4, features = VariableFeatures(seurat_Lane4))
seurat_Lane5 <- ScaleData(seurat_Lane5, features = VariableFeatures(seurat_Lane5))
```

#### **Explanation:**

-   **NormalizeData**: Log-normalizes the gene expression data.

-   **FindVariableFeatures**: Identifies genes that exhibit high
    variability across cells, which are informative for downstream
    analysis.

-   **ScaleData**: Scales and centers the data, which is necessary for
    PCA.

# ==============================================================================
# SECTION 8: CMO DATA NORMALIZATION
# ==============================================================================

#### CMO Data

We need to normalize the CMO (Cell Multiplexing Oligos) data.

```{r eval=FALSE}
# ==============================================================================
# Normalize CMO Data Using Centered Log-Ratio (CLR) Method
# ==============================================================================
# NormalizeData() with CLR normalization is recommended for CMO/HTO data
# CLR accounts for the compositional nature of multiplexing oligo counts
# normalization.method = "CLR": Centered Log-Ratio transformation
# assay = "CMO": Specifies to normalize the CMO assay, not RNA

seurat_Lane1 <- NormalizeData(seurat_Lane1, 
                               assay = "CMO", 
                               normalization.method = "CLR")

seurat_Lane2 <- NormalizeData(seurat_Lane2, 
                               assay = "CMO", 
                               normalization.method = "CLR")

seurat_Lane3 <- NormalizeData(seurat_Lane3, 
                               assay = "CMO", 
                               normalization.method = "CLR")

seurat_Lane4 <- NormalizeData(seurat_Lane4, 
                               assay = "CMO", 
                               normalization.method = "CLR")

seurat_Lane5 <- NormalizeData(seurat_Lane5, 
                               assay = "CMO", 
                               normalization.method = "CLR")

```

### **Explanation:**

-   The **CLR normalization** is recommended for CMO data to normalize
    across cells.

# ==============================================================================
# SECTION 9: HTODEMUX DEMULTIPLEXING
# ==============================================================================

### HTODemux Demultiplexing

We will perform demultiplexing using the **HTODemux** function with a
positive quantile of 0.99.

```{r eval=FALSE}
# ==============================================================================
# Run HTODemux Demultiplexing Algorithm
# ==============================================================================
# HTODemux(): Seurat's demultiplexing function for multiplexed samples
# Parameters:
# - assay = "CMO": Use the CMO assay for demultiplexing
# - positive.quantile = 0.99: Threshold for positive classification
#   Cells above the 99th percentile of the negative distribution are positive
# 
# Output adds metadata columns:
# - CMO_maxID: CMO with highest signal
# - CMO_secondID: CMO with second-highest signal
# - CMO_margin: Difference between top two CMO signals
# - CMO_classification: Specific CMO or Doublet/Negative
# - CMO_classification.global: Singlet/Doublet/Negative
# - hash.ID: Final sample assignment

seurat_Lane1 <- HTODemux(seurat_Lane1, 
                          assay = "CMO", 
                          positive.quantile = 0.99)

seurat_Lane2 <- HTODemux(seurat_Lane2, 
                          assay = "CMO", 
                          positive.quantile = 0.99)

seurat_Lane3 <- HTODemux(seurat_Lane3, 
                          assay = "CMO", 
                          positive.quantile = 0.99)

seurat_Lane4 <- HTODemux(seurat_Lane4, 
                          assay = "CMO", 
                          positive.quantile = 0.99)

seurat_Lane5 <- HTODemux(seurat_Lane5, 
                          assay = "CMO", 
                          positive.quantile = 0.99)

```

**Explanation:**

-   **HTODemux** classifies cells based on their CMO signals into
    Singlets, Doublets, or Negative.

-   The **positive.quantile** parameter sets the threshold for
    classifying positive cells.

# ==============================================================================
# SECTION 10: HTODEMUX RESULTS ANALYSIS
# ==============================================================================

### Results of HTODemux

We will examine the results of the demultiplexing.

#### Display Demultiplexing Results

```{r}
# ==============================================================================
# View Global Classification Results (Singlet/Doublet/Negative)
# ==============================================================================
# table(): Creates a frequency table of categorical values
# CMO_classification.global: Contains the overall classification for each cell
# - Singlet: Cell has one dominant CMO (from one sample)
# - Doublet: Cell has multiple CMOs (multiplet from different samples)
# - Negative: Cell has no clear CMO signal

# Display classification counts for each lane
table(seurat_Lane1$CMO_classification.global)
table(seurat_Lane2$CMO_classification.global)
table(seurat_Lane3$CMO_classification.global)
table(seurat_Lane4$CMO_classification.global)
table(seurat_Lane5$CMO_classification.global)

```

**Explanation:**

-   The **CMO_classification.global** metadata field contains the global
    classification of each cell (Singlet, Doublet, Negative).

#### Highest CMOs for Each Cell

```{r}
# ==============================================================================
# View Maximum CMO Signal Per Cell
# ==============================================================================
# CMO_maxID: The CMO barcode with the highest signal in each cell
# This shows which sample's CMO was most abundant, regardless of classification

# Display the distribution of maximum CMO signals for each lane
table(seurat_Lane1$CMO_maxID)
table(seurat_Lane2$CMO_maxID)
table(seurat_Lane3$CMO_maxID)
table(seurat_Lane4$CMO_maxID)
table(seurat_Lane5$CMO_maxID)

```

**Explanation:**

-   **CMO_maxID** indicates the CMO with the highest signal in each
    cell.

### Assignment of HTODemux

```{r}
# ==============================================================================
# View Final Sample Assignments (hash.ID)
# ==============================================================================
# hash.ID: Final assignment from HTODemux
# - For Singlets: Contains the specific CMO ID (CMO301-CMO312)
# - For Doublets: "Doublet"
# - For Negatives: "Negative"

# Display hash.ID distribution for each lane
table(seurat_Lane1$hash.ID)
table(seurat_Lane2$hash.ID)
table(seurat_Lane3$hash.ID)
table(seurat_Lane4$hash.ID)
table(seurat_Lane5$hash.ID)

```

**Explanation:**

-   **hash.ID** contains the specific CMO assignment for each cell or
    indicates if it's a Doublet or Negative.

#### Merged Results Summary

We can create a summary table to compare cell counts and assignments
across lanes.

**Example Table:**

| Parameter       | Lane 1 | Lane 2 | Lane 3 | Lane 4 | Lane 5 |
|-----------------|--------|--------|--------|--------|--------|
| Number of Cells | 31,833 | 26,969 | 26,951 | 28,537 | 30,214 |
| Singlets        | 17,294 | 17,252 | 17,902 | 19,034 | 20,151 |
| Doublets        | 11,566 | 6,973  | 6,620  | 6,797  | 7,429  |
| Negatives       | 2,973  | 2,744  | 2,429  | 2,706  | 2,634  |

**Explanation:**

-   This summary provides an overview of the demultiplexing results for
    each lane.

# ==============================================================================
# SECTION 11: VISUALIZATION - HEATMAPS
# ==============================================================================

### Visualizations

We will generate heatmaps and violin plots to visualize the
demultiplexing results.

#### Heatmaps

```{r}
# ==============================================================================
# Generate HTODemux Heatmaps
# ==============================================================================
# HTOHeatmap(): Visualizes normalized CMO signals across cells
# Parameters:
# - assay = "CMO": Use the CMO assay data
# - ncells = 10000: Randomly sample 10,000 cells for visualization
# Cells are grouped by their classification, showing CMO signal patterns

# Generate heatmap for Lane 1 showing CMO signal distribution
HTOHeatmap(seurat_Lane1, assay = "CMO", ncells = 10000) + ggtitle("Lane 1")

# Generate heatmap for Lane 2
HTOHeatmap(seurat_Lane2, assay = "CMO", ncells = 10000) + ggtitle("Lane 2")

# Generate heatmap for Lane 3
HTOHeatmap(seurat_Lane3, assay = "CMO", ncells = 10000) + ggtitle("Lane 3")

# Generate heatmap for Lane 4
HTOHeatmap(seurat_Lane4, assay = "CMO", ncells = 10000) + ggtitle("Lane 4")

# Generate heatmap for Lane 5
HTOHeatmap(seurat_Lane5, assay = "CMO", ncells = 10000) + ggtitle("Lane 5")

```

**Explanation:**

-   **HTOHeatmap** visualizes the normalized CMO signals across cells,
    helping to assess the quality of demultiplexing.

# ==============================================================================
# SECTION 12: VISUALIZATION - VIOLIN PLOTS
# ==============================================================================

#### Violin Plots

```{r}
# ==============================================================================
# Generate Quality Control Violin Plots
# ==============================================================================
# VlnPlot(): Creates violin plots of specified features
# Parameters:
# - features: Metadata columns to plot (nCount_RNA, nFeature_RNA)
#   - nCount_RNA: Total UMI counts per cell
#   - nFeature_RNA: Number of detected genes per cell
# - pt.size = 0.1: Size of individual data points
# - log = TRUE: Log-transform the y-axis for better visualization
# - ncol = 2: Arrange plots in 2 columns
# Cells are grouped by their CMO classification

# Generate violin plots for Lane 1
VlnPlot(seurat_Lane1, 
        features = c("nCount_RNA", "nFeature_RNA"), 
        pt.size = 0.1, 
        log = TRUE, 
        ncol = 2) + NoLegend()

# Repeat for other lanes (Lane 2 to Lane 5)

```

**Explanation:**

-   Violin plots show the distribution of RNA counts and detected
    features across different classifications (Singlets, Doublets,
    Negatives).

# ==============================================================================
# SECTION 13: DIMENSIONALITY REDUCTION - CMO-BASED UMAP
# ==============================================================================

#### UMAPs

We will visualize the results using UMAP dimensionality reduction.

##### CMO-Based UMAP

```{r}
# ==============================================================================
# Perform Dimensionality Reduction on CMO Data
# ==============================================================================
# This creates a UMAP visualization based on CMO signal patterns
# Helps identify how well samples separate based on their multiplexing oligos

# Set CMO as the default assay for subsequent operations
DefaultAssay(seurat_Lane1) <- "CMO"

# ScaleData(): Scale all CMO features (CMO301-CMO312)
# rownames(seurat_Lane1): Gets all feature names in the current assay
# verbose = FALSE: Suppress progress messages
seurat_Lane1 <- ScaleData(seurat_Lane1, 
                           features = rownames(seurat_Lane1), 
                           verbose = FALSE)

# RunPCA(): Perform Principal Component Analysis on CMO data
# approx = FALSE: Use exact PCA computation (more accurate for small feature sets)
seurat_Lane1 <- RunPCA(seurat_Lane1, 
                        features = rownames(seurat_Lane1), 
                        approx = FALSE)

# FindNeighbors(): Compute k-nearest neighbors graph
# reduction = "pca": Use PCA coordinates
# dims = 1:10: Use first 10 principal components
seurat_Lane1 <- FindNeighbors(seurat_Lane1, 
                               reduction = "pca", 
                               dims = 1:10)

# FindClusters(): Identify cell clusters using Louvain algorithm
# resolution = 0.6: Controls cluster granularity (higher = more clusters)
seurat_Lane1 <- FindClusters(seurat_Lane1, 
                              resolution = 0.6, 
                              verbose = FALSE)

# RunUMAP(): Generate UMAP embedding for visualization
# reduction.name = "cmo_umap": Name this UMAP specifically for CMO data
seurat_Lane1 <- RunUMAP(seurat_Lane1, 
                         reduction = "pca", 
                         dims = 1:10, 
                         reduction.name = "cmo_umap")

# ==============================================================================
# REPEAT FOR LANES 2-5
# ==============================================================================
# Apply the same dimensional reduction steps to seurat_Lane2 through seurat_Lane5

```

**Explanation:**

-   We perform PCA and UMAP on the CMO data to visualize how cells
    cluster based on their CMO profiles.

##### Visualization of CMO-Based UMAP

```{r}
# ==============================================================================
# Visualize CMO-Based UMAP Colored by Sample Assignment
# ==============================================================================
# DimPlot(): Create dimensional reduction plot
# reduction = "cmo_umap": Use the CMO-based UMAP coordinates
# group.by = "hash.ID": Color cells by their sample assignment

# Generate UMAP plot for Lane 1 colored by hash.ID
DimPlot(seurat_Lane1, 
        reduction = "cmo_umap", 
        group.by = "hash.ID") + ggtitle("Lane 1 hash.ID")

# Repeat for other lanes (Lane 2 to Lane 5)

```

# ==============================================================================
# SECTION 14: DIMENSIONALITY REDUCTION - RNA-BASED UMAP
# ==============================================================================

##### RNA-Based UMAP

We will process the RNA data to get UMAP plots.

```{r}
# ==============================================================================
# Perform Dimensionality Reduction on RNA Data
# ==============================================================================
# This creates a UMAP visualization based on gene expression patterns
# Shows how cells cluster based on their transcriptional profiles

# Set RNA as the default assay
DefaultAssay(seurat_Lane1) <- "RNA"

# NormalizeData(): Log-normalize the RNA counts
seurat_Lane1 <- NormalizeData(seurat_Lane1)

# FindVariableFeatures(): Identify highly variable genes
# Uses default method (vst) and default number of features (2000)
seurat_Lane1 <- FindVariableFeatures(seurat_Lane1)

# ScaleData(): Center and scale expression values for variable features
seurat_Lane1 <- ScaleData(seurat_Lane1, 
                           features = VariableFeatures(seurat_Lane1))

# RunPCA(): Perform PCA on scaled data
seurat_Lane1 <- RunPCA(seurat_Lane1)

# FindNeighbors(): Build nearest neighbor graph using PCA
# dims = 1:10: Use first 10 principal components
seurat_Lane1 <- FindNeighbors(seurat_Lane1, 
                               reduction = "pca", 
                               dims = 1:10)

# FindClusters(): Cluster cells based on their expression profiles
seurat_Lane1 <- FindClusters(seurat_Lane1, 
                              resolution = 0.6, 
                              verbose = FALSE)

# RunUMAP(): Generate UMAP for RNA data
# reduction.name = "rna_umap": Name this UMAP specifically for RNA data
seurat_Lane1 <- RunUMAP(seurat_Lane1, 
                         reduction = "pca", 
                         dims = 1:10, 
                         reduction.name = "rna_umap")

```

**Explanation:**

-   We perform PCA and UMAP on the RNA data to visualize the
    transcriptional landscape of the cells.

##### Visualization of RNA-Based UMAP

```{r}
# ==============================================================================
# Visualize RNA-Based UMAP with Various Groupings
# ==============================================================================

# Plot 1: Color by Seurat clusters (transcriptional clusters)
# label = TRUE: Show cluster numbers on the plot
DimPlot(seurat_Lane1, 
        reduction = "rna_umap", 
        group.by = "seurat_clusters", 
        label = TRUE) + ggtitle("Lane 1 Clusters")

# Plot 2: Color by CMO classification (Singlet/Doublet/Negative)
# Shows how demultiplexing results map onto transcriptional space
DimPlot(seurat_Lane1, 
        reduction = "rna_umap", 
        group.by = "CMO_classification.global") + ggtitle("Lane 1 CMO Classification")

# Repeat for other lanes (Lane 2 to Lane 5)

```

**Explanation:**

-   These plots help identify how demultiplexing results correspond to
    transcriptional clusters.

# ==============================================================================
# SECTION 15: CLUSTER-LEVEL DEMULTIPLEXING ANALYSIS
# ==============================================================================

#### Bar Plots with Classifications of HTODemux per Cluster

We will create bar plots to show the proportion of different
classifications within each cluster.

```{r}
# ==============================================================================
# Create Stacked Bar Plot of Classification per Cluster
# ==============================================================================

# Extract relevant metadata into a data frame
# seurat_clusters: Cluster assignments from RNA-based clustering
# CMO_classification.global: Demultiplexing classification
df_lane1 <- data.frame(
  seurat_clusters = seurat_Lane1$seurat_clusters,
  CMO_classification.global = seurat_Lane1$CMO_classification.global
)

# Calculate percentages using dplyr pipeline
df_lane1 <- df_lane1 %>%
  # Group by both cluster and classification
  group_by(seurat_clusters, CMO_classification.global) %>%
  # Count cells in each group
  summarise(Nb = n()) %>%
  # Calculate total cells per cluster
  mutate(C = sum(Nb)) %>%
  # Calculate percentage within each cluster
  mutate(Percent = Nb / C * 100)

# Generate stacked bar plot using ggplot2
ggplot(df_lane1, 
       aes(fill = CMO_classification.global,  # Color by classification
           y = Percent,                         # Y-axis shows percentage
           x = seurat_clusters)) +              # X-axis shows clusters
  # position = "fill": Stack bars to 100%
  # stat = "identity": Use actual values (not counts)
  geom_bar(position = "fill", stat = "identity") +
  # Customize theme: rotate x-axis labels for readability
  theme(legend.position = 'right', 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggtitle("Lane 1")

```

**Explanation:**

-   This plot shows the distribution of Singlets, Doublets, and
    Negatives within each cluster.

# ==============================================================================
# SECTION 16: CELLRANGER VS HTODEMUX COMPARISON
# ==============================================================================

### Cell Ranger and Comparison with HTODemux()

We will compare the demultiplexing results from CellRanger with those
from HTODemux.

#### Transferring CellRanger Demultiplexing Data to Seurat Objects

```{r}
# ==============================================================================
# Add CellRanger Demultiplexing Results to Seurat Metadata
# ==============================================================================

# Extract relevant columns from CellRanger confidence table
# Columns: Multiplet score, Blank score, Assignment, Assignment Probability
df_for_transfer <- confidence_tb_cellranger$CellRanger_L1[, 
  c("Multiplet", "Blank", "Assignment", "Assignment Probability")]

# Rename columns for clarity when stored in Seurat metadata
colnames(df_for_transfer) <- c("CellRanger_Multiplet_score", 
                                "CellRanger_Blank_score", 
                                "CellRanger_Assignment", 
                                "Assignment_Probability")

# Set cell barcodes as row names (required for AddMetaData)
rownames(df_for_transfer) <- confidence_tb_cellranger$CellRanger_L1$Barcode

# AddMetaData(): Adds new metadata columns to Seurat object
# Matches cells by row names (barcodes)
seurat_Lane1 <- AddMetaData(object = seurat_Lane1, 
                             metadata = df_for_transfer)

# ==============================================================================
# REPEAT FOR LANES 2-5
# ==============================================================================
# Apply the same metadata transfer for seurat_Lane2 through seurat_Lane5

```

**Explanation:**

-   We add the CellRanger demultiplexing results as metadata to the
    Seurat objects for comparison.

# ==============================================================================
# SECTION 17: VENN DIAGRAM COMPARISON
# ==============================================================================

#### Venn Diagrams to Compare Assignments

```{r}
# ==============================================================================
# Compare Singlet Assignments Between HTODemux and CellRanger
# ==============================================================================

# Create a list containing singlet cell barcodes from each method
singlets_L1 <- list(
  # HTODemux singlets: Cells classified as "Singlet" by HTODemux
  # rownames() extracts cell barcodes from metadata subset
  HTODemux_Singlet = rownames(
    seurat_Lane1@meta.data[seurat_Lane1$CMO_classification.global == "Singlet",]
  ),
  
  # CellRanger singlets: Cells assigned as "singlet" by CellRanger
  # Note: CellRanger uses lowercase "singlet"
  CellRanger_singlet = rownames(
    seurat_Lane1@meta.data[seurat_Lane1$CellRanger_Assignment == "singlet",]
  )
)

# Generate Venn diagram using ggvenn
# fill_color: Colors for each set
# stroke_size: Line thickness
# set_name_size: Font size for set labels
ggvenn(singlets_L1, 
       fill_color = c("#0073C2FF", "#EFC000FF"), 
       stroke_size = 0.5, 
       set_name_size = 5)

```

**Explanation:**

-   The Venn diagram visualizes the overlap between Singlet assignments
    from HTODemux and CellRanger.

# ==============================================================================
# SECTION 18: MERGE ALL LANES
# ==============================================================================

### Merging Datasets

We will merge all lanes into a single Seurat object for comprehensive
analysis.

```{r}
# ==============================================================================
# Merge All Seurat Objects Into One Combined Object
# ==============================================================================

# merge(): Combines multiple Seurat objects
# x: First object to merge
# y: List of additional objects to merge
# Cell barcodes are made unique by appending object identifiers
seurat_Lane0 <- merge(x = seurat_Lane1, 
                       y = list(seurat_Lane2, 
                               seurat_Lane3, 
                               seurat_Lane4, 
                               seurat_Lane5))

# ==============================================================================
# Process Merged Dataset
# ==============================================================================

# Normalize the merged data
seurat_Lane0 <- NormalizeData(seurat_Lane0)

# Find variable features across all cells
seurat_Lane0 <- FindVariableFeatures(seurat_Lane0)

# Scale the variable features
seurat_Lane0 <- ScaleData(seurat_Lane0, 
                           features = VariableFeatures(seurat_Lane0))

# ==============================================================================
# Dimensionality Reduction on Merged Data
# ==============================================================================

# Perform PCA on merged dataset
seurat_Lane0 <- RunPCA(seurat_Lane0)

# Build neighbor graph using more PCs (25) for larger dataset
seurat_Lane0 <- FindNeighbors(seurat_Lane0, 
                               reduction = "pca", 
                               dims = 1:25)

# Cluster with lower resolution (0.4) for broader groupings
seurat_Lane0 <- FindClusters(seurat_Lane0, 
                              resolution = 0.4, 
                              verbose = FALSE)

# Generate UMAP for merged data
# reduction.name = "rna_umap_merged": Distinguishes from lane-specific UMAPs
seurat_Lane0 <- RunUMAP(seurat_Lane0, 
                         reduction = "pca", 
                         dims = 1:25, 
                         reduction.name = "rna_umap_merged")

```

**Explanation:**

-   We process the merged dataset similarly to individual lanes for
    integrated analysis.

# ==============================================================================
# SECTION 19: CELL CYCLE SCORING
# ==============================================================================

### Cell Cycle Scoring

```{r}
# ==============================================================================
# Prepare Cell Cycle Gene Lists
# ==============================================================================

# Access Seurat's built-in cell cycle gene lists (human genes)
# cc.genes.updated.2019: Updated version of cell cycle gene markers
# s.genes: Genes associated with S phase (DNA synthesis)
# g2m.genes: Genes associated with G2/M phase (cell division)
s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

# gorth(): Convert gene symbols between organisms (gprofiler2 function)
# source_organism = 'hsapiens': Input genes are human
# target_organism = 'mmusculus': Convert to mouse orthologs
# $ortholog_name: Extract the mouse gene symbols
mmus_s <- gorth(s.genes, 
                source_organism = 'hsapiens', 
                target_organism = 'mmusculus')$ortholog_name

mmus_g2m <- gorth(g2m.genes, 
                  source_organism = 'hsapiens', 
                  target_organism = 'mmusculus')$ortholog_name

# ==============================================================================
# Score Cells for Cell Cycle Phase
# ==============================================================================

# Set RNA assay as default (cell cycle scoring uses RNA expression)
DefaultAssay(seurat_Lane0) <- "RNA"

# CellCycleScoring(): Assigns cell cycle phase to each cell
# s.features: Genes marking S phase
# g2m.features: Genes marking G2/M phase
# set.ident = TRUE: Set cell identities to cell cycle phase
# Adds metadata: S.Score, G2M.Score, Phase (G1, S, or G2M)
seurat_Lane0 <- CellCycleScoring(seurat_Lane0, 
                                  s.features = mmus_s, 
                                  g2m.features = mmus_g2m, 
                                  set.ident = TRUE)

```

**Explanation:**

-   We score cells based on their cell cycle phase, which can be
    important for interpreting clusters.

# ==============================================================================
# SECTION 20: FINAL VISUALIZATION OF MERGED DATA
# ==============================================================================

### Visualization of Merged Data

```{r}
# ==============================================================================
# Generate UMAP Plots for Merged Dataset
# ==============================================================================

# Plot 1: Color by CellRanger demultiplexing assignment
# Shows distribution of sample assignments in merged UMAP space
DimPlot(seurat_Lane0, 
        reduction = "rna_umap_merged", 
        group.by = "CellRanger_Assignment") + ggtitle("CellRanger Assignment")

# Plot 2: Color by HTODemux assignment (hash.ID)
# Allows comparison with CellRanger assignments
DimPlot(seurat_Lane0, 
        reduction = "rna_umap_merged", 
        group.by = "hash.ID") + ggtitle("HTODemux Assignment")

# Plot 3: Color by cell cycle phase
# Shows distribution of G1, S, and G2M cells across clusters
DimPlot(seurat_Lane0, 
        reduction = "rna_umap_merged", 
        group.by = "Phase") + ggtitle("Cell Cycle Phase")

```

**Explanation:**

-   These plots help visualize how cells from different lanes and
    classifications are distributed in the integrated dataset.

# Conclusions from the Analysis

-   **Singlet Ratios:** Lanes 2-5 have similar singlet ratios
    (\~62-66%), with both HTODemux and CellRanger showing comparable
    results. Lane 1 has a lower singlet ratio and a higher difference
    between methods, possibly due to stricter parameters in CellRanger.

-   **Cell Types with High Doublets or Negatives:**

    -   **Naive Pluripotent Embryonic Stem Cells** have a high
        proportion of Negative cells.

    -   **Erythroid Cells** have a high proportion of Doublets.

    -   Certain clusters are enriched for Doublets or Negatives,
        indicating specific cell types may be more prone to multiplets
        or have issues with labeling.

-   **Lane-Specific Observations:**

    -   **Lane 1:** Clusters 0 and 14 are enriched for Doublets; Cluster
        8 has a high number of Negatives and represents Naive ES cells.

    -   **Other Lanes:** Similar patterns are observed with specific
        clusters enriched for Doublets or Negatives, corresponding to
        certain cell types.

# ==============================================================================
# SECTION 21: SESSION INFORMATION
# ==============================================================================

# Session Info

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================
# sessionInfo(): Displays R version, platform, and all loaded packages
# Essential for reproducibility - others can match your exact environment
sessionInfo()
```

**Explanation:**

-   Including session information helps others reproduce the analysis by
    knowing the package versions and environment details.

# Extras

|        | Lane 1 | Lane 2 | Lane 3 | Lane 4 | Lane 5 |
|--------|--------|--------|--------|--------|--------|
| CMO301 | 66     | 105    | 93     | 56     | 50     |
| CMO302 | 89     | 164    | 156    | 94     | 84     |
| CMO303 | 107    | 212    | 136    | 109    | 108    |
| CMO304 | 74     | 124    | 102    | 90     | 81     |
| CMO305 | 78     | 163    | 105    | 102    | 108    |
| CMO306 | 156    | 302    | 163    | 108    | 107    |
| CMO307 | 135    | 182    | 167    | 109    | 110    |
| CMO308 | 110    | 222    | 154    | 128    | 79     |
| CMO309 | 191    | 163    | 165    | 131    | 127    |
| CMO310 | 99     | 155    | 107    | 92     | 111    |
| CMO311 | 112    | 244    | 201    | 172    | 127    |
| CMO312 | 275    | 131    | 149    | 99     | 98     |

: Cutoffs for CMOs per Lane.

|        | Lane1 | Lane2 | Lane3 | Lane4 | Lane5 |
|--------|-------|-------|-------|-------|-------|
| CMO301 | 3975  | 2710  | 2408  | 2560  | 2692  |
| CMO302 | 2939  | 2335  | 2416  | 2523  | 2728  |
| CMO303 | 2734  | 2287  | 2187  | 2357  | 2507  |
| CMO304 | 2994  | 2130  | 2159  | 2221  | 2396  |
| CMO305 | 2385  | 2086  | 2070  | 2206  | 2343  |
| CMO306 | 2155  | 2055  | 2156  | 2251  | 2415  |
| CMO307 | 2173  | 2132  | 2165  | 2294  | 2390  |
| CMO308 | 2412  | 2277  | 2208  | 2382  | 2598  |
| CMO309 | 2317  | 2305  | 2242  | 2347  | 2546  |
| CMO310 | 2718  | 2286  | 2341  | 2421  | 2549  |
| CMO311 | 3104  | 2575  | 2751  | 3012  | 2947  |
| CMO312 | 1927  | 1791  | 1848  | 1963  | 2103  |

: Highest CMO per cell.

+----------+----------+----------+----------+----------+----------+
| hash.ID  | seur     | seur     | seur     | seur     | seur     |
|          | at_Lane1 | at_Lane2 | at_Lane3 | at_Lane4 | at_Lane5 |
+==========+==========+==========+==========+==========+==========+
| Doublet  | 11566    | 6973     | 6620     | 6797     | 7429     |
+----------+----------+----------+----------+----------+----------+
| CMO306   | 1391     | 1504     | 1453     | 1599     | 1463     |
+----------+----------+----------+----------+----------+----------+
| CMO312   | 1137     | 1403     | 1309     | 1314     | 1391     |
+----------+----------+----------+----------+----------+----------+
| CMO308   | 1536     | 1743     | 1350     | 2046     | 1893     |
+----------+----------+----------+----------+----------+----------+
| CMO303   | 1348     | 1310     | 1332     | 1708     | 1534     |
+----------+----------+----------+----------+----------+----------+
| CMO304   | 1236     | 1345     | 1314     | 1424     | 1422     |
+----------+----------+----------+----------+----------+----------+
| CMO307   | 1445     | 1574     | 1497     | 1622     | 1709     |
+----------+----------+----------+----------+----------+----------+
| CMO311   | 1876     | 1438     | 1542     | 1599     | 1728     |
+----------+----------+----------+----------+----------+----------+
| CMO309   | 1549     | 1482     | 1558     | 1682     | 1756     |
+----------+----------+----------+----------+----------+----------+
| CMO305   | 1307     | 1270     | 1314     | 1414     | 1659     |
+----------+----------+----------+----------+----------+----------+
| CMO302   | 1517     | 1213     | 1567     | 1625     | 2151     |
+----------+----------+----------+----------+----------+----------+
| CMO310   | 1462     | 1406     | 2429     | 1493     | 1600     |
+----------+----------+----------+----------+----------+----------+
| CMO301   | 1490     | 1564     | 1514     | 1543     | 1845     |
+----------+----------+----------+----------+----------+----------+
| Negative | 2973     | 2744     | 2429     | 2706     | 2634     |
+----------+----------+----------+----------+----------+----------+

: HTODemux(), hash.ID

#### Merged results

+----------+----------+----------+----------+----------+----------+
| P        | seur     | seur     | seur     | seur     | seur     |
| arameter | at_Lane1 | at_Lane2 | at_Lane3 | at_Lane4 | at_Lane5 |
+==========+==========+==========+==========+==========+==========+
| Amount   | 3065453  | 3029894  | 3216261  | 3305265  | 3010848  |
| of Cells |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Cells in | 31833    | 26969    | 26951    | 28537    | 30214    |
| Seurat   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Doublet  | 11566    | 6973     | 6620     | 6797     | 7429     |
+----------+----------+----------+----------+----------+----------+
| CMO306   | 1391     | 1504     | 1453     | 1599     | 1463     |
+----------+----------+----------+----------+----------+----------+
| CMO312   | 1137     | 1403     | 1309     | 1314     | 1391     |
+----------+----------+----------+----------+----------+----------+
| CMO308   | 1536     | 1743     | 1350     | 2046     | 1893     |
+----------+----------+----------+----------+----------+----------+
| CMO303   | 1348     | 1310     | 1332     | 1708     | 1534     |
+----------+----------+----------+----------+----------+----------+
| CMO304   | 1236     | 1345     | 1314     | 1424     | 1422     |
+----------+----------+----------+----------+----------+----------+
| CMO307   | 1445     | 1574     | 1497     | 1622     | 1709     |
+----------+----------+----------+----------+----------+----------+
| CMO311   | 1876     | 1438     | 1542     | 1599     | 1728     |
+----------+----------+----------+----------+----------+----------+
| CMO309   | 1549     | 1482     | 1558     | 1682     | 1756     |
+----------+----------+----------+----------+----------+----------+
| CMO305   | 1307     | 1270     | 1314     | 1414     | 1659     |
+----------+----------+----------+----------+----------+----------+
| CMO302   | 1517     | 1213     | 1567     | 1625     | 2151     |
+----------+----------+----------+----------+----------+----------+
| CMO310   | 1462     | 1406     | 2429     | 1493     | 1600     |
+----------+----------+----------+----------+----------+----------+
| CMO301   | 1490     | 1564     | 1514     | 1543     | 1845     |
+----------+----------+----------+----------+----------+----------+
| Negative | 2973     | 2744     | 2429     | 2706     | 2634     |
+----------+----------+----------+----------+----------+----------+
| Lane 1   | 66       | 105      | 93       | 56       | 50       |
| Cutoff   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Lane 2   | 89       | 164      | 156      | 94       | 84       |
| Cutoff   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Lane 3   | 107      | 212      | 136      | 109      | 108      |
| Cutoff   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Lane 4   | 74       | 124      | 102      | 90       | 81       |
| Cutoff   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
| Lane 5   | 78       | 163      | 105      | 102      | 108      |
| Cutoff   |          |          |          |          |          |
+----------+----------+----------+----------+----------+----------+
