---
title: "Unique Clusters Marker Search"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP
# ==============================================================================

```{r}
# ==============================================================================
# Load Required Package
# ==============================================================================

library(Seurat)  # Single-cell analysis toolkit
```


We observed there are three small clusters/populations. Due to limitations of the cell numbers, we cannot get proper DESeq2 results from pseudbulked data. For this reason we will use seurat for finding markers.

# ==============================================================================
# SECTION 2: MARKER ANALYSIS - MESODERM_IFIT1 (baIfit1)
# ==============================================================================

# baIfit1

```{r}
# ==============================================================================
# Set Identity and Find Markers for Mesoderm_Ifit1
# ==============================================================================

# Set active identity to cluster level (cl_2)
s_cmo$ident <- s_cmo$cl_2
Idents(s_cmo) <- "cl_2"

# FindMarkers(): Identify differentially expressed genes between clusters
# ident.1 = "Mesoderm_Ifit1": Target cluster (positive logFC = higher in this cluster)
# ident.2 = c("Mesoderm_Hoxb","Mesoderm_Hoxa"): Reference clusters for comparison
# min.pct = 0: No minimum percentage filter (include all genes)
# logfc.threshold = 0: No fold change filter (include all genes)
# Using no thresholds to get complete results for downstream analysis (GSEA, etc.)
baIfit1_cmo_CvsN_3 <- FindMarkers(s_cmo, 
                                   ident.1 = "Mesoderm_Ifit1", 
                                   ident.2 = c("Mesoderm_Hoxb", "Mesoderm_Hoxa"),
                                   min.pct = 0, 
                                   logfc.threshold = 0)
```

```{r}
# ==============================================================================
# Handle Zero P-values
# ==============================================================================

# Replace exact zero p-values with very small number
# Zero p-values occur when Wilcoxon test returns exact zeros
# 1e-300: Smallest practical value for -log10 transformation
# This prevents infinite values in downstream calculations

# Replace adjusted p-values of 0
baIfit1_cmo_CvsN_3$p_val_adj[baIfit1_cmo_CvsN_3$p_val_adj == 0] <- 1e-300

# Replace raw p-values of 0
baIfit1_cmo_CvsN_3$p_val[baIfit1_cmo_CvsN_3$p_val == 0] <- 1e-300
```

## adding ranking column (which partially tries to mimic the stat values from DESeq2)

```{r}
# ==============================================================================
# Create DESeq2-like Statistic for Ranking
# ==============================================================================

# statlike: Signed -log10(p-value) similar to DESeq2 stat column
# sign(avg_log2FC): +1 for upregulated, -1 for downregulated
# -log10(p_val): Significance magnitude (higher = more significant)
# Result: Positive values = significantly upregulated in ident.1
#         Negative values = significantly downregulated in ident.1
# Useful for GSEA preranked analysis
baIfit1_cmo_CvsN_3$statlike <- sign(baIfit1_cmo_CvsN_3$avg_log2FC) * (-log10(baIfit1_cmo_CvsN_3$p_val))
```

# ==============================================================================
# SECTION 3: MARKER ANALYSIS - ENDRSAD2
# ==============================================================================

# EndRsad2

```{r}
# ==============================================================================
# Set Identity and Find Markers for EndRsad2
# ==============================================================================

# Switch to subcluster level (cl_3) for endothelial subclusters
s_cmo$ident <- s_cmo$cl_3
Idents(s_cmo) <- "cl_3"

# Compare EndRsad2 against other endothelial subclusters
# EndRsad2: Endothelial cells expressing Rsad2 (interferon-stimulated gene)
# Reference: EndGadd45g, EndCdh5, EndCyp26b1 (other endothelial populations)
EndRsad2_cmo_CvsN_3 <- FindMarkers(s_cmo, 
                                    ident.1 = "EndRsad2", 
                                    ident.2 = c("EndGadd45g", "EndCdh5", "EndCyp26b1"),
                                    min.pct = 0, 
                                    logfc.threshold = 0)
```

```{r}
# ==============================================================================
# Handle Zero P-values
# ==============================================================================

# Replace adjusted p-values of 0
EndRsad2_cmo_CvsN_3$p_val_adj[EndRsad2_cmo_CvsN_3$p_val_adj == 0] <- 1e-300

# Replace raw p-values of 0
EndRsad2_cmo_CvsN_3$p_val[EndRsad2_cmo_CvsN_3$p_val == 0] <- 1e-300
```

## adding ranking column (which partially tries to mimic the stat values from DESeq2)

```{r}
# ==============================================================================
# Create DESeq2-like Statistic
# ==============================================================================

EndRsad2_cmo_CvsN_3$statlike <- sign(EndRsad2_cmo_CvsN_3$avg_log2FC) * (-log10(EndRsad2_cmo_CvsN_3$p_val))
```

# ==============================================================================
# SECTION 4: MARKER ANALYSIS - ENDCYP26B1
# ==============================================================================

# EndCyp26b1

```{r}
# ==============================================================================
# Set Identity and Find Markers for EndCyp26b1
# ==============================================================================

# Ensure subcluster identity is set
s_cmo$ident <- s_cmo$cl_3
Idents(s_cmo) <- "cl_3"

# Compare EndCyp26b1 against other endothelial subclusters
# EndCyp26b1: Endothelial cells expressing Cyp26b1 (retinoic acid metabolism)
# Reference: EndGadd45g, EndCdh5, EndRsad2 (other endothelial populations)
EndCyp26b1_cmo_CvsN_3 <- FindMarkers(s_cmo, 
                                      ident.1 = "EndCyp26b1", 
                                      ident.2 = c("EndGadd45g", "EndCdh5", "EndRsad2"),
                                      min.pct = 0, 
                                      logfc.threshold = 0)
```

```{r}
# ==============================================================================
# Handle Zero P-values
# ==============================================================================

# Replace adjusted p-values of 0
EndCyp26b1_cmo_CvsN_3$p_val_adj[EndCyp26b1_cmo_CvsN_3$p_val_adj == 0] <- 1e-300

# Replace raw p-values of 0
EndCyp26b1_cmo_CvsN_3$p_val[EndCyp26b1_cmo_CvsN_3$p_val == 0] <- 1e-300
```

```{r}
# ==============================================================================
# Create DESeq2-like Statistic
# ==============================================================================

EndCyp26b1_cmo_CvsN_3$statlike <- sign(EndCyp26b1_cmo_CvsN_3$avg_log2FC) * (-log10(EndCyp26b1_cmo_CvsN_3$p_val))
```

# ==============================================================================
# SECTION 5: SESSION INFORMATION
# ==============================================================================

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================

sessionInfo()
```
