---
title: "DAA-miloR"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

Differential abundance analysis with miloR is applied for each ko and control pair for each group (Early Differentiation, Late Mesoderm, Hematopoietic).

# ==============================================================================
# SECTION 1: COLOR PALETTES FOR VISUALIZATION
# ==============================================================================

```{r eval=FALSE,}
# ==============================================================================
# Define Custom Color Palettes
# ==============================================================================
# r3dcol: A list containing multiple color palette options for miloR
# visualization and differential abundance plots

# Initialize empty list to store all color palettes
r3dcol <- list()

# cols_94: Large palette with 94 distinct colors for many neighborhoods
r3dcol$cols_94 <- c(
  "#F0E442","#1CE6FF","#FF34FF","#FF4A46","#008941","#006FA6","#A30059","#FFDBE5","#7A4900","#0089A3",
  "#63FFAC","#B79762",
  "#004D43","#8FB0FF","#997D87","#809693","#6A3A4C","#b4b701","#4FC601","#3B5DFF","#FF2F80","#61615A",
  "#BA0900","#6B7900",
  "#00C2A0","#FFAA92","#FF90C9","#B903AA","#D16100","#922329","#7B4F4B","#A1C299","#0AA6D8","#00846F",
  "#FFB500","#C2FFED",
  "#A079BF","#CC0744","#C0B9B2","#C2FF99","#00489C","#6F0062","#0CBD66","#EEC3FF","#456D75","#B77B68",
  "#7A87A1","#788D66",
  "#885578","#FAD09F","#FF8A9A","#D157A0","#BEC459","#456648","#0086ED","#886F4C","#34362D","#B4A8BD",
  "#00A6AA","#452C2C",
  "#636375","#A3C8C9","#FF913F","#938A81","#575329","#00FECF","#B05B6F","#8CD0FF","#3B9700","#04F757",
  "#C8A1A1","#1E6E00",
  "#7900D7","#A77500","#6367A9","#A05837","#6B002C","#772600","#D790FF","#9B9700","#549E79","#FFF69F",
  "#201625","#72418F",
  "#BC23FF","#99ADC0","#3A2465","#DDEFFF","#5B4534","#FDE8DC","#404E55","#CB7E98","#A4E804","#324E72"
)

# cols_46: Medium palette with 46 colors
r3dcol$cols_46 <- c(
  "#FFB500","#00C2A0","#D157A0","#8CD0FF","#FF4A46","#FDE8DC","#63FFAC","#B79762","#8FB0FF","#997D87",
  "#809693","#b4b701",
  "#4FC601","#FFAA92","#FF90C9","#A1C299","#F0E442","#C2FFED","#A079BF","#C0B9B2","#C2FF99","#EEC3FF",
  "#B77B68","#7A87A1",
  "#788D66","#FAD09F","#FF8A9A","#BEC459","#B4A8BD","#A3C8C9","#FF913F","#938A81","#00FECF","#FF34FF",
  "#1CE6FF","#04F757",
  "#C8A1A1","#D790FF","#9B9700","#549E79","#FFF69F","#99ADC0","#DDEFFF","#FFDBE5","#CB7E98","#A4E804"
)

# cols_pjsala: Palette inspired by Pijuan-Sala et al. mouse gastrulation atlas
r3dcol$cols_pjsala <- c(
  "#005579","#0F4A9C","#139992","#1A1A1A","#354E23","#3F84AA","#532C8A","#635547","#647a4f","#65A83E",
  "#7F6874","#8870ad",
  "#8DB5CE","#8EC792","#989898","#9e6762","#B51D8D","#c19f70","#C3C388","#C594BF","#C72228","#c9a997",
  "#C9EBFB","#cc7818",
  "#CDE088","#DABE99","#DFCDE4","#EF4E22","#EF5A9D","#F397C0","#F6BFCB","#f79083","#f7f79e","#f9decf",
  "#FACB12","#FBBE92",
  "#ff891c"
)

# cols_4x3: 12 colors arranged as 4 color families x 3 shades
r3dcol$cols_4x3 <- c(
  "#551153","#970098","#9c24cd","#412c00","#997600","#c6a000","#1e3604","#355e07","#4f890e","#002f64",
  "#0050ae","#1077f3"
)

# cols_12x1: 12 distinct colors from ColorBrewer paired palette
r3dcol$cols_12x1 <- c(
  '#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a',
  '#ffff99','#b15928'
)

# cols_4x3b: Alternative 12-color palette
r3dcol$cols_4x3b <- c(
  "#551153","#970098","#cc34cd","#c85b00","#f98517","#fdbf6f","#00603d","#008c5c","#33b983","#002f64",
  "#0050ae","#1077f3"
)

# cols_4x1: 4 colors for condition comparisons
r3dcol$cols_4x1 <- c(
  "#970098","#997600","#4f890e","#0050ae"
)

# cols_4x1b: Alternative 4-color palette
r3dcol$cols_4x1b <- c(
  "#970098","#f98517","#008c5c","#1077f3"
)

# cols_3a: 3 colors for three-group comparisons
r3dcol$cols_3a <- c(
  "#D157A0","#FFB500","#00C2A0"
)

# cols_3b: Alternative 3-color palette
r3dcol$cols_3b <- c(
  "#D454B5","#c85b00","#008c5c"
)
```

# ==============================================================================
# SECTION 2: ENVIRONMENT SETUP AND PACKAGE LOADING
# ==============================================================================

```{r eval=FALSE}
# ==============================================================================
# Load Required Packages
# ==============================================================================

suppressPackageStartupMessages({
  library(SingleCellExperiment)  # Bioconductor container for scRNA-seq data
  library(miloR)                 # Neighborhood-based differential abundance analysis
  library(dplyr)                 # Data manipulation
  library(scater)                # Single-cell analysis toolkit (visualization)
  library(speckle)               # Alternative DAA method (for comparison)
  library(ggplot2)               # Visualization
})
```



# MiloR - Analysis

# ==============================================================================
# SECTION 3: EARLY DIFFERENTIATION - DATA LOADING AND MILO SETUP
# ==============================================================================

## Early

```{r eval=FALSE}
# ==============================================================================
# Load Seurat Subsets for Early Differentiation Group
# ==============================================================================

# Load pre-subsetted Seurat objects containing only Early differentiation cells
# Each file contains one KO condition paired with Control
early_ControlAtf3 <- readRDS("~/cmo_chapter/introduction/rds/Early/early_CA.RDS")
early_ControlBcl6b <- readRDS("~/cmo_chapter/introduction/rds/Early/early_BA.RDS")
early_ControlZfp711 <- readRDS("~/cmo_chapter/introduction/rds/Early/early_ZA.RDS")

# ==============================================================================
# Convert Seurat to SingleCellExperiment and Create Milo Objects
# ==============================================================================

# miloR requires SingleCellExperiment format, not Seurat
# as.SingleCellExperiment(): Convert Seurat object to SCE

# --- Atf3 vs Control ---
sceAtf3NT <- as.SingleCellExperiment(early_ControlAtf3)
# reducedDim(): Add PCA embeddings (must extract from Seurat reduction slot)
# pcaearly: The PCA reduction name specific to the Early subset
reducedDim(sceAtf3NT, "PCA") <- early_ControlAtf3@reductions$pcaearly@cell.embeddings
# Milo(): Create Milo object from SCE (extends SCE with neighborhood slots)
Atf3NT_milo <- Milo(sceAtf3NT)

# --- Bcl6b vs Control ---
sceBcl6bNT <- as.SingleCellExperiment(early_ControlBcl6b)
reducedDim(sceBcl6bNT, "PCA") <- early_ControlBcl6b@reductions$pcaearly@cell.embeddings
Bcl6bNT_milo <- Milo(sceBcl6bNT)

# --- Zfp711 vs Control ---
sceZfp711NT <- as.SingleCellExperiment(early_ControlZfp711)
reducedDim(sceZfp711NT, "PCA") <- early_ControlZfp711@reductions$pcaearly@cell.embeddings
Zfp711NT_milo <- Milo(sceZfp711NT)

# ==============================================================================
# Build k-Nearest Neighbor Graphs
# ==============================================================================

# buildGraph(): Construct k-NN graph for neighborhood definition
# k = 35: Number of nearest neighbors (higher = larger neighborhoods)
# d = 50: Number of PCA dimensions to use for distance calculation
Atf3NT_milo <- buildGraph(Atf3NT_milo, k = 35, d = 50)
Bcl6bNT_milo <- buildGraph(Bcl6bNT_milo, k = 35, d = 50)
Zfp711NT_milo <- buildGraph(Zfp711NT_milo, k = 35, d = 50)

# ==============================================================================
# Create Neighborhoods
# ==============================================================================

# makeNhoods(): Define neighborhoods for differential abundance testing
# prop = 0.3: Proportion of cells to sample as neighborhood centers
# k = 35: Same k as buildGraph for consistency
# d = 50: Same dimensions as buildGraph
# refined = TRUE: Use refined sampling for better neighborhood representation
Atf3NT_milo <- makeNhoods(Atf3NT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)
Bcl6bNT_milo <- makeNhoods(Bcl6bNT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)
Zfp711NT_milo <- makeNhoods(Zfp711NT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)

# ==============================================================================
# Visualize Neighborhood Size Distribution
# ==============================================================================

# plotNhoodSizeHist(): Histogram of neighborhood sizes
# Ideal: Peak around 5*N_samples (rule of thumb)
# Too small neighborhoods = underpowered tests
# Too large neighborhoods = loss of resolution
plotNhoodSizeHist(Atf3NT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
  ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                   size = 20, face = "bold.italic"))

plotNhoodSizeHist(Bcl6bNT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
  ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                    size = 20, face = "bold.italic"))

plotNhoodSizeHist(Zfp711NT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
  ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                     size = 20, face = "bold.italic"))

# ==============================================================================
# Prepare Sample Identifiers
# ==============================================================================

# Convert sample IDs to character (miloR requirement)
# colData@listData: Access column data as a list
Atf3NT_milo@colData@listData[["Samples"]] <- as.character(Atf3NT_milo@colData@listData$Samples_lv)
Bcl6bNT_milo@colData@listData[["Samples"]] <- as.character(Bcl6bNT_milo@colData@listData$Samples_lv)
Zfp711NT_milo@colData@listData[["Samples"]] <- as.character(Zfp711NT_milo@colData@listData$Samples_lv)

# ==============================================================================
# Count Cells Per Neighborhood Per Sample
# ==============================================================================

# countCells(): Count how many cells from each sample are in each neighborhood
# This creates the count matrix for differential abundance testing
# samples = "Samples": Column name in metadata containing sample identifiers
Atf3NT_milo <- countCells(Atf3NT_milo, meta.data = data.frame(colData(Atf3NT_milo)),
                          samples = "Samples")
Bcl6bNT_milo <- countCells(Bcl6bNT_milo, meta.data = data.frame(colData(Bcl6bNT_milo)), 
                           samples = "Samples")
Zfp711NT_milo <- countCells(Zfp711NT_milo, meta.data = data.frame(colData(Zfp711NT_milo)),
                            samples = "Samples")

# ==============================================================================
# Create Experimental Design Data Frames
# ==============================================================================

# Extract sample-level information for the statistical model
# distinct(): Keep only unique sample-condition combinations

# --- Atf3 Design ---
Atf3NT_design <- data.frame(colData(Atf3NT_milo))[, c("Samples", "Conditions")]
Atf3NT_design <- distinct(Atf3NT_design)
rownames(Atf3NT_design) <- Atf3NT_design$Samples
# Set factor levels with Control as reference (important for interpretation)
Atf3NT_design$Conditions <- factor(Atf3NT_design$Conditions, levels = c("Control", "Atf3KO"))

# --- Bcl6b Design ---
Bcl6bNT_design <- data.frame(colData(Bcl6bNT_milo))[, c("Samples", "Conditions")]
Bcl6bNT_design <- distinct(Bcl6bNT_design)
rownames(Bcl6bNT_design) <- Bcl6bNT_design$Samples

# --- Zfp711 Design ---
Zfp711NT_design <- data.frame(colData(Zfp711NT_milo))[, c("Samples", "Conditions")]
Zfp711NT_design <- distinct(Zfp711NT_design)
rownames(Zfp711NT_design) <- Zfp711NT_design$Samples

# ==============================================================================
# Calculate Neighborhood Distances
# ==============================================================================

# calcNhoodDistance(): Compute distances between neighborhoods
# Required for spatial FDR correction
# d = 50: Number of PCA dimensions (same as buildGraph)
Atf3NT_milo <- calcNhoodDistance(Atf3NT_milo, d = 50)
Bcl6bNT_milo <- calcNhoodDistance(Bcl6bNT_milo, d = 50)
Zfp711NT_milo <- calcNhoodDistance(Zfp711NT_milo, d = 50)

# ==============================================================================
# Perform Differential Abundance Testing
# ==============================================================================

# testNhoods(): Run negative binomial GLM for each neighborhood
# design = ~ Conditions: Formula for the model (KO vs Control)
# design.df: Data frame with sample-level covariates
# Returns: logFC, PValue, FDR, SpatialFDR for each neighborhood
Atf3NT_da_results <- testNhoods(Atf3NT_milo, design = ~ Conditions, design.df = Atf3NT_design)
Bcl6bNT_da_results <- testNhoods(Bcl6bNT_milo, design = ~ Conditions, design.df = Bcl6bNT_design)
Zfp711NT_da_results <- testNhoods(Zfp711NT_milo, design = ~ Conditions, design.df = Zfp711NT_design)

# ==============================================================================
# Display Top Results
# ==============================================================================

# arrange(-SpatialFDR): Sort by Spatial FDR (descending = least significant first)
# head(): Show top 6 results
Atf3NT_da_results %>% arrange(-SpatialFDR) %>% head()
Bcl6bNT_da_results %>% arrange(-SpatialFDR) %>% head()
Zfp711NT_da_results %>% arrange(-SpatialFDR) %>% head()

# ==============================================================================
# P-Value Distribution Histograms
# ==============================================================================

# Uniform p-value distribution = no signal
# Peak near 0 = many significant neighborhoods
ggplot(Atf3NT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
  ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                   size = 20, face = "bold.italic"))

ggplot(Bcl6bNT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
  ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                    size = 20, face = "bold.italic"))

ggplot(Zfp711NT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
  ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                                     size = 20, face = "bold.italic"))

# ==============================================================================
# P-Value vs Spatial FDR Scatter Plots
# ==============================================================================

# Shows relationship between raw p-value and spatially-corrected FDR
plot_grid(
  ggplot(Atf3NT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                     size = 20, face = "bold.italic")),
  ggplot(Bcl6bNT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                      size = 20, face = "bold.italic")),
  ggplot(Zfp711NT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                                       size = 20, face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Volcano Plots (logFC vs -log10 SpatialFDR)
# ==============================================================================

# Volcano plot: Effect size (logFC) vs significance (-log10 SpatialFDR)
# geom_hline at y=1: SpatialFDR = 0.1 threshold
plot_grid(
  ggplot(Atf3NT_da_results, aes(logFC, -log10(SpatialFDR))) + geom_point() + 
    geom_hline(yintercept = 1) + theme_classic() +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                     hjust = 0.5, size = 20, 
                                                                     face = "bold.italic")),
  ggplot(Bcl6bNT_da_results, aes(logFC, -log10(SpatialFDR))) + geom_point() + 
    geom_hline(yintercept = 1) + theme_classic() +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = 
                                                                        "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ggplot(Zfp711NT_da_results, aes(logFC, -log10(SpatialFDR))) + geom_point() + 
    geom_hline(yintercept = 1) + theme_classic() +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = 
                                                                         "black", hjust = 0.5, size = 20,
                                                                       face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# MA Plots
# ==============================================================================

# plotNhoodMA(): MA plot (log mean abundance vs logFC)
# Shows whether effect size depends on neighborhood size
plot_grid(
  plotNhoodMA(Atf3NT_da_results) + theme_classic() +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                                     size = 20, face = "bold.italic")),
  plotNhoodMA(Bcl6bNT_da_results) + theme_classic() +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                      size = 20, face = "bold.italic")),
  plotNhoodMA(Zfp711NT_da_results) + theme_classic() +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                                       size = 20, face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Build Neighborhood Graphs for UMAP Visualization
# ==============================================================================

# buildNhoodGraph(): Create graph connecting neighborhoods for visualization
Atf3NT_milo <- buildNhoodGraph(Atf3NT_milo)
Bcl6bNT_milo <- buildNhoodGraph(Bcl6bNT_milo)
Zfp711NT_milo <- buildNhoodGraph(Zfp711NT_milo)

# ==============================================================================
# UMAP with Neighborhood DA Overlay
# ==============================================================================

# plot_grid(): Combine multiple plots from cowplot
# plotReducedDim(): UMAP colored by cluster identity
# plotNhoodGraphDA(): UMAP with neighborhoods colored by logFC

plot_grid(
  plotReducedDim(Atf3NT_milo, dimred = "UMAPEARLY", colour_by = "cl_3", text_size = 4,
                 text_colour = 'black') +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black",
                                                                     hjust = 0.5, size = 20, 
                                                                     face = "bold.italic")) +
    theme(legend.text = element_text(angle = 0, size = 15), legend.title = 
            element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) +
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout = "UMAPEARLY", alpha = 0.5) +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                     hjust = 0.5, size = 20,
                                                                     face = "bold.italic")),
  ncol = 2
)

plot_grid(
  plotReducedDim(Bcl6bNT_milo, dimred = "UMAPEARLY", colour_by = "cl_3", text_size = 4,
                 text_colour = 'black') +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                      hjust = 0.5, size = 20, 
                                                                      face = "bold.italic")) +
    theme(legend.text = element_text(angle = 0, size = 15), legend.title = 
            element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) +
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Bcl6bNT_milo, Bcl6bNT_da_results, layout = "UMAPEARLY", alpha = 0.5) +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                      hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 2
)

plot_grid(
  plotReducedDim(Zfp711NT_milo, dimred = "UMAPEARLY", colour_by = "cl_3", 
                 text_size = 4, text_colour = 'black') +
    ggtitle("Zfp711-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")) +
    theme(legend.text = element_text(angle = 0, 
                                     size = 15), legend.title =
            element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) +
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Zfp711NT_milo, 
                   Zfp711NT_da_results, layout = "UMAPEARLY", alpha = 0.5) +
    ggtitle("Zfp711-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  ncol = 2
)

# ==============================================================================
# Annotate Neighborhoods with Cluster Identity
# ==============================================================================

# annotateNhoods(): Assign cluster labels to neighborhoods
# coldata_col = "cl_3": Which metadata column to use for annotation
# Adds: cl_3 (most common cluster), cl_3_fraction (proportion of that cluster)
Atf3NT_da_results <- annotateNhoods(Atf3NT_milo, Atf3NT_da_results, coldata_col = "cl_3")
lv_early <- levels(early_ControlAtf3$cl_3)
Atf3NT_da_results$cl_3_lv <- factor(Atf3NT_da_results$cl_3, levels = lv_early)

Bcl6bNT_da_results <- annotateNhoods(Bcl6bNT_milo, Bcl6bNT_da_results, coldata_col = "cl_3")
Bcl6bNT_da_results$cl_3_lv <- factor(Bcl6bNT_da_results$cl_3, levels = lv_early)

Zfp711NT_da_results <- annotateNhoods(Zfp711NT_milo, Zfp711NT_da_results, coldata_col = "cl_3")
Zfp711NT_da_results$cl_3_lv <- factor(Zfp711NT_da_results$cl_3, levels = lv_early)

# ==============================================================================
# Handle Mixed Neighborhoods
# ==============================================================================

# Neighborhoods with cl_3_fraction < 0.5 are "Mixed" (no dominant cluster)
# ifelse(): Create new annotation that labels mixed neighborhoods
Atf3NT_da_results$cl_3_mx <-
  ifelse(Atf3NT_da_results$cl_3_fraction < 0.5, "Mixed", Atf3NT_da_results$cl_3)
Bcl6bNT_da_results$cl_3_mx <- 
  ifelse(Bcl6bNT_da_results$cl_3_fraction < 0.5, "Mixed", Bcl6bNT_da_results$cl_3)
Zfp711NT_da_results$cl_3_mx <- 
  ifelse(Zfp711NT_da_results$cl_3_fraction < 0.5, "Mixed", Zfp711NT_da_results$cl_3)

# Add "Mixed" to factor levels
lv_early_p <- c(lv_early, "Mixed")
Atf3NT_da_results$cl_3_mx_lv <- factor(Atf3NT_da_results$cl_3_mx, levels = lv_early_p)
Bcl6bNT_da_results$cl_3_mx_lv <- factor(Bcl6bNT_da_results$cl_3_mx, levels = lv_early_p)
Zfp711NT_da_results$cl_3_mx_lv <- factor(Zfp711NT_da_results$cl_3_mx, levels = lv_early_p)

# droplevels(): Remove unused factor levels
Atf3NT_da_results$cl_3_mx_lv2 <- droplevels(Atf3NT_da_results$cl_3_mx_lv)
Bcl6bNT_da_results$cl_3_mx_lv2 <- droplevels(Bcl6bNT_da_results$cl_3_mx_lv)
Zfp711NT_da_results$cl_3_mx_lv2 <- droplevels(Zfp711NT_da_results$cl_3_mx_lv)

# ==============================================================================
# Final Visualization: UMAP + Neighborhood Graph + Beeswarm
# ==============================================================================

# plotDAbeeswarm(): Beeswarm plot showing logFC distribution per cluster
# Combines: Cell type UMAP, DA neighborhood graph, and logFC by cluster
plot_grid(
  plotReducedDim(Atf3NT_milo, dimred = "UMAPEARLY", colour_by = "cl_3",
                 text_size = 4, text_colour = 'black') +
    theme(legend.text = element_text(angle = 0, size = 15),
          legend.title = element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) +
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout = "UMAPEARLY", alpha = 0.5),
  plotDAbeeswarm(Atf3NT_da_results, group.by = "cl_3_mx_lv2", alpha = 1) + xlab("") +
    theme(axis.title.x = element_text(angle = 0, size = 12), 
          axis.text.y = element_text(angle = 0, size = 15)),
  ncol = 3, rel_widths = c(1.1, 1.1, 1)
)

```


# ==============================================================================
# SECTION 4: HEMATOPOIETIC - DATA LOADING AND MILO SETUP
# ==============================================================================

## Heamtopoietic

```{r eval=FALSE}
# ==============================================================================
# Load Seurat Subsets for Hematopoietic Group
# ==============================================================================

# Load pre-subsetted Seurat objects for hematopoietic cells
hemato_ControlAtf3 <- readRDS("~/cmo_chapter/introduction/rds/Hemato/hemato_CA.RDS")
hemato_ControlBcl6b <- readRDS("~/cmo_chapter/introduction/rds/Hemato/hemato_BA.RDS")
hemato_ControlZfp711 <- readRDS("~/cmo_chapter/introduction/rds/Hemato/hemato_ZA.RDS")

# Display sample composition tables
table(hemato_ControlAtf3$Samples_lv, hemato_ControlAtf3$CellRanger_Assignment)
table(hemato_ControlBcl6b$Samples_lv, hemato_ControlBcl6b$CellRanger_Assignment)
table(hemato_ControlZfp711$Samples_lv, hemato_ControlZfp711$CellRanger_Assignment)

# ==============================================================================
# Convert to SingleCellExperiment and Create Milo Objects
# ==============================================================================

# --- Atf3 vs Control ---
sceAtf3NT <- as.SingleCellExperiment(hemato_ControlAtf3)
# pcahema: PCA reduction specific to hematopoietic subset
reducedDim(sceAtf3NT, "PCA") <- hemato_ControlAtf3@reductions$pcahema@cell.embeddings
Atf3NT_milo <- Milo(sceAtf3NT)
Atf3NT_milo

# --- Bcl6b vs Control ---
sceBcl6bNT <- as.SingleCellExperiment(hemato_ControlBcl6b)
reducedDim(sceBcl6bNT, "PCA") <- hemato_ControlBcl6b@reductions$pcahema@cell.embeddings
Bcl6bNT_milo <- Milo(sceBcl6bNT)
Bcl6bNT_milo

# --- Zfp711 vs Control ---
sceZfp711NT <- as.SingleCellExperiment(hemato_ControlZfp711)
reducedDim(sceZfp711NT, "PCA") <- hemato_ControlZfp711@reductions$pcahema@cell.embeddings
Zfp711NT_milo <- Milo(sceZfp711NT)
Zfp711NT_milo

# ==============================================================================
# Build k-NN Graphs
# ==============================================================================

Atf3NT_milo <- buildGraph(Atf3NT_milo, k = 35, d = 50)
Bcl6bNT_milo <- buildGraph(Bcl6bNT_milo, k = 35, d = 50)
Zfp711NT_milo <- buildGraph(Zfp711NT_milo, k = 35, d = 50)

# ==============================================================================
# Create Neighborhoods
# ==============================================================================

Atf3NT_milo <- makeNhoods(Atf3NT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)
Bcl6bNT_milo <- makeNhoods(Bcl6bNT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)
Zfp711NT_milo <- makeNhoods(Zfp711NT_milo, prop = 0.3, k = 35, d = 50, refined = TRUE)

# ==============================================================================
# Visualize Neighborhood Size Distribution
# ==============================================================================

plot_grid(
  plotNhoodSizeHist(Atf3NT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black",
                                                                     hjust = 0.5, size = 20,
                                                                    face = "bold.italic")),
  plotNhoodSizeHist(Bcl6bNT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                      hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  plotNhoodSizeHist(Zfp711NT_milo) + scale_x_continuous(breaks = seq(0, 300, 25)) +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                       hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Prepare Sample Identifiers and Count Cells
# ==============================================================================

Atf3NT_milo@colData@listData[["Samples"]] <- as.character(Atf3NT_milo@colData@listData$Samples_lv)
Bcl6bNT_milo@colData@listData[["Samples"]] <- as.character(Bcl6bNT_milo@colData@listData$Samples_lv)
Zfp711NT_milo@colData@listData[["Samples"]] <- as.character(Zfp711NT_milo@colData@listData$Samples_lv)

Atf3NT_milo <- countCells(Atf3NT_milo, meta.data = data.frame(colData(Atf3NT_milo)),
                          samples = "Samples")
Bcl6bNT_milo <- countCells(Bcl6bNT_milo, meta.data = data.frame(colData(Bcl6bNT_milo)), 
                           samples = "Samples")
Zfp711NT_milo <- countCells(Zfp711NT_milo, meta.data = data.frame(colData(Zfp711NT_milo)), 
                            samples = "Samples")

# ==============================================================================
# Create Experimental Design Data Frames
# ==============================================================================

Atf3NT_design <- data.frame(colData(Atf3NT_milo))[, c("Samples", "Conditions")]
Atf3NT_design <- distinct(Atf3NT_design)
rownames(Atf3NT_design) <- Atf3NT_design$Samples
Atf3NT_design

Atf3NT_design$Conditions <- factor(Atf3NT_design$Conditions, levels = c("Control", "Atf3KO"))

Bcl6bNT_design <- data.frame(colData(Bcl6bNT_milo))[, c("Samples", "Conditions")]
Bcl6bNT_design <- distinct(Bcl6bNT_design)
rownames(Bcl6bNT_design) <- Bcl6bNT_design$Samples

Zfp711NT_design <- data.frame(colData(Zfp711NT_milo))[, c("Samples", "Conditions")]
Zfp711NT_design <- distinct(Zfp711NT_design)
rownames(Zfp711NT_design) <- Zfp711NT_design$Samples

# ==============================================================================
# Calculate Neighborhood Distances and Run DA Testing
# ==============================================================================

Atf3NT_milo <- calcNhoodDistance(Atf3NT_milo, d = 50)
Bcl6bNT_milo <- calcNhoodDistance(Bcl6bNT_milo, d = 50)
Zfp711NT_milo <- calcNhoodDistance(Zfp711NT_milo, d = 50)

Atf3NT_da_results <- testNhoods(Atf3NT_milo, design = ~ Conditions, design.df = Atf3NT_design)
Bcl6bNT_da_results <- testNhoods(Bcl6bNT_milo, design = ~ Conditions, design.df = Bcl6bNT_design)
Zfp711NT_da_results <- testNhoods(Zfp711NT_milo, design = ~ Conditions, design.df = Zfp711NT_design)

# Display top results
Atf3NT_da_results %>% arrange(-SpatialFDR) %>% head()
Bcl6bNT_da_results %>% arrange(-SpatialFDR) %>% head()
Zfp711NT_da_results %>% arrange(-SpatialFDR) %>% head()

# ==============================================================================
# Visualization: P-value Histograms
# ==============================================================================

plot_grid(
  ggplot(Atf3NT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
    ggtitle("Atf3-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                    face = "bold.italic")),
  ggplot(Bcl6bNT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
    ggtitle("Bcl6b-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ggplot(Zfp711NT_da_results, aes(PValue)) + geom_histogram(bins = 100) + theme_classic() +
    ggtitle("Zfp711-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Visualization: P-value vs SpatialFDR
# ==============================================================================

plot_grid(
  ggplot(Atf3NT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Atf3-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                    face = "bold.italic")),
  ggplot(Bcl6bNT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Bcl6b-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ggplot(Zfp711NT_da_results, aes(PValue, SpatialFDR)) + geom_point() + theme_classic() +
    ggtitle("Zfp711-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Visualization: Volcano Plots
# ==============================================================================

plot_grid(
  ggplot(Atf3NT_da_results, aes(logFC, -log10(SpatialFDR))) + geom_point() +
    geom_hline(yintercept = 1) +
    theme_classic() + ggtitle("Atf3-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                size = 20, face = "bold.italic")),
  ggplot(Bcl6bNT_da_results, aes(logFC, -log10(SpatialFDR))) + 
    geom_point() + geom_hline(yintercept = 1) +
    theme_classic() + ggtitle("Bcl6b-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  ggplot(Zfp711NT_da_results, aes(logFC, -log10(SpatialFDR))) + geom_point() +
    geom_hline(yintercept = 1) +
    theme_classic() + ggtitle("Zfp711-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Visualization: MA Plots
# ==============================================================================

plot_grid(
  plotNhoodMA(Atf3NT_da_results) + theme_classic() + ggtitle("Atf3-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  plotNhoodMA(Bcl6bNT_da_results) + theme_classic() + ggtitle("Bcl6b-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  plotNhoodMA(Zfp711NT_da_results) + theme_classic() + ggtitle("Zfp711-KO and Control") +
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20, face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Build Neighborhood Graphs and Visualize
# ==============================================================================

Atf3NT_milo <- buildNhoodGraph(Atf3NT_milo)
Bcl6bNT_milo <- buildNhoodGraph(Bcl6bNT_milo)
Zfp711NT_milo <- buildNhoodGraph(Zfp711NT_milo)

# UMAPHEMA: UMAP reduction specific to hematopoietic cells
plot_grid(
  plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout = "UMAPHEMA", alpha = 0.1) +
    ggtitle("Atf3-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                     size = 20,
                                                                    face = "bold.italic")),
  plotNhoodGraphDA(Bcl6bNT_milo, Bcl6bNT_da_results, layout = "UMAPHEMA", alpha = 0.1) +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5,
                                                                      size = 20,
                                                                      face = "bold.italic")),
  plotNhoodGraphDA(Zfp711NT_milo, Zfp711NT_da_results, layout = "UMAPHEMA", alpha = 0.1) +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = element_text(color = "black", hjust = 0.5, 
                                                                       size = 20,
                                                                      face = "bold.italic")),
  ncol = 3
)

# ==============================================================================
# Combined UMAP and Neighborhood Visualization
# ==============================================================================

plot_grid(
  plotReducedDim(Bcl6bNT_milo, dimred = "UMAPHEMA", colour_by = "cl_3", text_size = 4,
                 text_colour = 'black') +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                      hjust = 0.5, size = 20,
                                                                    face = "bold.italic")) +
    theme(legend.text = element_text(angle = 0, size = 15), 
          legend.title = element_text(size = 0,  hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) +
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Bcl6bNT_milo, Bcl6bNT_da_results, layout = "UMAPHEMA", alpha = 0.5) +
    ggtitle("Bcl6b-KO and Control") + theme(plot.title = element_text(color = "black", 
                                                                      hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 2
)

plot_grid(
  plotReducedDim(Zfp711NT_milo, dimred = "UMAPHEMA", colour_by = "cl_3",
                 text_size = 4, text_colour = 'black') +
    ggtitle("Zfp711-KO and Control") + 
    theme(plot.title = element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")) +
    theme(legend.text = element_text(angle = 0, size = 15), 
          legend.title = element_text(size = 0, hjust = 0.5, vjust = 1,  face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) + 
    theme(legend.direction = "vertical",egend.position = "left"),
  plotNhoodGraphDA(Zfp711NT_milo, Zfp711NT_da_results, layout = "UMAPHEMA", alpha = 0.5) +
    ggtitle("Zfp711-KO and Control") + theme(plot.title = 
                                               element_text(color = "black", hjust = 0.5, size = 20,
                                                                      face = "bold.italic")),
  ncol = 2
)

# ==============================================================================
# Annotate Neighborhoods and Handle Mixed Clusters
# ==============================================================================

Atf3NT_da_results <- annotateNhoods(Atf3NT_milo, Atf3NT_da_results, coldata_col = "cl_3")
lv_hemato <- levels(hemato_ControlAtf3$cl_3)
Atf3NT_da_results$cl_3_lv <- factor(Atf3NT_da_results$cl_3, levels = lv_hemato)

Bcl6bNT_da_results <- annotateNhoods(Bcl6bNT_milo, Bcl6bNT_da_results, coldata_col = "cl_3")
Bcl6bNT_da_results$cl_3_lv <- factor(Bcl6bNT_da_results$cl_3, levels = lv_hemato)

Zfp711NT_da_results <- annotateNhoods(Zfp711NT_milo, Zfp711NT_da_results, coldata_col = "cl_3")
Zfp711NT_da_results$cl_3_lv <- factor(Zfp711NT_da_results$cl_3, levels = lv_hemato)

# Handle mixed neighborhoods
Atf3NT_da_results$cl_3_mx <- ifelse(Atf3NT_da_results$cl_3_fraction < 0.5, "Mixed",
                                    Atf3NT_da_results$cl_3)
Bcl6bNT_da_results$cl_3_mx <- ifelse(Bcl6bNT_da_results$cl_3_fraction < 0.5, "Mixed",
                                     Bcl6bNT_da_results$cl_3)
Zfp711NT_da_results$cl_3_mx <- ifelse(Zfp711NT_da_results$cl_3_fraction < 0.5, "Mixed", 
                                      Zfp711NT_da_results$cl_3)

lv_hemato_p <- c(lv_hemato, "Mixed")
Atf3NT_da_results$cl_3_mx_lv <- factor(Atf3NT_da_results$cl_3_mx, levels = lv_hemato_p)
Bcl6bNT_da_results$cl_3_mx_lv <- factor(Bcl6bNT_da_results$cl_3_mx, levels = lv_hemato_p)
Zfp711NT_da_results$cl_3_mx_lv <- factor(Zfp711NT_da_results$cl_3_mx, levels = lv_hemato_p)

Atf3NT_da_results$cl_3_mx_lv2 <- droplevels(Atf3NT_da_results$cl_3_mx_lv)
Bcl6bNT_da_results$cl_3_mx_lv2 <- droplevels(Bcl6bNT_da_results$cl_3_mx_lv)
Zfp711NT_da_results$cl_3_mx_lv2 <- droplevels(Zfp711NT_da_results$cl_3_mx_lv)

Atf3NT_da_results$cl_3_lv <- droplevels(Atf3NT_da_results$cl_3_lv)
Bcl6bNT_da_results$cl_3_lv <- droplevels(Bcl6bNT_da_results$cl_3_lv)
Zfp711NT_da_results$cl_3_lv <- droplevels(Zfp711NT_da_results$cl_3_lv)

# ==============================================================================
# Final Visualization with Beeswarm Plots
# ==============================================================================

plot_grid(
  plotReducedDim(Atf3NT_milo, dimred = "UMAPHEMA", colour_by = "cl_3", text_by = "cl_3",
                 text_size = 6, text_colour = 'black') +
    theme(legend.text = element_text(angle = 0, size = 15), legend.title = 
            element_text(size = 0, hjust = 0.5, vjust = 1,face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) + 
    theme(legend.direction = "vertical", legend.position = "left"),
  plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout = "UMAPHEMA", alpha = 0.1),
  plotDAbeeswarm(Atf3NT_da_results, group.by = "cl_3_mx_lv2", alpha = 1) + xlab("") +
    theme(axis.title.x = element_text(angle = 0, size = 12), axis.text.y = 
            element_text(angle = 0, size = 15)),
  ncol = 3, rel_widths = c(1.1, 1.1, 1)
)

plot_grid(
  plotReducedDim(Bcl6bNT_milo, dimred = "UMAPHEMA", colour_by = "cl_3", text_by = "cl_3",
                 text_size = 6, text_colour = 'black') +
    theme(legend.text = element_text(angle = 0, size = 15), legend.title = 
            element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) + 
    theme(legend.direction = "vertical",legend.position = "left"),
  plotNhoodGraphDA(Bcl6bNT_milo, Atf3NT_da_results, layout = "UMAPHEMA", alpha = 0.1),
  plotDAbeeswarm(Bcl6bNT_da_results, group.by = "cl_3_mx_lv2", alpha = 1) + xlab("") +
    theme(axis.title.x = element_text(angle = 0, size = 12), axis.text.y = 
            element_text(angle = 0, size = 15)),
  ncol = 3, rel_widths = c(1.1, 1.1, 1)
)

plot_grid(
  plotReducedDim(Zfp711NT_milo, dimred = "UMAPHEMA", colour_by = "cl_3", text_by = "cl_3",
                 text_size = 6, text_colour = 'black') +
    theme(legend.text = element_text(angle = 0, size = 15), legend.title = 
            element_text(size = 0, hjust = 0.5, vjust = 1, face = 'bold')) +
    guides(color = guide_legend(override.aes = list(size = 7), ncol = 1)) + 
    theme(legend.direction = "vertical",  legend.position = "left"),
  plotNhoodGraphDA(Zfp711NT_milo, Atf3NT_da_results, layout = "UMAPHEMA", alpha = 0.1),
  plotDAbeeswarm(Zfp711NT_da_results, group.by = "cl_3_mx_lv2", alpha = 1) + xlab("") +
    theme(axis.title.x = element_text(angle = 0, size = 12), axis.text.y = 
            element_text(angle = 0, size = 15)),
  ncol = 3, rel_widths = c(1.1, 1.1, 1)
)

```


# ==============================================================================
# SECTION 5: LATE MESODERM - DATA LOADING AND MILO SETUP
# ==============================================================================

## LateMesoderm

```{r eval=FALSE}
# ==============================================================================
# Load Seurat Subsets for Late Mesoderm Group
# ==============================================================================

mesoderm_ControlAtf3 <- readRDS("~/cmo_chapter/introduction/rds/Mesoderm/mesoderm_CA.RDS")
mesoderm_ControlBcl6b <- readRDS("~/cmo_chapter/introduction/rds/Mesoderm/mesoderm_BA.RDS")
mesoderm_ControlZfp711 <- readRDS("~/cmo_chapter/introduction/rds/Mesoderm/mesoderm_ZA.RDS")

# ==============================================================================
# Convert to SingleCellExperiment and Create Milo Objects
# ==============================================================================

# pcamesoderm: PCA reduction specific to mesoderm subset
sceAtf3NT <- as.SingleCellExperiment(mesoderm_ControlAtf3)
reducedDim(sceAtf3NT, "PCA") <-mesoderm_ControlAtf3@reductions$pcamesoderm@cell.embeddings
Atf3NT_milo <- Milo(sceAtf3NT)

sceBcl6bNT <- as.SingleCellExperiment(mesoderm_ControlBcl6b)
reducedDim(sceBcl6bNT, "PCA") <- mesoderm_ControlBcl6b@reductions$pcamesoderm@cell.embeddings
Bcl6bNT_milo <- Milo(sceBcl6bNT)

sceZfp711NT <- as.SingleCellExperiment(mesoderm_ControlZfp711)
reducedDim(sceZfp711NT, "PCA") <- mesoderm_ControlZfp711@reductions$pcamesoderm@cell.embeddings
Zfp711NT_milo <- Milo(sceZfp711NT)

# ==============================================================================
# Build k-NN Graphs and Create Neighborhoods
# ==============================================================================

Atf3NT_milo <- buildGraph(Atf3NT_milo, k = 35, d = 50)
Bcl6bNT_milo <- buildGraph(Bcl6bNT_milo, k = 35, d = 50)
Zfp711NT_milo <- buildGraph(Zfp711NT_milo, k = 35, d = 50)

Atf3NT_milo   <- makeNhoods(Atf3NT_milo, prop = 0.3, k = 35, d=50, refined = TRUE)
Bcl6bNT_milo  <- makeNhoods(Bcl6bNT_milo, prop = 0.3, k = 35, d=50, refined = TRUE)
Zfp711NT_milo <- makeNhoods(Zfp711NT_milo, prop = 0.3, k = 35, d=50, refined = TRUE)

# ==============================================================================
# Visualize Neighborhood Size Distribution
# ==============================================================================

plot_grid(
	plotNhoodSizeHist(Atf3NT_milo) + scale_x_continuous(breaks = seq(0, 300, 25))+
	  ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	plotNhoodSizeHist(Bcl6bNT_milo)+ scale_x_continuous(breaks = seq(0, 300, 25))+
	  ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	plotNhoodSizeHist(Zfp711NT_milo)+ scale_x_continuous(breaks = seq(0, 300, 25))+
	  ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=3
)

# ==============================================================================
# Prepare Sample Identifiers and Count Cells
# ==============================================================================

Atf3NT_milo@colData@listData[["Samples"]] <- as.character(Atf3NT_milo@colData@listData$Samples_lv)
Bcl6bNT_milo@colData@listData[["Samples"]] <- as.character(Bcl6bNT_milo@colData@listData$Samples_lv)
Zfp711NT_milo@colData@listData[["Samples"]] <- as.character(Zfp711NT_milo@colData@listData$Samples_lv)

Atf3NT_milo <- countCells(Atf3NT_milo, meta.data = data.frame(colData(Atf3NT_milo)),
                          samples="Samples")
Bcl6bNT_milo <- countCells(Bcl6bNT_milo, meta.data = data.frame(colData(Bcl6bNT_milo)), 
                           samples="Samples")
Zfp711NT_milo <- countCells(Zfp711NT_milo, meta.data = data.frame(colData(Zfp711NT_milo)),
                            samples="Samples")

# ==============================================================================
# Create Experimental Design Data Frames
# ==============================================================================

Atf3NT_design <- data.frame(colData(Atf3NT_milo))[,c("Samples", "Conditions")]
Atf3NT_design <- distinct(Atf3NT_design)
rownames(Atf3NT_design) <- Atf3NT_design$Samples

Atf3NT_design$Conditions <- factor(Atf3NT_design$Conditions, levels = c("Control", "Atf3KO"))

Bcl6bNT_design <- data.frame(colData(Bcl6bNT_milo))[,c("Samples", "Conditions")]
Bcl6bNT_design <- distinct(Bcl6bNT_design)
rownames(Bcl6bNT_design) <- Bcl6bNT_design$Samples


Zfp711NT_design <- data.frame(colData(Zfp711NT_milo))[,c("Samples", "Conditions")]
Zfp711NT_design <- distinct(Zfp711NT_design)
rownames(Zfp711NT_design) <- Zfp711NT_design$Samples

# ==============================================================================
# Calculate Neighborhood Distances and Run DA Testing
# ==============================================================================

Atf3NT_milo <- calcNhoodDistance(Atf3NT_milo, d=50)
Bcl6bNT_milo <- calcNhoodDistance(Bcl6bNT_milo, d=50)
Zfp711NT_milo <- calcNhoodDistance(Zfp711NT_milo, d=50)

Atf3NT_da_results <- testNhoods(Atf3NT_milo, design = ~ Conditions, design.df = Atf3NT_design)
Bcl6bNT_da_results <- testNhoods(Bcl6bNT_milo, design = ~ Conditions, design.df = Bcl6bNT_design)
Zfp711NT_da_results <- testNhoods(Zfp711NT_milo, design = ~ Conditions, design.df = Zfp711NT_design)

# Display top results
Atf3NT_da_results %>%
	arrange(- SpatialFDR) %>%
	head()

Bcl6bNT_da_results %>%
	arrange(- SpatialFDR) %>%
	head()

Zfp711NT_da_results %>%
	arrange(- SpatialFDR) %>%
	head()

# ==============================================================================
# Visualization: P-value Histograms
# ==============================================================================

plot_grid(
	ggplot(Atf3NT_da_results, aes(PValue)) + geom_histogram(bins=100) + theme_classic()+
	  ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Bcl6bNT_da_results, aes(PValue)) + geom_histogram(bins=100) + theme_classic()+
	  ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Zfp711NT_da_results, aes(PValue)) + geom_histogram(bins=100) + theme_classic()+
	  ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=3
)

# ==============================================================================
# Visualization: P-value vs SpatialFDR
# ==============================================================================

plot_grid(
	ggplot(Atf3NT_da_results, aes(PValue, SpatialFDR)) + 
		geom_point() + theme_classic()+ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Bcl6bNT_da_results, aes(PValue, SpatialFDR)) + 
		geom_point() + theme_classic()+ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Zfp711NT_da_results, aes(PValue, SpatialFDR)) + 
		geom_point() + theme_classic()+ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=3
)

# ==============================================================================
# Visualization: Volcano Plots
# ==============================================================================

plot_grid(
	ggplot(Atf3NT_da_results, aes(logFC, -log10(SpatialFDR))) + 
		geom_point() +
		geom_hline(yintercept = 1)+ theme_classic()+ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Bcl6bNT_da_results, aes(logFC, -log10(SpatialFDR))) + 
		geom_point() +
		geom_hline(yintercept = 1)+ theme_classic()+ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	ggplot(Zfp711NT_da_results, aes(logFC, -log10(SpatialFDR))) + 
		geom_point() +
		geom_hline(yintercept = 1)+ theme_classic()+ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=3
)

# ==============================================================================
# Visualization: MA Plots
# ==============================================================================

plot_grid(
	plotNhoodMA(Atf3NT_da_results) + theme_classic()+ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	plotNhoodMA(Bcl6bNT_da_results) + theme_classic()+ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	
	plotNhoodMA(Zfp711NT_da_results) + theme_classic()+ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=3
)

# ==============================================================================
# Build Neighborhood Graphs and Visualize
# ==============================================================================

Atf3NT_milo <- buildNhoodGraph(Atf3NT_milo)
Bcl6bNT_milo <- buildNhoodGraph(Bcl6bNT_milo)
Zfp711NT_milo <- buildNhoodGraph(Zfp711NT_milo)

# UMAPMESODERM: UMAP reduction specific to mesoderm cells
plot_grid(
	plotReducedDim(Atf3NT_milo, dimred = "UMAPMESODERM",  colour_by="cl_3", text_size = 4,
	               text_colour = 'black')+ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic"))+
	  theme(legend.text = element_text(angle = 0,size = 15),
	        legend.title = element_text(size = 0,hjust=0.5,vjust = 1, face = 'bold'))+ 
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout="UMAPMESODERM",alpha=0.5) +
	  ggtitle("Atf3-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=2
)

plot_grid(
	plotReducedDim(Bcl6bNT_milo, dimred = "UMAPMESODERM", 
	               colour_by="cl_3", text_size = 4,text_colour = 'black')+
	  ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic"))+
	  theme(legend.text = element_text(angle = 0,size = 15),
	        legend.title = element_text(size = 0, hjust=0.5,vjust = 1, face = 'bold'))+ 
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	plotNhoodGraphDA(Bcl6bNT_milo, Bcl6bNT_da_results, layout="UMAPMESODERM",alpha=0.5) +
	  ggtitle("Bcl6b-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=2
)

plot_grid(
	plotReducedDim(Zfp711NT_milo, dimred = "UMAPMESODERM",  colour_by="cl_3", text_size = 4,
	               text_colour = 'black')+ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic"))+
	  theme(legend.text = element_text(angle = 0,size = 15),legend.title = 
	          element_text(size = 0, hjust=0.5,vjust = 1, face = 'bold'))+ 
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	plotNhoodGraphDA(Zfp711NT_milo, Zfp711NT_da_results, layout="UMAPMESODERM",alpha=0.5) +
	  ggtitle("Zfp711-KO and Control") +
		theme(plot.title = element_text(color="black", hjust = 0.5,size=20, face="bold.italic")),
	ncol=2
)

# ==============================================================================
# Annotate Neighborhoods and Handle Mixed Clusters
# ==============================================================================

Atf3NT_da_results <- annotateNhoods(Atf3NT_milo, Atf3NT_da_results, coldata_col = "cl_3")

lv_mesoderm <- levels(mesoderm_ControlAtf3$cl_3)

Atf3NT_da_results$cl_3_lv <- factor(Atf3NT_da_results$cl_3, levels =lv_mesoderm) 

Bcl6bNT_da_results <- annotateNhoods(Bcl6bNT_milo, Bcl6bNT_da_results, coldata_col = "cl_3")

Bcl6bNT_da_results$cl_3_lv <- factor(Bcl6bNT_da_results$cl_3, levels =lv_mesoderm) 
Zfp711NT_da_results <- annotateNhoods(Zfp711NT_milo, Zfp711NT_da_results, coldata_col = "cl_3")
Zfp711NT_da_results$cl_3_lv <- factor(Zfp711NT_da_results$cl_3, levels =lv_mesoderm) 

Atf3NT_da_results$cl_3_mx <- ifelse(Atf3NT_da_results$cl_3_fraction < 0.5, "Mixed", 
                                    Atf3NT_da_results$cl_3)

Bcl6bNT_da_results$cl_3_mx <- ifelse(Bcl6bNT_da_results$cl_3_fraction < 0.5, "Mixed",
                                     Bcl6bNT_da_results$cl_3)

Zfp711NT_da_results$cl_3_mx <- ifelse(Zfp711NT_da_results$cl_3_fraction < 0.5, "Mixed", 
                                      Zfp711NT_da_results$cl_3)


lv_mesoderm
lv_mesoderm_p <- c(lv_mesoderm, "Mixed")
lv_mesoderm_p 

Atf3NT_da_results$cl_3_mx_lv <- factor(Atf3NT_da_results$cl_3_mx, levels =lv_mesoderm_p) 
Bcl6bNT_da_results$cl_3_mx_lv <- factor(Bcl6bNT_da_results$cl_3_mx, levels =lv_mesoderm_p) 
Zfp711NT_da_results$cl_3_mx_lv <- factor(Zfp711NT_da_results$cl_3_mx, levels =lv_mesoderm_p) 

Atf3NT_da_results$cl_3_mx_lv2 <- droplevels(Atf3NT_da_results$cl_3_mx_lv)
Bcl6bNT_da_results$cl_3_mx_lv2 <- droplevels(Bcl6bNT_da_results$cl_3_mx_lv)
Zfp711NT_da_results$cl_3_mx_lv2 <- droplevels(Zfp711NT_da_results$cl_3_mx_lv)

Zfp711NT_da_results$cl_3_lv <- droplevels(Zfp711NT_da_results$cl_3_lv)

# ==============================================================================
# Final Visualization with Beeswarm Plots
# ==============================================================================

plot_grid(
	plotReducedDim(Atf3NT_milo, dimred = "UMAPMESODERM",  colour_by="cl_3", text_by = "cl_3",
	               text_size = 6,text_colour = 'black')+
	  theme(legend.text = element_text(angle = 0,size = 15),legend.title = 
	          element_text(size = 0, hjust=0.5,vjust = 1, face = 'bold'))+ 
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	
	plotNhoodGraphDA(Atf3NT_milo, Atf3NT_da_results, layout="UMAPMESODERM",alpha=0.1) ,

		plotDAbeeswarm(Atf3NT_da_results, group.by = "cl_3_mx_lv2",alpha = 1) +xlab("")+
	  theme(axis.title.x = element_text(angle = 0,size = 12), axis.text.y =
	          element_text(angle = 0,size = 15)),
	ncol=3,rel_widths = c(1.1,1.1,1)
)

plot_grid(
	plotReducedDim(Bcl6bNT_milo, dimred = "UMAPMESODERM",  colour_by="cl_3", text_by = "cl_3",
	               text_size = 6,text_colour = 'black')+
	  theme(legend.text = element_text(angle = 0,size = 15),legend.title =
	          element_text(size = 0, hjust=0.5,vjust = 1, face = 'bold'))+
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	
	plotNhoodGraphDA(Bcl6bNT_milo, Atf3NT_da_results, layout="UMAPMESODERM",alpha=0.1) ,

	plotDAbeeswarm(Bcl6bNT_da_results, group.by = "cl_3_mx_lv2",alpha = 1) +xlab("")+
	  theme(axis.title.x = element_text(angle = 0,size = 12), axis.text.y =
	          element_text(angle = 0,size = 15)),
	ncol=3,rel_widths = c(1.1,1.1,1)
)

plot_grid(
	plotReducedDim(Zfp711NT_milo, dimred = "UMAPMESODERM",  colour_by="cl_3", text_by = "cl_3",
	               text_size = 6,text_colour = 'black')+
	  theme(legend.text = element_text(angle = 0,size = 15),legend.title =
	          element_text(size = 0, hjust=0.5,vjust = 1, face = 'bold'))+ 
	  guides(color = guide_legend(override.aes = list(size = 7),ncol = 1))+
	  theme(legend.direction ="vertical",legend.position = "left"),
	
	plotNhoodGraphDA(Zfp711NT_milo, Atf3NT_da_results, layout="UMAPMESODERM",alpha=0.1) ,

		plotDAbeeswarm(Zfp711NT_da_results, group.by = "cl_3_mx_lv2",alpha = 1) +xlab("")+
	  theme(axis.title.x = element_text(angle = 0,size = 12), axis.text.y = 
	          element_text(angle = 0,size = 15)),
	ncol=3,rel_widths = c(1.1,1.1,1)
)




```

# ==============================================================================
# SECTION 6: SESSION INFORMATION
# ==============================================================================

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================

sessionInfo()
```
