---
title: "Gene Set Enrichment Analysis"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

We will perform GSEA for one condition.

# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP AND PACKAGE LOADING
# ==============================================================================

```{r}
# ==============================================================================
# Load Required Packages
# ==============================================================================

suppressPackageStartupMessages({
  library(dplyr)    # Data manipulation
  library(fgsea)    # Fast Gene Set Enrichment Analysis
  library(ggplot2)  # Visualization
})
```

# ==============================================================================
# SECTION 2: LOAD DIFFERENTIAL GENE EXPRESSION DATA
# ==============================================================================

# Loading the data

## Loading the Differential Gene Expression (DGE) Data from DESeq2

Import DGE results from DESeq2 analysis. Ensure data integrity and format.

```{r}
# ==============================================================================
# Load DESeq2 Results
# ==============================================================================

# read.delim(): Load tab-separated DESeq2 output
# Contains: gene, baseMean, log2FoldChange, stat, pvalue, padj
dge_results <- read.delim(
  "/DESeq2_results_Atf3_cl_1/Late-Mesoderm_resOrdered_enhanced.txt")

# summary(): Check for missing values and data distribution
summary(dge_results)

# Visualize stat value distribution (used for ranking)
# stat: Wald statistic from DESeq2 (sign indicates direction, magnitude indicates confidence)
hist(dge_results$stat, breaks = 50, main = "Distribution of Gene Expression Values", xlab = "Stat Value")
```

```{r}
# ==============================================================================
# Visualize Stat vs Adjusted P-value
# ==============================================================================

# Scatter plot showing relationship between stat and significance
ggplot(dge_results, aes(x=stat, y=padj))+geom_point()
```

# ==============================================================================
# SECTION 3: LOAD MSIGDB GENE SETS
# ==============================================================================

## Loading the MSigDB (Molecular Signatures Database)

Load gene sets and pathways from MSigDB for enrichment analysis.

```{r}
# ==============================================================================
# Load MSigDB Gene Set Collections (Mouse Symbols)
# ==============================================================================

# gmtPathways(): Read GMT (Gene Matrix Transposed) format files
# GMT format: Each line is a pathway with gene members
# Version: v2023.2.Mm (Mouse symbols)

# -----------------------------------------------------------------------------
# MH: Hallmark Gene Sets
# -----------------------------------------------------------------------------
# Summarize well-defined biological states/processes with coherent expression
# 50 gene sets representing specific biological states
mh_hallmark <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/mh.all.v2023.2.Mm.symbols.gmt")

# -----------------------------------------------------------------------------
# M1: Positional Gene Sets
# -----------------------------------------------------------------------------
# Gene sets corresponding to chromosome cytogenetic bands
m1_positional <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m1.all.v2023.2.Mm.symbols.gmt")

# -----------------------------------------------------------------------------
# M2: Curated Gene Sets
# -----------------------------------------------------------------------------

# CGP: Chemical and Genetic Perturbations
# Expression signatures from genetic/chemical perturbations
# Often paired: xxx_UP (induced) and xxx_DN (repressed)
m2_curated_cgp <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m2.cgp.v2023.2.Mm.symbols.gmt")

# CP: Canonical Pathways (from various databases)

# BioCarta pathways
m2_cp_biocarta <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m2.cp.biocarta.v2023.2.Mm.symbols.gmt")

# Reactome pathways
m2_cp_reactome <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m2.cp.reactome.v2023.2.Mm.symbols.gmt")

# WikiPathways
m2_cp_wikipathways <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m2.cp.wikipathways.v2023.2.Mm.symbols.gmt")

# -----------------------------------------------------------------------------
# M3: Regulatory Target Gene Sets
# -----------------------------------------------------------------------------

# miRDB: microRNA targets
# High-confidence miRNA target predictions from miRDB v6.0
m3_miRDB <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m3.mirdb.v2023.2.Mm.symbols.gmt")

# GTRD: Transcription Factor Targets
# Genes with predicted TF binding sites in promoter region (-1000, +100 bp)
m3_gtrd <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m3.gtrd.v2023.2.Mm.symbols.gmt")

# -----------------------------------------------------------------------------
# M5: Ontology Gene Sets (Gene Ontology)
# -----------------------------------------------------------------------------

# GO Biological Process
m5_go_bp <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m5.go.bp.v2023.2.Mm.symbols.gmt")

# GO Cellular Component
m5_go_cc <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m5.go.cc.v2023.2.Mm.symbols.gmt")

# GO Molecular Function
m5_go_mf <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m5.go.mf.v2023.2.Mm.symbols.gmt")

# MPT: Mammalian Phenotype Ontology (tumor-related subset)
m5_go_mpt <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m5.mpt.v2023.2.Mm.symbols.gmt")

# -----------------------------------------------------------------------------
# M8: Cell Type Signature Gene Sets
# -----------------------------------------------------------------------------
# Curated cluster markers from single-cell studies
m8_celltype <- fgsea::gmtPathways(
  "~/cmo_chapter_v1/Analysis/4a_gsea/gmt/m8.all.v2023.2.Mm.symbols.gmt")
```

# ==============================================================================
# SECTION 4: CREATE RANKED GENE LIST
# ==============================================================================

## Creating Ranked Gene List with "stat"

Rank genes based on the "stat" column from the DESeq2 results. Prepare the ranked gene list for GSEA

```{r}
# ==============================================================================
# Prepare Ranked Gene List for GSEA
# ==============================================================================

# Filter low-expressed genes (baseMean > 15)
# baseMean: Mean normalized counts across all samples
dge_results_1 <- dge_results %>%
  filter(baseMean > 15)

# Extract gene names and stat values only
dge_results_rank <- dge_results_1[c("gene", "stat")]

# Handle empty strings (convert to NA)
dge_results_rank[dge_results_rank == ""] <- NA

# Remove rows with NA values
dge_results_rank <- na.omit(dge_results_rank)

# Sort by stat value (highest to lowest)
# Positive stat = upregulated in KO
# Negative stat = downregulated in KO
dge_results_rank <- dge_results_rank[order(-dge_results_rank$stat),]

# Create named vector for fgsea
# fgsea requires: named numeric vector where names are genes
res <- dge_results_rank$stat
names(res) <- dge_results_rank$gene

# Check number of genes in ranked list
length(res)
```

# ==============================================================================
# SECTION 5: RUN FGSEA ANALYSIS
# ==============================================================================

# Gene Set Enrichment Analysis (GSEA)

## Running fgsea

Perform GSEA using the fgsea package with the ranked gene list and MSigDB pathways.

```{r,results='hide'}
# ==============================================================================
# Run fgsea for All Gene Set Collections
# ==============================================================================

# fgsea(): Fast preranked GSEA
# pathways: List of gene sets (from GMT files)
# stats: Named vector of gene-level statistics (ranked list)
# minSize = 15: Minimum gene set size (exclude smaller sets)
# maxSize = 500: Maximum gene set size (exclude larger sets)
# nproc = 4: Number of parallel processes

fgseaRes_mh_hallmark <- 
  fgsea(pathways = mh_hallmark, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m1_positional <-
  fgsea(pathways = m1_positional, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m2_cgp <-
  fgsea(pathways = m2_curated_cgp, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m2_cp_biocarta <-
  fgsea(pathways = m2_cp_biocarta, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m2_cp_reactome <-
  fgsea(pathways = m2_cp_reactome, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m2_cp_wikipathways <-
  fgsea(pathways = m2_cp_wikipathways, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m3_miRDB <-
  fgsea(pathways = m3_miRDB, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m3_gtrd <-
  fgsea(pathways = m3_gtrd, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m5_go_bp <-
  fgsea(pathways = m5_go_bp, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m5_go_cc <-
  fgsea(pathways = m5_go_cc, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m5_go_mf <-
  fgsea(pathways = m5_go_mf, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m5_mpt <-
  fgsea(pathways = m5_go_mpt, stats = res, minSize=15, maxSize=500, nproc = 4)

fgseaRes_m8_celltype <-
  fgsea(pathways = m8_celltype, stats = res, minSize=15, maxSize=500, nproc = 4)

# ==============================================================================
# Sort Results by Adjusted P-value
# ==============================================================================

# order(padj): Sort by significance (most significant first)
fgseaRes_mh_hallmark <- fgseaRes_mh_hallmark[order(padj),]
fgseaRes_m1_positional <- fgseaRes_m1_positional[order(padj),]
fgseaRes_m2_cgp <- fgseaRes_m2_cgp[order(padj),]
fgseaRes_m2_cp_biocarta <- fgseaRes_m2_cp_biocarta[order(padj),]
fgseaRes_m2_cp_reactome <- fgseaRes_m2_cp_reactome[order(padj),]
fgseaRes_m2_cp_wikipathways <- fgseaRes_m2_cp_wikipathways[order(padj),]
fgseaRes_m3_miRDB <- fgseaRes_m3_miRDB[order(padj),]
fgseaRes_m3_gtrd <- fgseaRes_m3_gtrd[order(padj),]
fgseaRes_m5_go_bp <- fgseaRes_m5_go_bp[order(padj),]
fgseaRes_m5_go_cc <- fgseaRes_m5_go_cc[order(padj),]
fgseaRes_m5_go_mf <- fgseaRes_m5_go_mf[order(padj),]
fgseaRes_m5_mpt <- fgseaRes_m5_mpt[order(padj),]
fgseaRes_m8_celltype <- fgseaRes_m8_celltype[order(padj),]

```

# ==============================================================================
# SECTION 6: SAVE UNFILTERED RESULTS
# ==============================================================================

```{r}
# ==============================================================================
# Save All GSEA Results (Unfiltered)
# ==============================================================================

# saveRDS(): Save R objects for later use
saveRDS(fgseaRes_mh_hallmark,"./fgseaRes_mh_hallmark.RDS")
saveRDS(fgseaRes_m1_positional,"./fgseaRes_m1_positional.RDS")
saveRDS(fgseaRes_m2_cgp,"./fgseaRes_m2_cgp.RDS")
saveRDS(fgseaRes_m2_cp_biocarta,"./fgseaRes_m2_cp_biocarta.RDS")
saveRDS(fgseaRes_m2_cp_reactome,"./fgseaRes_m2_cp_reactome.RDS")
saveRDS(fgseaRes_m2_cp_wikipathways,"./fgseaRes_m2_cp_wikipathways.RDS")
saveRDS(fgseaRes_m3_miRDB,"./fgseaRes_m3_miRDB.RDS")
saveRDS(fgseaRes_m3_gtrd,"./fgseaRes_m3_gtrd.RDS")
saveRDS(fgseaRes_m5_go_bp,"./fgseaRes_m5_go_bp.RDS")
saveRDS(fgseaRes_m5_go_cc,"./fgseaRes_m5_go_cc.RDS")
saveRDS(fgseaRes_m5_go_mf,"./fgseaRes_m5_go_mf.RDS")
saveRDS(fgseaRes_m5_mpt,"./fgseaRes_m5_mpt.RDS")
saveRDS(fgseaRes_m8_celltype,"./fgseaRes_m8_celltype.RDS")
```

# ==============================================================================
# SECTION 7: FILTER SIGNIFICANT PATHWAYS
# ==============================================================================

## Filtering the GSEA Results with padj

Apply a cutoff to filter pathways based on adjusted p-value (padj).

```{r}
# ==============================================================================
# Filter Pathways by padj < 0.05
# ==============================================================================

# Keep only significantly enriched pathways
fgseaRes_mh_hallmark <-
  fgseaRes_mh_hallmark%>%
  filter(padj<0.05)

fgseaRes_m1_positional <-
  fgseaRes_m1_positional%>%
  filter(padj<0.05)
  
fgseaRes_m2_cgp <-
  fgseaRes_m2_cgp%>%
  filter(padj<0.05)
  
fgseaRes_m2_cp_biocarta <-
  fgseaRes_m2_cp_biocarta%>%
  filter(padj<0.05)

fgseaRes_m2_cp_reactome <-
  fgseaRes_m2_cp_reactome%>%
  filter(padj<0.05)
  
fgseaRes_m2_cp_wikipathways <-
  fgseaRes_m2_cp_wikipathways%>%
  filter(padj<0.05)
  
fgseaRes_m3_miRDB <-
  fgseaRes_m3_miRDB%>%
  filter(padj<0.05)

fgseaRes_m3_gtrd <-
  fgseaRes_m3_gtrd%>%
  filter(padj<0.05)
  
fgseaRes_m5_go_bp <-
  fgseaRes_m5_go_bp%>%
  filter(padj<0.05)
  
fgseaRes_m5_go_cc <-
  fgseaRes_m5_go_cc%>%
  filter(padj<0.05)
  
fgseaRes_m5_go_mf <-
  fgseaRes_m5_go_mf%>%
  filter(padj<0.05)
  
fgseaRes_m5_mpt <-
  fgseaRes_m5_mpt%>%
  filter(padj<0.05)
  
fgseaRes_m8_celltype <-
  fgseaRes_m8_celltype%>%
  filter(padj<0.05)
```

# ==============================================================================
# SECTION 8: SAVE FILTERED RESULTS
# ==============================================================================

```{r}
# ==============================================================================
# Save Filtered GSEA Results
# ==============================================================================

saveRDS(fgseaRes_mh_hallmark,"./fgseaRes_mh_hallmark_filtered.RDS")
saveRDS(fgseaRes_m1_positional,"./fgseaRes_m1_positional_filtered.RDS")
saveRDS(fgseaRes_m2_cgp,"./fgseaRes_m2_cgp_filtered.RDS")
saveRDS(fgseaRes_m2_cp_biocarta,"./fgseaRes_m2_cp_biocarta_filtered.RDS")
saveRDS(fgseaRes_m2_cp_reactome,"./fgseaRes_m2_cp_reactome_filtered.RDS")
saveRDS(fgseaRes_m2_cp_wikipathways,"./fgseaRes_m2_cp_wikipathways_filtered.RDS")
saveRDS(fgseaRes_m3_miRDB,"./fgseaRes_m3_miRDB_filtered.RDS")
saveRDS(fgseaRes_m3_gtrd,"./fgseaRes_m3_gtrd_filtered.RDS")
saveRDS(fgseaRes_m5_go_bp,"./fgseaRes_m5_go_bp_filtered.RDS")
saveRDS(fgseaRes_m5_go_cc,"./fgseaRes_m5_go_cc_filtered.RDS")
saveRDS(fgseaRes_m5_go_mf,"./fgseaRes_m5_go_mf_filtered.RDS")
saveRDS(fgseaRes_m5_mpt,"./fgseaRes_m5_mpt_filtered.RDS")
saveRDS(fgseaRes_m8_celltype,"./fgseaRes_m8_celltype_filtered.RDS")
```

# ==============================================================================
# SECTION 9: VISUALIZATION - BUBBLE PLOTS
# ==============================================================================

## Visualization of GSEA Results

Generate visualizations (e.g., enrichment plots) to summarize GSEA results.

```{r,fig.width=10,fig.height=10}
# ==============================================================================
# Basic Bubble Plot - Hallmark Pathways
# ==============================================================================

# Bubble plot: x = -log10(padj), size = gene set size, fill = NES
# NES (Normalized Enrichment Score): Positive = upregulated, Negative = downregulated
ggplot(fgseaRes_mh_hallmark[1:50,], aes(reorder(pathway, -log10(padj)), -log10(padj), size=size)) +
  geom_point(aes(fill=NES), shape=21) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
  labs(title ="Hallmarks", x = "", y = "-log10(padj)") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size=10),
        axis.text.y.top = element_text(margin = margin(t = 20)),
        axis.text.y = element_text(size=10),
        title = element_text(size=18),
        legend.position = "right")
```

```{r,fig.width=15, fig.height=20}
# ==============================================================================
# Generate Bubble Plots for All Collections
# ==============================================================================

# Reusable function for consistent plotting
generate_plot <- function(data, title) {
  ggplot(data[1:100,], aes(reorder(pathway, -log10(padj)), -log10(padj), size=size)) +
    geom_point(aes(fill=NES), shape=21) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
    scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
    labs(title = title, x = "", y = "-log10(padj)") + 
    theme_minimal() +
    theme(axis.text.x = element_text(size=10),
          axis.text.x.top = element_text(margin = margin(t = 20)),
          axis.text.y = element_text(size=10),
          title = element_text(size=18),
          legend.position = "right")
}

# Generate plots for all collections
print(generate_plot(fgseaRes_mh_hallmark, "Hallmarks"))
print(generate_plot(fgseaRes_m1_positional, "Positional"))
print(generate_plot(fgseaRes_m2_cgp, "CGP"))
print(generate_plot(fgseaRes_m2_cp_biocarta, "BioCarta"))
print(generate_plot(fgseaRes_m2_cp_reactome, "Reactome"))
print(generate_plot(fgseaRes_m2_cp_wikipathways, "WikiPathways"))
print(generate_plot(fgseaRes_m3_miRDB, "miRDB"))
print(generate_plot(fgseaRes_m3_gtrd, "GTRD"))
print(generate_plot(fgseaRes_m5_go_bp, "GO BP"))
print(generate_plot(fgseaRes_m5_go_cc, "GO CC"))
print(generate_plot(fgseaRes_m5_go_mf, "GO MF"))
print(generate_plot(fgseaRes_m5_mpt, "MPT"))
print(generate_plot(fgseaRes_m8_celltype, "Celltype"))

# ==============================================================================
# Export All Plots to PDF
# ==============================================================================

pdf("gsea_results_plots.pdf",width = 15, height = 20)
print(generate_plot(fgseaRes_mh_hallmark, "Hallmarks"))
print(generate_plot(fgseaRes_m1_positional, "Positional"))
print(generate_plot(fgseaRes_m2_cgp, "CGP"))
print(generate_plot(fgseaRes_m2_cp_biocarta, "BioCarta"))
print(generate_plot(fgseaRes_m2_cp_reactome, "Reactome"))
print(generate_plot(fgseaRes_m2_cp_wikipathways, "WikiPathways"))
print(generate_plot(fgseaRes_m3_miRDB, "miRDB"))
print(generate_plot(fgseaRes_m3_gtrd, "GTRD"))
print(generate_plot(fgseaRes_m5_go_bp, "GO BP"))
print(generate_plot(fgseaRes_m5_go_cc, "GO CC"))
print(generate_plot(fgseaRes_m5_go_mf, "GO MF"))
print(generate_plot(fgseaRes_m5_mpt, "MPT"))
print(generate_plot(fgseaRes_m8_celltype, "Celltype"))
dev.off()
```

# ==============================================================================
# SECTION 10: SEPARATE UP/DOWN REGULATED PATHWAYS
# ==============================================================================

```{r}
# ==============================================================================
# Split Hallmark Pathways by Direction
# ==============================================================================

# NES < 0: Pathway genes are downregulated (enriched at bottom of ranked list)
fgseaRes_mh_hallmark_down <-
  fgseaRes_mh_hallmark%>%
  filter(NES< 0)

fgseaRes_mh_hallmark_down$pathway

# NES > 0: Pathway genes are upregulated (enriched at top of ranked list)
fgseaRes_mh_hallmark_up <-
  fgseaRes_mh_hallmark%>%
  filter(NES> 0)

fgseaRes_mh_hallmark_up$pathway
```

```{r}
# ==============================================================================
# Split Reactome Pathways by Direction
# ==============================================================================

fgseaRes_m2_cp_reactome_down <-
  fgseaRes_m2_cp_reactome%>%
  filter(NES< 0)

fgseaRes_m2_cp_reactome_down$pathway


fgseaRes_m2_cp_reactome_up <-
  fgseaRes_m2_cp_reactome%>%
  filter(NES> 0)

fgseaRes_m2_cp_reactome_up$pathway
```

```{r}
# ==============================================================================
# Split GO Biological Process by Direction
# ==============================================================================

fgseaRes_m5_go_bp_down <-
  fgseaRes_m5_go_bp%>%
  filter(NES< 0)

fgseaRes_m5_go_bp_down$pathway


fgseaRes_m5_go_bp_up <-
  fgseaRes_m5_go_bp%>%
  filter(NES> 0)

fgseaRes_m5_go_bp_up$pathway
```

# ==============================================================================
# SECTION 11: ENRICHMENT PLOT FUNCTION
# ==============================================================================

## Selecting

```{r}
# ==============================================================================
# Initialize Pathways Storage and Define Enrichment Plot Function
# ==============================================================================

# Create list to store selected pathway results
if (!exists("pathways")) {
  pathways <- list()
}

# generate_enrichment_plot(): Create and save individual enrichment plots
# pathway_name: Name of pathway to plot
# fgsea_result_table: GSEA results data frame
# ranked_list: Named vector used for GSEA
# mh_hallmark: Pathway gene sets list
# output_directory: Where to save PDFs
generate_enrichment_plot <- function(pathway_name, fgsea_result_table, ranked_list,
                                     mh_hallmark, output_directory = "./", 
                                     plot_width = 5, plot_height = 3) {
  # Extract pathway statistics
  stats <- fgsea_result_table[fgsea_result_table$pathway == pathway_name, ]
  
  if(nrow(stats) == 0) {
    stop("Pathway not found in the fgsea result table.")
  }

  # plotEnrichment(): Classic GSEA running enrichment plot
  # Shows: Running enrichment score, gene positions, leading edge
  plot <- plotEnrichment(mh_hallmark[[pathway_name]], ranked_list, ticksSize = 0.2) +
    labs(title = pathway_name,
         caption = paste("padj: ", formatC(stats$padj, format = "e", digits = 2),
                         "NES: ", round(stats$NES, 2),
                         "Size: ", stats$size)) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          axis.title.x = element_text(size = 12),
          axis.title.y = element_text(size = 12))

  # Save plot to PDF
  ggsave(
    filename = paste0(output_directory, "/", pathway_name, ".pdf"),
    plot = plot,
    width = plot_width,
    height = plot_height,
    units = "in",
    dpi = 300,
    limitsize = TRUE
  )
  
  print(plot)
  
  # Store pathway stats in global list
  # <<-: Assign to global environment
  pathways[[pathway_name]] <<- stats
  
  # Return leading edge genes
  return(stats$leadingEdge)
}
```

# ==============================================================================
# SECTION 12: INDIVIDUAL PATHWAY ENRICHMENT PLOTS
# ==============================================================================

```{r,fig.width=5,fig.height=3}
# ==============================================================================
# Generate Enrichment Plots for Selected Pathways
# ==============================================================================

generate_enrichment_plot("HALLMARK_TGF_BETA_SIGNALING",
                         fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_TNFA_SIGNALING_VIA_NFKB", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_KRAS_SIGNALING_UP",
                         fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_WNT_BETA_CATENIN_SIGNALING",
                         fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_NOTCH_SIGNALING", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_MITOTIC_SPINDLE", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_APOPTOSIS", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_ESTROGEN_RESPONSE_EARLY", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_HYPOXIA", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_INTERFERON_GAMMA_RESPONSE", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_INTERFERON_ALPHA_RESPONSE", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_IL6_JAK_STAT3_SIGNALING", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_OXIDATIVE_PHOSPHORYLATION", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_GLYCOLYSIS", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_FATTY_ACID_METABOLISM", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_ADIPOGENESIS", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_BILE_ACID_METABOLISM", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_PEROXISOME", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_PROTEIN_SECRETION", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_MTORC1_SIGNALING", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

HALLMARK_ANGIOGENESIS HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION HALLMARK_MYOGENESIS

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_ANGIOGENESIS", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_MYOGENESIS", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

```{r,fig.width=5,fig.height=3}
generate_enrichment_plot("HALLMARK_UV_RESPONSE_DN", fgseaRes_mh_hallmark, res, mh_hallmark)
```

# ==============================================================================
# SECTION 13: COMBINED VISUALIZATION
# ==============================================================================

```{r}
# ==============================================================================
# Load Additional Packages
# ==============================================================================

library(stringr)  # String manipulation (str_wrap)
```

```{r}
# ==============================================================================
# Load Pre-saved Results
# ==============================================================================

pathways <- readRDS("./pathways.RDS")
```

```{r}
# Load filtered results
fgseaRes_mh_hallmark_filtered <- readRDS("./fgseaRes_mh_hallmark_filtered.RDS")
fgseaRes_m5_go_bp_filtered <- readRDS("./fgseaRes_m5_go_bp_filtered.RDS")
fgseaRes_m2_cp_reactome <- readRDS("./fgseaRes_m2_cp_reactome_filtered.RDS")
```

```{r}
# ==============================================================================
# Combine Results from Multiple Collections
# ==============================================================================

# rbind(): Combine rows from different gene set collections
gsea <- rbind(fgseaRes_mh_hallmark_filtered,fgseaRes_m5_go_bp_filtered,fgseaRes_m2_cp_reactome)
```

# ==============================================================================
# SECTION 14: PUBLICATION-QUALITY PLOTS
# ==============================================================================

```{r,fig.width=6.5,fig.height=5.5}
# ==============================================================================
# Publication-Quality Hallmark Plot
# ==============================================================================

set.seed(123)

# wrap_text(): Wrap long pathway names and replace underscores with spaces
wrap_text <- function(y) str_wrap(gsub("_", " ", y), width = 45)

# Create publication-quality bubble plot
pl_emp_hall <-
  ggplot(fgseaRes_mh_hallmark_filtered[1:20,], aes(y=reorder(pathway, -log10(padj)), x=-log10(padj), size=size)) +
  geom_point(aes(fill=NES), shape=21) +
  scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
  labs(title ="EMP GSEA:HALLMARK",y="") + 
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black", face = "bold"),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.text.y = element_text(size = 8, margin = margin(t = 0, r = 0, b = 0, l = 0)),
    legend.position = "inside",
    legend.position.inside = c(0.92, 0.45),
    legend.background = element_rect(fill = "transparent"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    axis.line = element_line(size = 0.2)
  ) +
  scale_y_discrete(labels = function(y) wrap_text(y)) +
  labs(size="Genes")

print(pl_emp_hall)
```


```{r,fig.width=6.5,fig.height=5.5}
# Duplicate plot (same as above)
set.seed(123)
wrap_text <- function(y) str_wrap(gsub("_", " ", y), width = 45)

pl_emp_hall <-
  ggplot(fgseaRes_mh_hallmark_filtered[1:20,], aes(y=reorder(pathway, -log10(padj)), x=-log10(padj), size=size)) +
  geom_point(aes(fill=NES), shape=21) +
  scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
  labs(title ="EMP GSEA:HALLMARK",y="") + 
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black", face = "bold"),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.text.y = element_text(size = 8, margin = margin(t = 0, r = 0, b = 0, l = 0)),
    legend.position = "inside",
    legend.position.inside = c(0.92, 0.45),
    legend.background = element_rect(fill = "transparent"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    axis.line = element_line(size = 0.2)
  ) +
  scale_y_discrete(labels = function(y) wrap_text(y)) +
  labs(size="Genes")

print(pl_emp_hall)
```

```{r,fig.width=6.5,fig.height=5.5}
# ==============================================================================
# Publication-Quality GO Biological Process Plot
# ==============================================================================

set.seed(123)
wrap_text <- function(y) str_wrap(gsub("_", " ", y), width = 45)

pl_emp_gobp <-
  ggplot(fgseaRes_m5_go_bp_filtered[1:20,], aes(y=reorder(pathway, -log10(padj)), x=-log10(padj), size=size)) +
  geom_point(aes(fill=NES), shape=21) +
  scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
  labs(title ="EMP GSEA:GO-BP",y="") + 
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black", face = "bold"),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.text.y = element_text(size = 8, margin = margin(t = 0, r = 0, b = 0, l = 0)),
    legend.position = "inside",
    legend.position.inside = c(0.92, 0.45),
    legend.background = element_rect(fill = "transparent"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    axis.line = element_line(size = 0.2)
  ) +
  scale_y_discrete(labels = function(y) wrap_text(y)) +
  labs(size="Genes")

print(pl_emp_gobp)
```

```{r,fig.width=6.5,fig.height=5.5}
# ==============================================================================
# Publication-Quality Reactome Plot
# ==============================================================================

set.seed(123)
wrap_text <- function(y) str_wrap(gsub("_", " ", y), width = 45)

pl_emp_reactome <- 
  ggplot(fgseaRes_m2_cp_reactome[1:20,], aes(y=reorder(pathway, -log10(padj)), x=-log10(padj), size=size)) +
  geom_point(aes(fill=NES), shape=21) +
  scale_fill_gradient2(low="red", mid="white", high="blue", midpoint=0) +
  labs(title ="EMP GSEA-Reactome",y="") + 
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black", face = "bold"),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.text.y = element_text(size = 8, margin = margin(t = 0, r = 0, b = 0, l = 0)),
    legend.position = "inside",
    legend.position.inside = c(0.92, 0.45),
    legend.background = element_rect(fill = "transparent"),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    axis.line = element_line(size = 0.2)
  ) +
  scale_y_discrete(labels = function(y) wrap_text(y)) +
  labs(size="Genes")

print(pl_emp_reactome)

```

# ==============================================================================
# SECTION 15: SESSION INFORMATION
# ==============================================================================

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================

sessionInfo()
```
