---
title: "Differential Gene Expression Analysis with DESeq2"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---


Pseudobulked differential gene expression analysis with DESeq2. Process will be repeated for each annotations in groups, clusters and sub-clusters. And number of significant genes will be stored in a table as a summary table.

gene_info can be found in either in github or in the gse.

# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP AND PACKAGE LOADING
# ==============================================================================

```{r eval=FALSE}
# ==============================================================================
# Load Required Packages
# ==============================================================================

suppressPackageStartupMessages({
  library(Seurat)      # Single-cell analysis and pseudobulk aggregation
  library(DESeq2)      # Differential expression analysis (negative binomial model)
  library(dplyr)       # Data manipulation
  library(tibble)      # Enhanced data frames
  library(ggplot2)     # Visualization
  library(ggrepel)     # Non-overlapping text labels
  library(ggvenn)      # Venn diagrams with ggplot2
  library(venn)        # Venn diagram utilities
  library(tidyverse)   # Tidyverse ecosystem
  library(VennDiagram) # Publication-quality Venn diagrams
  library(cowplot)     # Plot arrangement
  library(UpSetR)      # UpSet plots for set intersections
})
```


# ==============================================================================
# SECTION 2: DATA LOADING
# ==============================================================================

# Data Preprocessing

Example code for analysis. It will be repeated for conditions which desired.

```{r  eval=FALSE}
# ==============================================================================
# Load Seurat Object and Gene Annotation
# ==============================================================================

# Load the Atf3 KO vs Control subset
Atf3NT <- readRDS("~/cmo_chapter_v1/Analysis/RDS/CMO/cmo_CA.RDS")

# Load gene annotation file (gene symbol to Ensembl ID mapping)
# Extracted from CellRanger's annotation file
gene_info <- read.csv("~/cmo_chapter/figs/DEG/gene_info.csv")
# Gene_info is extracted from cellranger's annotation file contains gene symbol and ensmbl name. 
```

# ==============================================================================
# SECTION 3: PSEUDOBULK AGGREGATION - SUBCLUSTER LEVEL (cl_3)
# ==============================================================================

```{r  eval=FALSE}
# ==============================================================================
# Setup Analysis Parameters
# ==============================================================================

# Set active identity to subcluster level
Idents(Atf3NT) <- "cl_3"

# Define the gene being knocked out (used to identify KO samples)
gene_of_interest <- "Atf3"

# Create output directory for results
output_folder <- "DESeq2_results_Atf3_cl_3"
dir.create(output_folder, showWarnings = FALSE)

# ==============================================================================
# Clean Factor Levels
# ==============================================================================

# droplevels(): Remove unused factor levels from previous subsetting
# Critical for correct pseudobulk aggregation
Atf3NT$Samples <- droplevels(Atf3NT$Samples_lv)
Atf3NT$c_groups <- droplevels(Atf3NT$cl_3)

# Get unique cluster names for iteration
cluster_names <- unique(Atf3NT$c_groups)

# ==============================================================================
# Calculate Average Expression of Target Gene
# ==============================================================================

# AverageExpression(): Calculate mean expression per cluster
# Used to track target gene expression across clusters
exp <- AverageExpression(Atf3NT, features = "Atf3")
exp <- exp$RNA
exp <- as.data.frame(exp)

# ==============================================================================
# Create Pseudobulk Count Matrix
# ==============================================================================

# AggregateExpression(): Sum counts across cells within each group
# group.by: Aggregate by cluster AND sample (for replicates)
# slot = "counts": Use raw counts (required for DESeq2)
# return.seurat = F: Return matrix, not Seurat object
cts <- AggregateExpression(Atf3NT, group.by = c("c_groups", "Samples"),
                           assays = "RNA",
                           slot = "counts",
                           return.seurat = F)

# ==============================================================================
# Reshape Pseudobulk Matrix for DESeq2
# ==============================================================================

# Extract RNA counts
cts <- cts$RNA

# Transpose: genes become rows, samples become columns
cts.t <- t(cts)
cts.t <- as.data.frame(cts.t)

# Extract cluster names from row names (format: "ClusterName_SampleName")
# gsub('_.*', '', ...): Remove everything after first underscore
splitRows <- gsub('_.*', '', rownames(cts.t))

# Split data frame by cluster
# Creates list: one element per cluster
cts.split <- split.data.frame(cts.t, f = factor(splitRows))

# Fix row names (keep only sample name) and transpose back
# lapply: Apply function to each cluster's data frame
cts.split.modified <- lapply(cts.split, function(x){
  # gsub('.*_(.*)', '\\1', ...): Keep only text after underscore
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)  # Transpose: genes as rows, samples as columns
})
```

# ==============================================================================
# SECTION 4: DESEQ2 ANALYSIS LOOP - SUBCLUSTER LEVEL
# ==============================================================================

```{r  eval=FALSE}
# ==============================================================================
# Initialize Summary Data Frame
# ==============================================================================

# Create empty data frame to store DEG counts across all clusters
# Tracks multiple significance thresholds and directions
summary_dataframe <- data.frame(
  cluster = character(),
  average_expression = integer(),      # Target gene expression in cluster
  total_genes = integer(),             # Total genes before filtering
  included_genes = integer(),          # Genes passing filter
  excluded_genes = integer(),          # Genes filtered out
  padj_0_01 = integer(),               # DEGs at padj < 0.01
  upregulated_genes_padj_0_01 = integer(),
  downregulated_genes_padj_0_01 = integer(),
  padj_0_05 = integer(),               # DEGs at padj < 0.05
  upregulated_genes_padj_0_05 = integer(),
  downregulated_genes_padj_0_05 = integer(),
  padj_0_1 = integer(),                # DEGs at padj < 0.1
  upregulated_genes_padj_0_1 = integer(),
  downregulated_genes_padj_0_1 = integer(),
  pvalue_0_05 = integer(),             # DEGs at raw p < 0.05
  upregulated_genes_pvalue_0_05 = integer(),
  downregulated_genes_pvalue_0_05 = integer(),
  stringsAsFactors = FALSE
)

# ==============================================================================
# Loop Through Each Cluster
# ==============================================================================

for (cluster_name in cluster_names) {
  if (cluster_name %in% names(cts.split.modified)) {
    # Extract count matrix for this cluster
    counts_cluster <- cts.split.modified[[cluster_name]]

    # Get average expression of target gene in this cluster
    avg_exp <- exp[[cluster_name]]
    
    # ==============================================================================
    # Create Sample Metadata (colData)
    # ==============================================================================
    
    # Create data frame with sample names
    colData <- data.frame(samples = colnames(counts_cluster))
    
    # Assign condition based on sample name
    # grepl(): Check if gene name appears in sample name
    # KO samples contain the gene name, NT (Non-Targeting) do not
    colData <- colData %>%
      mutate(condition = ifelse(grepl(gene_of_interest, samples), 'KO', 'NT')) %>%
      column_to_rownames(var = 'samples')
    
    # ==============================================================================
    # Create DESeq2 Dataset
    # ==============================================================================
    
    # DESeqDataSetFromMatrix(): Create DESeq2 object from count matrix
    # countData: Gene x Sample count matrix
    # colData: Sample metadata with conditions
    # design: Formula specifying the comparison (KO vs NT)
    dds <- DESeqDataSetFromMatrix(countData = counts_cluster,
                                  colData = colData,
                                  design = ~ condition)

    # ==============================================================================
    # Pre-filtering Low-Count Genes
    # ==============================================================================
    
    # Filter genes with very low counts (improves power)
    smallestGroupSize <- 3  # Minimum replicates per condition
    total_genes_before <- nrow(dds)
    
    # Keep genes with >= 10 counts in at least 3 samples
    keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
    dds <- dds[keep,]
    total_genes_after <- nrow(dds)
    
    # ==============================================================================
    # Run DESeq2 Analysis
    # ==============================================================================
    
    # relevel(): Set reference level for comparison
    # ref = "NT": Non-targeting control as reference
    # Positive logFC = higher in KO, Negative logFC = lower in KO
    dds$condition <- relevel(dds$condition, ref = "NT")
    
    # DESeq(): Run full DESeq2 pipeline
    # 1. Estimate size factors (normalization)
    # 2. Estimate dispersions
    # 3. Fit negative binomial GLM
    # 4. Wald test for significance
    dds <- DESeq(dds)
    
    # results(): Extract results table
    res <- results(dds)
    
    # Order by p-value (most significant first)
    resOrdered <- res[order(res$pvalue),]
    
    # Convert to tibble with gene names as column
    resOrdered <- resOrdered %>%
      data.frame() %>%
      rownames_to_column(var = "gene") %>%
      as_tibble()
    
    # ==============================================================================
    # Merge with Gene Annotation
    # ==============================================================================
    
    # Add Ensembl IDs and other gene info
    # merge(): Left join to keep all DESeq2 results
    resOrdered_enhanced <- merge(resOrdered, gene_info, 
                                 by.x = "gene", by.y = "gene_symbol", 
                                 all.x = TRUE)

    # ==============================================================================
    # Save Results
    # ==============================================================================
    
    # Save DESeq2 results object (for further analysis)
    save(res, file = paste0(output_folder, "/", cluster_name, "_res.RData"))
    
    # Save enhanced results with gene info
    save(resOrdered_enhanced, file = paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.RData"))
    
    # Export as tab-separated text
    write.table(resOrdered_enhanced, 
                file = paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.txt"), 
                sep = '\t', quote = F, row.names = F)
    
    # Save DESeq2 dataset object (for QC plots)
    save(dds, file = paste0(output_folder, "/", cluster_name, "_dds.RData"))
    
    # Save raw pseudobulk counts
    write.table(counts_cluster, 
                file = paste0(output_folder, "/", cluster_name, "_counts_cluster_table.txt"), 
                sep = '\t', quote = F, row.names = F)

    # ==============================================================================
    # Calculate Summary Statistics
    # ==============================================================================
    
    # Handle NA values (genes with no valid test)
    complete_cases <- complete.cases(resOrdered_enhanced$padj, 
                                     resOrdered_enhanced$log2FoldChange)

    # Add row to summary data frame with DEG counts at multiple thresholds
    summary_dataframe <- rbind(summary_dataframe, 
                               data.frame(
                                 cluster = cluster_name,
                                 average_expression = avg_exp,
                                 total_genes = total_genes_before,
                                 included_genes = total_genes_after,
                                 excluded_genes = total_genes_before - total_genes_after,
                                 # padj < 0.01 threshold
                                 padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01, na.rm = TRUE),
                                 upregulated_genes_padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 # padj < 0.05 threshold
                                 padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05, na.rm = TRUE),
                                 upregulated_genes_padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 # padj < 0.1 threshold
                                 padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1, na.rm = TRUE),
                                 upregulated_genes_padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 # Raw p-value < 0.05 (less stringent)
                                 pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05, na.rm = TRUE),
                                 upregulated_genes_pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE)
                               ))
  }
}
```

# ==============================================================================
# SECTION 5: EXPORT SUBCLUSTER SUMMARY
# ==============================================================================

```{r}
# ==============================================================================
# Save Summary Statistics
# ==============================================================================

# Export summary of DEG counts across all subclusters
write.csv(summary_dataframe, 
          file = paste0(output_folder, "/summary_dataframe.csv"), 
          row.names = FALSE)
```

# ==============================================================================
# SECTION 6: PSEUDOBULK AGGREGATION - CLUSTER LEVEL (cl_2)
# ==============================================================================

```{r  eval=FALSE}
# ==============================================================================
# Switch to Cluster Level Analysis
# ==============================================================================

# Change identity to cluster level (coarser than subclusters)
Idents(Atf3NT) <- "cl_2"
levels(Atf3NT)
```

```{r eval=FALSE}
# ==============================================================================
# Initialize Summary Data Frame for Cluster Level
# ==============================================================================

# Same structure as subcluster analysis
summary_dataframe <- data.frame(
  cluster = character(),
  average_expression = integer(),
  total_genes = integer(),
  included_genes = integer(),
  excluded_genes = integer(),
  padj_0_01 = integer(),
  upregulated_genes_padj_0_01 = integer(),
  downregulated_genes_padj_0_01 = integer(),
  padj_0_05 = integer(),
  upregulated_genes_padj_0_05 = integer(),
  downregulated_genes_padj_0_05 = integer(),
  padj_0_1 = integer(),
  upregulated_genes_padj_0_1 = integer(),
  downregulated_genes_padj_0_1 = integer(),
  pvalue_0_05 = integer(),
  upregulated_genes_pvalue_0_05 = integer(),
  downregulated_genes_pvalue_0_05 = integer(),
  stringsAsFactors = FALSE
)
```

```{r eval=FALSE}
# ==============================================================================
# Setup Cluster Level Analysis
# ==============================================================================

# Set identity and parameters for cluster level
Idents(Atf3NT) <- "cl_2"
Atf3NT$ident <- Atf3NT$cl_2

gene_of_interest <- "Atf3"
output_folder <- "DESeq2_results_Atf3_cl_2"
dir.create(output_folder, showWarnings = FALSE)

# Clean factor levels
Atf3NT$Samples <- droplevels(Atf3NT$Samples_lv)
Atf3NT$c_groups <- droplevels(Atf3NT$cl_2)

# ==============================================================================
# Calculate Average Expression
# ==============================================================================

exp <- AverageExpression(Atf3NT, features = "Atf3")
exp <- exp$RNA
exp <- as.data.frame(exp)

# ==============================================================================
# Create Pseudobulk Counts
# ==============================================================================

cts <- AggregateExpression(Atf3NT, group.by = c("c_groups", "Samples"),
                           assays = "RNA",
                           slot = "counts",
                           return.seurat = F)

# Reshape for DESeq2 (same process as subcluster level)
cts <- cts$RNA
cts.t <- t(cts)
cts.t <- as.data.frame(cts.t)
splitRows <- gsub('_.*', '', rownames(cts.t))
cts.split <- split.data.frame(cts.t, f = factor(splitRows))
cts.split.modified <- lapply(cts.split, function(x){
  rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x))
  t(x)
})

# Get cluster names from the split list
cluster_names <- names(cts.split.modified)
```

# ==============================================================================
# SECTION 7: DESEQ2 ANALYSIS LOOP - CLUSTER LEVEL
# ==============================================================================

```{r  eval=FALSE}
# ==============================================================================
# Initialize Summary Data Frame
# ==============================================================================

summary_dataframe <- data.frame(
  cluster = character(),
  average_expression = integer(),
  total_genes = integer(),
  included_genes = integer(),
  excluded_genes = integer(),
  padj_0_01 = integer(),
  upregulated_genes_padj_0_01 = integer(),
  downregulated_genes_padj_0_01 = integer(),
  padj_0_05 = integer(),
  upregulated_genes_padj_0_05 = integer(),
  downregulated_genes_padj_0_05 = integer(),
  padj_0_1 = integer(),
  upregulated_genes_padj_0_1 = integer(),
  downregulated_genes_padj_0_1 = integer(),
  pvalue_0_05 = integer(),
  upregulated_genes_pvalue_0_05 = integer(),
  downregulated_genes_pvalue_0_05 = integer(),
  stringsAsFactors = FALSE
)

# ==============================================================================
# Loop Through Each Cluster (Same Logic as Subcluster Level)
# ==============================================================================

for (cluster_name in cluster_names) {
  if (cluster_name %in% names(cts.split.modified)) {
    counts_cluster <- cts.split.modified[[cluster_name]]

    # Get average expression
    avg_exp <- exp[[cluster_name]]

    # Create sample metadata
    colData <- data.frame(samples = colnames(counts_cluster))
    colData <- colData %>%
      mutate(condition = ifelse(grepl(gene_of_interest, samples), 'KO', 'NT')) %>%
      column_to_rownames(var = 'samples')
    
    # Create DESeq2 object
    dds <- DESeqDataSetFromMatrix(countData = counts_cluster,
                                  colData = colData,
                                  design = ~ condition)

    # Pre-filtering
    smallestGroupSize <- 3
    total_genes_before <- nrow(dds)
    keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
    dds <- dds[keep,]
    total_genes_after <- nrow(dds)
    
    # DESeq2 analysis
    dds$condition <- relevel(dds$condition, ref = "NT")
    dds <- DESeq(dds)
    res <- results(dds)
    resOrdered <- res[order(res$pvalue),]
    
    resOrdered <- resOrdered %>%
      data.frame() %>%
      rownames_to_column(var = "gene") %>%
      as_tibble()
    
    # Merge with gene annotation
    resOrdered_enhanced <- merge(resOrdered, gene_info, 
                                 by.x = "gene", by.y = "gene_symbol", 
                                 all.x = TRUE)

    # Save results
    save(res, file = paste0(output_folder, "/", cluster_name, "_res.RData"))
    save(resOrdered_enhanced, file = paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.RData"))
    write.table(resOrdered_enhanced, 
                file = paste0(output_folder, "/", cluster_name, "_resOrdered_enhanced.txt"), 
                sep = '\t', quote = F, row.names = F)
    save(dds, file = paste0(output_folder, "/", cluster_name, "_dds.RData"))
    write.table(counts_cluster, 
                file = paste0(output_folder, "/", cluster_name, "_counts_cluster_table.txt"), 
                sep = '\t', quote = F, row.names = F)

    # Calculate summary statistics
    complete_cases <- complete.cases(resOrdered_enhanced$padj, 
                                     resOrdered_enhanced$log2FoldChange)

    # Add to summary
    summary_dataframe <- rbind(summary_dataframe, 
                               data.frame(
                                 cluster = cluster_name,
                                 average_expression = avg_exp,
                                 total_genes = total_genes_before,
                                 included_genes = total_genes_after,
                                 excluded_genes = total_genes_before - total_genes_after,
                                 padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01, na.rm = TRUE),
                                 upregulated_genes_padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_01 = sum(complete_cases & resOrdered_enhanced$padj < 0.01 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05, na.rm = TRUE),
                                 upregulated_genes_padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_05 = sum(complete_cases & resOrdered_enhanced$padj < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1, na.rm = TRUE),
                                 upregulated_genes_padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_padj_0_1 = sum(complete_cases & resOrdered_enhanced$padj < 0.1 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE),
                                 pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05, na.rm = TRUE),
                                 upregulated_genes_pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange > 0, na.rm = TRUE),
                                 downregulated_genes_pvalue_0_05 = sum(complete_cases & resOrdered_enhanced$pvalue < 0.05 & resOrdered_enhanced$log2FoldChange < 0, na.rm = TRUE)
                               ))
  }
}
```

# ==============================================================================
# SECTION 8: EXPORT CLUSTER SUMMARY
# ==============================================================================

```{r eval=FALSE}
# ==============================================================================
# Save Cluster Level Summary
# ==============================================================================

write.csv(summary_dataframe, 
          file = paste0(output_folder, "/summary_dataframe.csv"), 
          row.names = FALSE)
```

# ==============================================================================
# SECTION 9: SESSION INFORMATION
# ==============================================================================

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================

sessionInfo()
```
