---
title: "R Notebook"
output: html_notebook
---

# ==============================================================================
# PREP FOR PSEUDOTIME - PART 2: ATLAS ENDPOINT SELECTION
# ==============================================================================
# Purpose: Interactive selection of trajectory endpoint cells in the Atlas
# Method: CellSelector() for manual identification of terminal cells
# Output: Cell IDs for each lineage endpoint (used in Slingshot)
# ==============================================================================


# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP
# ==============================================================================
# Load Atlas and create UMAP for interactive selection

```{r}
# ------------------------------------------------------------------------------
# Load Extended Mouse Gastrulation Atlas
# ------------------------------------------------------------------------------
# 430,339 cells with celltype_extended_atlas annotations
library(Seurat)
atlas <- readRDS("~/scRIPT/000_io_dir/outputs/rds_files/extended_atlas_cellranger_features_seurat.rds")

# ------------------------------------------------------------------------------
# Create UMAP Plot for CellSelector
# ------------------------------------------------------------------------------
# DimPlot(): Visualize cells colored by cell type
# NoAxes(), NoLegend(): Clean display for selection
p <- DimPlot(atlas, reduction = "UMAP", group.by = 'celltype_extended_atlas',label = TRUE, alpha = 0.5)+ NoAxes()+NoLegend() 
```

```{r}
# ------------------------------------------------------------------------------
# Check Developmental Stage Distribution
# ------------------------------------------------------------------------------
table(atlas$stage)
```

```{r}

```


# ==============================================================================
# SECTION 2: BLOOD LINEAGE ENDPOINTS
# ==============================================================================


# EMP
# ------------------------------------------------------------------------------
# Erythro-Myeloid Progenitor - Terminal endpoint for EMP trajectory
# ------------------------------------------------------------------------------
```{r}
emp_candidates <- CellSelector(p)
emp_candidates
```
```{r}
# Verify cell type and stage of selected cell
atlas@meta.data[emp_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_328761 


# MkP
# ------------------------------------------------------------------------------
# Megakaryocyte Progenitor - Terminal endpoint for platelet lineage
# ------------------------------------------------------------------------------
```{r}
MkP_candidates <- CellSelector(p)
MkP_candidates
```
```{r}
atlas@meta.data[MkP_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_329685


# Erythroid
# ------------------------------------------------------------------------------
# Erythroid - Terminal endpoint for red blood cell lineage
# ------------------------------------------------------------------------------
```{r}
Erythroid_candidates <- CellSelector(p)
Erythroid_candidates
```
```{r}
atlas@meta.data[Erythroid_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_189604


# ==============================================================================
# SECTION 3: ENDOTHELIAL LINEAGE ENDPOINTS
# ==============================================================================


# YS Endothelium
# ------------------------------------------------------------------------------
# Yolk Sac Endothelium - Terminal endpoint for YS vascular lineage
# ------------------------------------------------------------------------------
```{r}
YSE_candidates <- CellSelector(p)
YSE_candidates
```
```{r}
atlas@meta.data[YSE_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_14984


# Endocardium
# ------------------------------------------------------------------------------
# Endocardium - Terminal endpoint for cardiac endothelial lineage
# ------------------------------------------------------------------------------
```{r}
Endocardium_candidates <- CellSelector(p)
Endocardium_candidates
```
```{r}
atlas@meta.data[Endocardium_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_343138


# ==============================================================================
# SECTION 4: MESODERM LINEAGE ENDPOINTS
# ==============================================================================


# YSM
# ------------------------------------------------------------------------------
# Yolk Sac Mesoderm - Terminal endpoint for YS mesoderm lineage
# ------------------------------------------------------------------------------
```{r}
YSM_candidates <- CellSelector(p)
YSM_candidates
```
```{r}
atlas@meta.data[YSM_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_343138


# Mesenchyme
# ------------------------------------------------------------------------------
# Mesenchyme - Terminal endpoint for mesenchymal lineage
# ------------------------------------------------------------------------------
```{r}
Mesenchyme_candidates <- CellSelector(p)
Mesenchyme_candidates

atlas@meta.data[Mesenchyme_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_325578


# Allantois
# ------------------------------------------------------------------------------
# Allantois - Terminal endpoint for allantois lineage
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_194268


# Epicardium
# ------------------------------------------------------------------------------
# Epicardium - Terminal endpoint for epicardial lineage
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_7917


# Cardiomyocytes
# ------------------------------------------------------------------------------
# Cardiomyocytes - Terminal endpoint for cardiac muscle lineage
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_172850


# Limb Mesoderm
# ------------------------------------------------------------------------------
# Limb Mesoderm - Terminal endpoint for limb bud mesoderm
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_318387


# Lateral Plate Mesoderm
# ------------------------------------------------------------------------------
# Lateral Plate Mesoderm - Terminal endpoint for LPM lineage
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
ext_cell_29525


# ==============================================================================
# SECTION 5: STARTING POINT
# ==============================================================================


# Epiblast
# ------------------------------------------------------------------------------
# Epiblast - Starting point for all trajectories
# ------------------------------------------------------------------------------
```{r}
Allantois_candidates <- CellSelector(p)
Allantois_candidates

atlas@meta.data[Allantois_candidates,c("celltype_extended_atlas","stage") ]
```
cell_5856


# ==============================================================================
# SECTION 6: VERIFY SELECTED CELLS
# ==============================================================================

```{r}
# ------------------------------------------------------------------------------
# Confirm Epiblast Starting Point Cells
# ------------------------------------------------------------------------------
atlas@meta.data[c('cell_5856',"cell_15629"),c("celltype_extended_atlas","stage") ]
```

