---
title: "Atf3 DEGs and Integration to other data"
output: html_notebook
---

# ==============================================================================
# SECTION 1: ENVIRONMENT SETUP
# ==============================================================================

```{r}
# ==============================================================================
# Set Working Directory
# ==============================================================================

setwd("~/cmo_chapter_v1/Analysis/3_dge/Atf3_0/4_DEGs_Categorization/1_Integration/")
getwd()
```

```{r}
# ==============================================================================
# Load Required Packages
# ==============================================================================

# suppressPackageStartupMessages(): Suppress loading messages for cleaner output
suppressPackageStartupMessages({
  library(dplyr)        # Data manipulation
  library(tibble)       # Tibble data frames
  library(ggplot2)      # Visualization
  library(ggrepel)      # Non-overlapping text labels
  library(ggvenn)       # Venn diagrams with ggplot2
  library(venn)         # Venn diagram calculations
  library(tidyverse)    # Tidyverse collection
  library(VennDiagram)  # Venn diagram generation
  library(cowplot)      # Plot grids and themes
  library(UpSetR)       # UpSet plots for set intersections
  library(stringr)      # String manipulation
})
```

# ==============================================================================
# SECTION 2: LOAD AND VISUALIZE DEG SUMMARY
# ==============================================================================

# Descriptive Information

```{r}
# ==============================================================================
# Load DESeq2 Summary Statistics
# ==============================================================================

# Load pseudobulk DEG summary file from DESeq2 analysis
# Contains: cluster identities, DEG counts at various thresholds, average Atf3 expression
Atf3_DESeq2 <- read.csv("~/cmo_chapter_v1/Analysis/3_dge/Atf3_summary_corrected",sep = "\t",dec = ",")

# Round average Atf3 expression to 2 decimal places
Atf3_DESeq2$avg_exp_Atf3 <- round(Atf3_DESeq2$avg_exp_Atf3,2)

# ==============================================================================
# Calculate Differential Expression Metrics
# ==============================================================================

# Use padj < 0.05 threshold for DEG counts
Atf3_DESeq2$DEG_counts <- Atf3_DESeq2$padj0_05

# DiffExprScore: Net difference between up and down regulated genes
# Positive = more upregulated, Negative = more downregulated
Atf3_DESeq2$DiffExprScore <- Atf3_DESeq2$uppadj0_05 - Atf3_DESeq2$downpadj0_05

# Normalize score to 0-1 range for color gradient
# Min-max normalization: (x - min) / (max - min)
Atf3_DESeq2$NormalizedScore <- (Atf3_DESeq2$DiffExprScore - min(Atf3_DESeq2$DiffExprScore)) / (max(Atf3_DESeq2$DiffExprScore) - min(Atf3_DESeq2$DiffExprScore))

# ==============================================================================
# Create Summary Visualization
# ==============================================================================

# Define divergent color scale for expression direction
# 0 = more downregulated (orange), 0.5 = balanced (white), 1 = more upregulated (green)
custom_scale <- scale_fill_gradient2(
  low = 'orange', 
  mid = 'white', 
  high = 'green', 
  midpoint = 0.50, 
  name = "Expression\nChange",
  breaks = c(0, 0.5, 1),
  labels = c("Down > Up", "Up=Down", "Up > Down")
)

# Bubble plot: x = Atf3 expression, y = DEG counts, size = DEG counts, fill = direction
plot_Atf3_deseq2 <- ggplot(Atf3_DESeq2, aes(x = avg_exp_Atf3, y = DEG_counts, label = ifelse(DEG_counts >=0, as.character(Idents), ''))) +
  geom_point(aes(fill = NormalizedScore, size = DEG_counts), shape = 21, color = "black", stroke = 0.5) +
  geom_label_repel(label.size = NA, fill = NA, size = 5) +  # Add cluster labels
  theme_classic() +
  ggtitle("Atf3KO vs Control; DESeq2 DGE Results padj < 0.05") +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "right"
  ) +
  custom_scale +
  scale_size_continuous(range = c(2, 10), name = "DEG Counts") +
  labs(x = "Average Expression", y = "DEG Counts") +
  annotate("text", x = Inf, y = -Inf, label = "Threshold: 0", hjust = 1.1, vjust = -0.5, size = 4, color = "darkgrey")
plot_Atf3_deseq2

# Export plot
ggsave("./plots/plot_Atf3_summary_1.pdf", width = 7, height = 5)
ggsave("./plots/plot_Atf3_summary_1.png", width = 7, height = 5)
```

```{r}
# ==============================================================================
# Plot: Included Genes vs DEG Counts
# ==============================================================================

# Visualize relationship between genes included in analysis and DEGs detected
# Point size = Atf3 expression level
plot_included <-ggplot(Atf3_DESeq2, aes(x = included_g, y = DEG_counts, label = ifelse(DEG_counts >= 8, as.character(Idents), ''))) +
  geom_point(aes(size = avg_exp_Atf3), shape = 21, color = "black", fill = "skyblue", stroke = 1) +
  geom_label_repel(label.size = NA, fill = NA, size = 5, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  theme_classic() +
  ggtitle("Atf3-KO: Included Genes and DEGs in DGE Analysis") +
  theme(
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18, color = "darkblue"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  labs(
    x = "Included Genes",
    y = "DEG Counts",
    size =  "Average\nExpression\nof Atf3"
  ) +
  scale_size_continuous(range = c(1, 8))

plot_included
ggsave("./plots/plot_summary_2.pdf", width = 7, height = 5)
ggsave("./plots/plot_summary_2.png", width = 7, height = 5)
```

# ==============================================================================
# SECTION 3: COMPILE DEG LIST FROM ALL CLUSTERS
# ==============================================================================

# DEG List

## padjo05

```{r}
# ==============================================================================
# Load All DESeq2 Results Files
# ==============================================================================

# Set directory containing DESeq2 results
setwd("~/cmo_chapter_v1/Analysis/3_dge/DESeq2_results/Combined_Atf3/")

# list.files(): Get all files matching pattern
# Pattern: *_resOrdered_enhanced.txt (DESeq2 output format)
files <- list.files(pattern = "_resOrdered_enhanced.txt")

# Initialize empty list for data frames
df_list <- list()

# ==============================================================================
# Loop Through Each Cluster's Results
# ==============================================================================

for (file in files) {
  # Extract cluster name from filename
  # sub(): Replace pattern with empty string
  sample_name <- sub("_resOrdered_enhanced.txt", "", file)
  
  # Read DESeq2 results
  df <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  
  # Filter for significant DEGs (padj < 0.05)
  filtered_df <- df %>% filter(padj < 0.05)
  
  # Only process if DEGs found
  if (nrow(filtered_df) > 0) {
    # Add cluster name column
    filtered_df <- filtered_df %>% mutate(Samplename = sample_name)
    
    # Append to list
    df_list[[length(df_list) + 1]] <- filtered_df
  }
}

# ==============================================================================
# Combine All Cluster Results
# ==============================================================================

# bind_rows(): Stack all data frames vertically
combined_df <- bind_rows(df_list)

head(combined_df)
```

```{r}
# Check dimensions of combined data
dim(combined_df)
```

# ==============================================================================
# SECTION 4: CONSOLIDATE GENE-LEVEL STATISTICS
# ==============================================================================

```{r}
# ==============================================================================
# Aggregate Statistics Per Gene Across All Clusters
# ==============================================================================

# For genes appearing in multiple clusters, consolidate statistics
cleaned_df <- combined_df %>%
  group_by(gene) %>%
  summarize(
    gene_id = first(gene_id),                          # Keep first Ensembl ID
    baseMean = max(baseMean, na.rm = TRUE),            # Maximum expression
    pvalue = min(pvalue, na.rm = TRUE),                # Minimum p-value (most significant)
    padj = min(padj, na.rm = TRUE),                    # Minimum adjusted p-value
    Samplename = paste(Samplename, collapse = ";"),    # Concatenate all cluster names
    # log2FoldChange: Keep most extreme value (max if any positive, min if all negative)
    log2FoldChange = if (all(log2FoldChange < 0)) min(log2FoldChange, na.rm = TRUE) else max(log2FoldChange, na.rm = TRUE),
    # stat: Keep most extreme Wald statistic
    stat = if (all(stat < 0)) min(stat, na.rm = TRUE) else max(stat, na.rm = TRUE),
    # updown: Classify as "up", "down", or "mix" based on direction consistency
    updown = case_when(
      all(log2FoldChange < 0) ~ "down",    # All clusters show downregulation
      all(log2FoldChange > 0) ~ "up",      # All clusters show upregulation
      TRUE ~ "mix"                          # Mixed direction across clusters
    )
  )
head(cleaned_df)
```

```{r}
# Check consolidated gene count
dim(cleaned_df)
```

```{r}
# Store as working copy
Atf3_degs_0 <- cleaned_df
```

```{r}
# Count genes by direction
# "up" = upregulated in KO, "down" = downregulated in KO, "mix" = inconsistent
table(cleaned_df$updown)
```

# ==============================================================================
# SECTION 5: ENCODE TF TARGET GENE INTEGRATION
# ==============================================================================

# ENCODE TARGET GENE

ATF3 Human 8397 genes (<https://maayanlab.cloud/Harmonizome/dataset/ENCODE+Transcription+Factor+Targets>) Last Updated
2015 Apr 06)

conversion <https://www.syngoportal.org/convert> (version 12-12-2020) (to mouse gene names)

```{r}
# ==============================================================================
# Load ENCODE ATF3 Target Genes
# ==============================================================================

# Human ATF3 targets from ENCODE converted to mouse orthologs via SynGO
encode_target_symbol <- read.delim2("../0_TF_Target_encode/SynGO_id_convert_2024-06-04 14;37/target_symbols.txt",header = F)
```

```{r}
# ==============================================================================
# Annotate DEGs as ENCODE Targets
# ==============================================================================

# Add boolean column: TRUE if gene is an ENCODE ATF3 target
# %in%: Set membership test
Atf3_degs_4 <- Atf3_degs_3 %>%
  mutate(encode_target = gene %in% encode_target_symbol$V1)
```

```{r}
# Count DEGs that are ENCODE targets
table(Atf3_degs_4$encode_target)
```

# ==============================================================================
# SECTION 6: LITERATURE GENE SET INTEGRATION
# ==============================================================================

# Literature

```{r}
# ==============================================================================
# Load Literature-Curated Gene Sets
# ==============================================================================

# Load pre-compiled gene sets from published ATF3 studies
# Contains: wang23up, wang23down, sood17deg, sykes21up, sykes21down,
#           sammons23up, sammons23down, nardo15up, nardo15down
litera <- readRDS("../00_Literature/all_merge_cleaned_lit.rds")
str(litera)
```

```{r}
# Store working copy
Atf3_degs_5 <- Atf3_degs_4
```

```{r}
# ==============================================================================
# Add Literature Gene Set Annotations
# ==============================================================================

# For each published study, add boolean columns indicating membership
# wang23: Wang et al. 2023
Atf3_degs_5$wang23up <- Atf3_degs_5$gene %in% litera$wang23up
Atf3_degs_5$wang23down <- Atf3_degs_5$gene %in% litera$wang23down

# sood17: Sood et al. 2017
Atf3_degs_5$sood17deg <- Atf3_degs_5$gene %in% litera$sood17deg

# sykes21: Sykes et al. 2021
Atf3_degs_5$sykes21up <- Atf3_degs_5$gene %in% litera$sykes21up
Atf3_degs_5$sykes21down <- Atf3_degs_5$gene %in% litera$sykes21down

# sammons23: Sammons et al. 2023
Atf3_degs_5$sammons23up <- Atf3_degs_5$gene %in% litera$sammons23up
Atf3_degs_5$sammons23down <- Atf3_degs_5$gene %in% litera$sammons23down

# nardo15: Nardo et al. 2015
Atf3_degs_5$nardo15up <- Atf3_degs_5$gene %in% litera$nardo15up
Atf3_degs_5$nardo15down <- Atf3_degs_5$gene %in% litera$nardo15down

# Combined sets (up + down)
Atf3_degs_5$wang23 <- Atf3_degs_5$gene %in% litera$wang23
Atf3_degs_5$sykes21 <- Atf3_degs_5$gene %in% litera$sykes21
Atf3_degs_5$sammons23 <- Atf3_degs_5$gene %in% litera$sammons23
Atf3_degs_5$nardo15 <- Atf3_degs_5$gene %in% litera$nardo15

# Any literature evidence
Atf3_degs_5$liter_all <- Atf3_degs_5$gene %in% litera$all
```

```{r}
# Count DEGs with any literature support
table(Atf3_degs_5$liter_all)
```

```{r}
# Store working copy
Atf3_degs_6 <-Atf3_degs_5
```

# ==============================================================================
# SECTION 7: CHIPSEQ DATA INTEGRATION
# ==============================================================================

# ChIPseq DATA

```{r}
# ==============================================================================
# Load ChIP-seq Peak-to-Gene Assignments
# ==============================================================================

# GREAT: Genomic Regions Enrichment of Annotations Tool
# Assigns regulatory peaks to nearest genes
data <- read.delim2("../../3_Chipseqs/5_Nearest_Gene/GREAT/20240709-public-4.0.4-qbtw0X-mm10-all-gene.txt")
head(data)

# Keep only gene ID and peak distance columns
data <- data[, 1:2]
colnames(data) <- c("gene_id", "peaks")
dim(data)
head(data)
tail(data)
```

```{r}
# ==============================================================================
# Clean Peak Distance Values
# ==============================================================================

# Extract numeric values from parentheses and clean format
# str_extract_all(): Extract all matches of pattern
# Pattern: \\(([^)]+)\\) matches content inside parentheses
data <- data %>%
  mutate(peaks = str_extract_all(peaks, "\\(([^)]+)\\)") %>% 
                 sapply(paste, collapse = ",") %>% 
                 str_replace_all("[()]", ""))

head(data)
```

```{r}
dim(data)
```

```{r}
# ==============================================================================
# Find Closest Peaks to TSS
# ==============================================================================

# Function to find closest peaks in each direction from TSS
# Positive values: downstream of TSS
# Negative values: upstream of TSS
find_closest_values_peaks <- function(peaks) {
  # Split comma-separated values and convert to numeric
  values <- as.numeric(unlist(strsplit(peaks, ",")))
  positive_values <- values[values > 0]
  non_positive_values <- values[values <= 0]
  
  # Find closest positive (smallest positive) and closest negative (largest non-positive)
  closest_positive <- if (length(positive_values) > 0) min(positive_values) else NA
  closest_non_positive <- if (length(non_positive_values) > 0) max(non_positive_values) else NA
  
  return(c(closest_positive, closest_non_positive))
}

# Apply function to each row
# rowwise(): Enable row-by-row operations
data <- data %>%
  rowwise() %>%
  mutate(
    closest_positive = find_closest_values_peaks(peaks)[1],
    closest_non_positive = find_closest_values_peaks(peaks)[2]
  )


head(data)
```

```{r}
tail(data)
```

```{r}
# ==============================================================================
# Count Total Peaks Per Gene
# ==============================================================================

# Function to count comma-separated values
count_values <- function(x) {
  lengths(strsplit(x, ","))
}

# Add column with total peak count per gene
data <- data %>%
  mutate(chip_all = sapply(peaks, count_values))

head(data)
```

```{r}
# Maximum peaks per gene
max(data$chip_all)
```

```{r}
# ==============================================================================
# Visualize Peak Distribution Per Gene
# ==============================================================================

# Histogram of peak counts per gene
ggplot(data, aes(x = chip_all)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Counts of Values in Peaks Column",
       x = "Count of Values",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# ==============================================================================
# Identify Genes with Most Peaks
# ==============================================================================

# Top 40 genes by peak count
top_40 <- data %>%
  arrange(desc(chip_all)) %>%
  head(40)

print(top_40[, -2], n=40)
```

```{r}
# ==============================================================================
# Calculate Peak Size Statistics
# ==============================================================================

# Load intersection file (ChIP-seq + ATAC-seq + Motif)
size_calculate_intersect_01_intersect_motif <- read.delim2("../../3_Chipseqs/4_intersections/2_with_motifs/ChIPseq_Atac_Motif.bed", header = F, sep = '\t')
head(size_calculate_intersect_01_intersect_motif)

# Calculate peak sizes (end - start)
size_calculate_intersect_01_intersect_motif$size <- size_calculate_intersect_01_intersect_motif$V3- size_calculate_intersect_01_intersect_motif$V2
head(size_calculate_intersect_01_intersect_motif)

# Summary statistics
mean(size_calculate_intersect_01_intersect_motif$size)
median(size_calculate_intersect_01_intersect_motif$size)
min(size_calculate_intersect_01_intersect_motif$size)
max(size_calculate_intersect_01_intersect_motif$size)
```

We can continue know. It is 500 bp fragment size.

```{r}
# ==============================================================================
# Extract All Peak Distances for Distribution Plot
# ==============================================================================

# Combine all peak distances into single vector
all_peaks <- data %>%
  pull(peaks) %>%               # Extract peaks column
  strsplit(",") %>%             # Split by comma
  unlist() %>%                  # Flatten to vector
  as.numeric()                  # Convert to numeric

# Remove NA values
all_peaks <- na.omit(all_peaks)

# Histogram of all peak distances to TSS
p1 <- ggplot(data.frame(peaks = all_peaks), aes(x = peaks)) +
  geom_histogram(binwidth = 10, fill = "black", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Chip Peak Distance to TSS",
       x = "TSS",
       y = "Count") +
  theme_classic()

p1
```

```{r}
ggsave("./plots/chip_peaks_all.pdf",width = 7, height = 5)
ggsave("./plots/chip_peaks_all.png",width = 7, height = 5)
```

```{r}
# ==============================================================================
# Zoom: ±100kb from TSS
# ==============================================================================

p2 <- ggplot(data.frame(peaks = all_peaks), aes(x = peaks)) +
  geom_histogram(binwidth = 10, fill = "black", color = "black", alpha = 0.7) +
  labs(title = "Histogram of ChipSeq Peak Distance to TSS (-100kb to +100kb)",
       x = "TSS",
       y = "Count") +
  theme_classic()+
  xlim(-100000, 100000) 
p2
```

```{r}
ggsave("./plots/chip_peaks_100kb.pdf",width = 7, height = 5)
ggsave("./plots/chip_peaks_100kb.png",width = 7, height = 5)
```

```{r}
# ==============================================================================
# Zoom: ±10kb from TSS
# ==============================================================================

ggplot(data.frame(peaks = all_peaks), aes(x = peaks)) +
  geom_histogram(binwidth = 1, fill = "black", color = "black", alpha = 0.7) +
  labs(title = "Histogram of ChipSeq Peak Distance to TSS (-10kb to +10kb)",
       x = "TSS",
       y = "Count") +
  theme_classic()+
  scale_x_continuous(limits = c(-10000, 10000), breaks = seq(-10000, 10000, by = 1000))
```

```{r}
ggsave("./plots/chip_peaks_10kb.pdf",width = 7, height = 5)
ggsave("./plots/chip_peaks_10kb.png",width = 7, height = 5)
```

# ==============================================================================
# SECTION 8: PUBLICATION FIGURE - TSS DISTANCE
# ==============================================================================

# Figure

```{r,fig.width=5, fig.height=3.5}
# ==============================================================================
# Publication-Quality TSS Distance Histogram
# ==============================================================================

# Focus on ±1kb from TSS (promoter region)
figure_atf3_tss_0 <-
  ggplot(data.frame(peaks = all_peaks), aes(x = peaks)) +
  geom_histogram(binwidth = 1, fill = "#2166ac", color = "#2166ac", alpha = 1) +
  labs( x = "TSS",
       y = "Count") +
  theme_classic()+  theme(
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    axis.text = element_text(size = 7),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  )+
  scale_x_continuous(limits = c(-1000, 1000), breaks = seq(-1000, 1000, by = 250))+
  geom_vline(xintercept = 0, linetype=2, color="red",linewidth=0.2, alpha=0.7)  # Mark TSS

figure_atf3_tss_0
ggsave("./plots/figure_atf3_tss_0.png", width = 5, height = 3.5, units = "in", dpi = 600)
ggsave("./plots/figure_atf3_tss_0.svg", width = 5, height = 3.5, units = "in", dpi = 600)
ggsave("./plots/figure_atf3_tss_0.pdf", width = 5, height = 3.5, units = "in", dpi = 600)
```

# ==============================================================================
# SECTION 9: UNIQUE CLUSTER MARKER ANALYSIS
# ==============================================================================

# Unique clusters

# Intersection Unique clusters

## 0.05 (o05)

```{r}
# ==============================================================================
# Define Gene Sets from Small Clusters
# ==============================================================================

# Extract significant genes (padj < 0.05) from each small cluster
# These clusters had insufficient cells for DESeq2 pseudobulk, used FindMarkers instead
Atf3_unique_o05 <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndCyp26b1     = rownames(EndCyp26b1_cmo_CvsN_3[EndCyp26b1_cmo_CvsN_3$p_val_adj< 0.05,])
)

upset_df <- Atf3_unique_o05

# ==============================================================================
# Create UpSet Plot
# ==============================================================================

# fromList(): Convert list of gene sets to binary matrix for UpSetR
upset_data <- fromList(upset_df)


# upset(): Create UpSet plot showing set intersections
# More scalable alternative to Venn diagrams for >3 sets
plot_Atf3_unique_o05 <-
  upset(
  upset_data, 
  keep.order = TRUE,            # Preserve set order as specified
  nsets = length(upset_df),     # Include all sets
  order.by = "degree",          # Order by intersection size
  sets.bar.color = "blue",      # Set size bar color
  matrix.color = "gray23",      # Matrix dot color
  main.bar.color ="red",        # Intersection bar color
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),        # Matrix to bar ratio
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o05
```

# ==============================================================================
# SECTION 10: INTERSECTION ANALYSIS - DIFFERENT THRESHOLDS
# ==============================================================================

## o05 baIfit1 - EndRsad2

```{r}
# ==============================================================================
# Two-Set Intersection: baIfit1 vs EndRsad2 (padj < 0.05)
# ==============================================================================

Atf3_unique_o05_be <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,])
)

upset_df <- Atf3_unique_o05_be
upset_data <- fromList(upset_df)


# UpSet plot for two-set comparison
plot_Atf3_unique_o05_be <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o05_be
```


```{r}
# ==============================================================================
# Two-Set Intersection: padj < 0.01 (More Stringent)
# ==============================================================================

Atf3_unique_o01_be <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.01,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.01,])
)

upset_df <- Atf3_unique_o01_be
upset_data <- fromList(upset_df)


plot_Atf3_unique_o01_be <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o01_be
```

```{r}
# ==============================================================================
# Two-Set Intersection: padj < 0.001 (Most Stringent)
# ==============================================================================

Atf3_unique_o001_be <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.001,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.001,])
)

upset_df <- Atf3_unique_o001_be
upset_data <- fromList(upset_df)


plot_Atf3_unique_o001_be <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o001_be
```

# ==============================================================================
# SECTION 11: INTEGRATE SMALL CLUSTERS WITH DESeq2 DEGs
# ==============================================================================

# I will use 0.05

```{r}
# ==============================================================================
# Four-Set Intersection: Three Small Clusters + DESeq2 DEGs
# ==============================================================================

Atf3_unique_o05_beX_deg <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndCyp26b1     = rownames(EndCyp26b1_cmo_CvsN_3[EndCyp26b1_cmo_CvsN_3$p_val_adj< 0.05,]),
  DEGs          = df$gene  # DESeq2 pseudobulk DEGs
)

upset_df <- Atf3_unique_o05_beX_deg
upset_data <- fromList(upset_df)


plot_Atf3_unique_o05_beX_deg <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o05_beX_deg
```

```{r}
# ==============================================================================
# Three-Set Intersection: Two Small Clusters + DESeq2 DEGs
# ==============================================================================

Atf3_unique_o05_be_deg <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  DEGs          = df$gene
)

upset_df <- Atf3_unique_o05_be_deg
upset_data <- fromList(upset_df)


plot_Atf3_unique_o05_be_deg <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o05_be_deg
```

# ==============================================================================
# SECTION 12: ANNOTATE DEGs WITH SMALL CLUSTER MEMBERSHIP
# ==============================================================================

### EndRsad2 - baIfit1 - DEGs

```{r}
# ==============================================================================
# Add Boolean Columns for Each Small Cluster
# ==============================================================================

# Check if each DEG is also significant in baIfit1
df$int_baIfit1 <- ifelse(df$gene %in% Atf3_unique_o05_be_deg$baIfit1, TRUE, FALSE)
```

```{r}
table(df$int_baIfit1)
```

```{r}
# Check if each DEG is also significant in EndRsad2
df$int_EndRsad2 <- ifelse(df$gene %in% Atf3_unique_o05_be_deg$EndRsad2, TRUE, FALSE)
table(df$int_EndRsad2)
```

```{r}
# Check if each DEG is also significant in EndCyp26b1
df$int_EndCyp26b1 <- ifelse(df$gene %in% Atf3_unique_o05_beX_deg$EndCyp26b1, TRUE, FALSE)
table(df$int_EndCyp26b1)
```

# ==============================================================================
# SECTION 13: CALCULATE SET INTERSECTIONS
# ==============================================================================

```{r}
# ==============================================================================
# Triple Intersection: All Three Sets
# ==============================================================================

# Reduce(intersect, ...): Sequentially apply intersect to list of vectors
# Genes significant in baIfit1 AND EndRsad2 AND DESeq2
genes51 <- Reduce(intersect, list(Atf3_unique_o05_be_deg$baIfit1,Atf3_unique_o05_be_deg$EndRsad2,Atf3_unique_o05_be_deg$DEGs))
genes51
```

```{r}
genes61 <-  Reduce(intersect, list(Atf3_unique_o05_be_deg$baIfit1,Atf3_unique_o05_be_deg$EndRsad2,Atf3_unique_o05_be_deg$DEGs))
genes51
```

```{r}
# Visualize with ggvenn
ggvenn(Atf3_unique_o05_be_deg)
```

```{r}
# ==============================================================================
# Define Outersect Function (Symmetric Difference)
# ==============================================================================

# outersect(): Find elements unique to one set or the other (XOR operation)
outersect <- function(x, y) {
  sort(c(setdiff(x, y),    # Elements in x but not y
         setdiff(y, x)))   # Elements in y but not x
}
```

```{r}
# ==============================================================================
# Find Genes in baIfit1 ∩ EndRsad2 but NOT in DEGs
# ==============================================================================

# Genes shared by small clusters but not detected in pseudobulk
genes_65 <- 
  outersect(
  intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$EndRsad2), 
  intersect(intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$EndRsad2),Atf3_unique_o05_be_deg$DEGs))
length(genes_65)
genes_65
```

```{r}
# ==============================================================================
# Find Genes in baIfit1 ∩ DEGs but NOT in EndRsad2 ∩ DEGs
# ==============================================================================

genes_13 <- 
  outersect(
  intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$DEGs), 
  intersect(intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$EndRsad2),Atf3_unique_o05_be_deg$DEGs))
length(genes_13)
genes_13
```

```{r}
# ==============================================================================
# Find Genes in EndRsad2 ∩ DEGs but NOT in baIfit1 ∩ DEGs
# ==============================================================================

genes_1 <- 
  outersect(
  intersect(Atf3_unique_o05_be_deg$EndRsad2, Atf3_unique_o05_be_deg$DEGs), 
  intersect(intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$EndRsad2),Atf3_unique_o05_be_deg$DEGs))
length(genes_1)
genes_1
```

```{r}
# ==============================================================================
# Triple Intersection: All Three Sets
# ==============================================================================

genes_51 <- 
  intersect(intersect(Atf3_unique_o05_be_deg$baIfit1, Atf3_unique_o05_be_deg$EndRsad2),Atf3_unique_o05_be_deg$DEGs)
length(genes_51)
genes_51
```

```{r}
# ==============================================================================
# Add Intersection Categories to Main Data Frame
# ==============================================================================

# Boolean columns for each intersection category
df$unq_bed_genes51 <- ifelse(df$gene %in% genes_51, TRUE, FALSE)  # All three sets
df$unq_bd_genes13 <- ifelse(df$gene %in% genes_13, TRUE, FALSE)   # baIfit1 + DEGs only
df$unq_ed_genes1 <- ifelse(df$gene %in% genes_1, TRUE, FALSE)     # EndRsad2 + DEGs only
```

# ==============================================================================
# SECTION 14: EXPORT ANNOTATED DEG TABLE
# ==============================================================================

```{r}
# ==============================================================================
# Save Final Annotated DEG Table
# ==============================================================================

# Save as RDS (preserves R object structure)
saveRDS(df, "./data_output/Atf3_degs_15.rds")

# Save as tab-separated text
write.table(df, "./data_output/Atf3_degs_15.txt",sep = '\t', row.names = F, col.names = T)
```

```{r}
# ==============================================================================
# Final UpSet Plot: Three-Set Comparison
# ==============================================================================

Atf3_unique_o05_be_deg <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  DEGs          = df$gene
)

upset_df <- Atf3_unique_o05_be_deg
upset_data <- fromList(upset_df)


plot_Atf3_unique_o05_be_deg <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3,
    
)
plot_Atf3_unique_o05_be_deg
```

```{r}
# Export UpSet plot
pdf("./plots/upsetr_bed.pdf",width = 6.5, height = 4.5)
plot_Atf3_unique_o05_be_deg
dev.off()

png(filename = "./plots/upsetr_bed.png",width = 6.5, height = 4.5,units = "in",res = 300)
plot_Atf3_unique_o05_be_deg
dev.off()
```

```{r}
# ==============================================================================
# Three Small Clusters UpSet Plot
# ==============================================================================

Atf3_unique_o05 <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndCyp26b1     = rownames(EndCyp26b1_cmo_CvsN_3[EndCyp26b1_cmo_CvsN_3$p_val_adj< 0.05,])
)

upset_df <- Atf3_unique_o05
upset_data <- fromList(upset_df)


plot_Atf3_unique_o05 <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)

plot_Atf3_unique_o05
```

```{r}
pdf("./plots/upsetr_bee.pdf",width = 6.5, height = 4.5)
plot_Atf3_unique_o05
dev.off()

png(filename = "./plots/upsetr_bee.png",width = 6.5, height = 4.5,units = "in",res = 300)
plot_Atf3_unique_o05
dev.off()
```

```{r}
# ==============================================================================
# Four-Set UpSet: All Small Clusters + DEGs
# ==============================================================================

Atf3_unique_o05_beX_deg <- list(
  baIfit1        = rownames(baIfit1_cmo_CvsN_3[baIfit1_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndRsad2       = rownames(EndRsad2_cmo_CvsN_3[EndRsad2_cmo_CvsN_3$p_val_adj< 0.05,]),
  EndCyp26b1     = rownames(EndCyp26b1_cmo_CvsN_3[EndCyp26b1_cmo_CvsN_3$p_val_adj< 0.05,]),
  DEGs          = df$gene
)

upset_df <- Atf3_unique_o05_beX_deg
upset_data <- fromList(upset_df)


plot_Atf3_unique_o05_beX_deg <-
  upset(
  upset_data, 
  keep.order = TRUE,
  nsets = length(upset_df),
  order.by = "degree",
  sets.bar.color = "blue",
  matrix.color = "gray23",
  main.bar.color ="red",
  att.color = c("red","green"),
  text.scale = 1.5,
  set_size.show = F,
  mb.ratio = c(0.7,0.3),
  matrix.dot.alpha = 0.1, line.size = 0.4, point.size = 3
    
)
plot_Atf3_unique_o05_beX_deg
```

```{r}
pdf("./plots/upsetr_beed.pdf",width = 6.5, height = 4.5)
plot_Atf3_unique_o05_beX_deg
dev.off()

png(filename = "./plots/upsetr_beed.png",width = 6.5, height = 4.5,units = "in",res = 300)
plot_Atf3_unique_o05_beX_deg
dev.off()
```

# ==============================================================================
# SECTION 15: VOLCANO PLOT PREPARATION
# ==============================================================================

# Volcano Plots

```{r}
# ==============================================================================
# Prepare Data Frames for Volcano Plots
# ==============================================================================

# Add gene names as column (currently in rownames)
EndRsad2_cmo_CvsN_3_df <- EndRsad2_cmo_CvsN_3
EndRsad2_cmo_CvsN_3_df$gene <- rownames(EndRsad2_cmo_CvsN_3)

EndCyp26b1_cmo_CvsN_3_df <- EndCyp26b1_cmo_CvsN_3
EndCyp26b1_cmo_CvsN_3_df$gene <- rownames(EndCyp26b1_cmo_CvsN_3)

baIfit1_cmo_CvsN_3_df <- baIfit1_cmo_CvsN_3
baIfit1_cmo_CvsN_3_df$gene <- rownames(baIfit1_cmo_CvsN_3)
```

# ==============================================================================
# SECTION 16: VOLCANO PLOT - baIfit1
# ==============================================================================

# baIfit1

```{r}
# ==============================================================================
# Volcano Plot: baIfit1 vs Late Mesoderm
# ==============================================================================

de <- baIfit1_cmo_CvsN_3_df

# Initialize differential expression category
de$diffexpressed <- "NO"

# Classify genes by direction and significance
de$diffexpressed[de$avg_log2FC > 0 & de$p_val_adj < 0.05] <- "UP"
de$diffexpressed[de$avg_log2FC < -0 & de$p_val_adj < 0.05] <- "DOWN"

# Add labels for significant genes only
de$delabel <- NA
de$delabel[de$diffexpressed != "NO"] <- de$gene[de$diffexpressed != "NO"]

# Remove Atf3 and Fam71a (confounders) for clean version
de_clean <- de %>% filter(!(gene %in% c("Atf3", "Fam71a")))

# Count significant genes
up_count <- sum(de$diffexpressed == "UP")
down_count <- sum(de$diffexpressed == "DOWN")


# Create volcano plot
plot_volcano_baIfit1 <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +  # DOWN=red, NO=black, UP=blue
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("baIfit1 vs Hoxa+ and Hoxb+ Mesoderm") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_baIfit1)

# ----------------------------------------------------------------------------------------------------
# Clean version (without Atf3 and Fam71a)
plot_volcano_baIfit1_clean <- ggplot(data=de_clean, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("baIfit1 vs Hoxa+ and Hoxb+ Mesoderm") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de_clean$avg_log2FC),y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de_clean$avg_log2FC), y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_baIfit1_clean)
```

```{r}
# ==============================================================================
# Alternative Color Scheme Volcano Plot
# ==============================================================================

plot_volcano_baIfit1x <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=5.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#B51D8D", "black", "#c85b00")) +  # Alternative color scheme
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("baIfit1 vs Late Mesoderm") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 15, color = "black",face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 9),
    legend.position = "none", 
    axis.line = element_line(size = 0.2),
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(padj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.05, vjust =-0, size = 5.5, color = "red", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 5, vjust = -0, size = 5.5, color = "blue", fontface = "italic")

print(plot_volcano_baIfit1x)
```

# ==============================================================================
# SECTION 17: VOLCANO PLOT - EndRsad2
# ==============================================================================

## EndRsad2

```{r}
# ==============================================================================
# Volcano Plot: EndRsad2 vs Other Endothelial Clusters
# ==============================================================================

de <- EndRsad2_cmo_CvsN_3_df

# Classify differential expression
de$diffexpressed <- "NO"
de$diffexpressed[de$avg_log2FC > 0 & de$p_val_adj < 0.05] <- "UP"
de$diffexpressed[de$avg_log2FC < -0 & de$p_val_adj < 0.05] <- "DOWN"

# Add gene labels
de$delabel <- NA
de$delabel[de$diffexpressed != "NO"] <- de$gene[de$diffexpressed != "NO"]

# Remove confounders
de_clean <- de %>% filter(!(gene %in% c("Atf3", "Fam71a")))

# Count significant genes
up_count <- sum(de$diffexpressed == "UP")
down_count <- sum(de$diffexpressed == "DOWN")


# Create volcano plot
plot_volcano_EndRsad2 <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndRsad2 vs EndGadd45g, EndCdh5, EndCyp26b1") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_EndRsad2)

# ----------------------------------------------------------------------------------------------------
# Clean version
plot_volcano_EndRsad2_clean <- ggplot(data=de_clean, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndRsad2 vs EndGadd45g, EndCdh5, EndCyp26b1") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de_clean$avg_log2FC),y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de_clean$avg_log2FC), y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_EndRsad2_clean)
```

```{r}
# Alternative color scheme
plot_volcano_EndRsad2x <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=5.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#551153", "black", "#c85b00")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndRsad2 vs Rest of the Endothelium") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 15, color = "black",face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 9),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(padj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.05, vjust =-0, size = 5.5, color = "red", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 2.5, vjust = -0, size = 5.5, color = "blue",  fontface = "italic")

print(plot_volcano_EndRsad2x)
```

# ==============================================================================
# SECTION 18: VOLCANO PLOT - EndCyp26b1
# ==============================================================================

## EndCyp26b1

```{r}
# ==============================================================================
# Volcano Plot: EndCyp26b1 vs Other Endothelial Clusters
# ==============================================================================

de <- EndCyp26b1_cmo_CvsN_3_df

# Classify differential expression
de$diffexpressed <- "NO"
de$diffexpressed[de$avg_log2FC > 0 & de$p_val_adj < 0.05] <- "UP"
de$diffexpressed[de$avg_log2FC < -0 & de$p_val_adj < 0.05] <- "DOWN"

# Add gene labels
de$delabel <- NA
de$delabel[de$diffexpressed != "NO"] <- de$gene[de$diffexpressed != "NO"]

# Remove confounders
de_clean <- de %>% filter(!(gene %in% c("Atf3", "Fam71a")))

# Count significant genes
up_count <- sum(de$diffexpressed == "UP")
down_count <- sum(de$diffexpressed == "DOWN")


# Create volcano plot
plot_volcano_EndCyp26b1 <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndCyp26b1 vs EndGadd45g, EndCdh5, EndRsad2") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_EndCyp26b1)

# ----------------------------------------------------------------------------------------------------
# Clean version
plot_volcano_EndCyp26b1_clean <- ggplot(data=de_clean, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3.5, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=3.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#b2182b", "black", "#2166ac")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndCyp26b1 vs EndGadd45g, EndCdh5, EndRsad2") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, color = "black",face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(p_val_adj)"
  )+
  annotate("text",x = min(de_clean$avg_log2FC),y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.1, vjust =-0.5, size = 3, color = "#525252", fontface = "italic") +
  annotate("text",x = max(de_clean$avg_log2FC), y = max(-log10(na.omit(de_clean$p_val_adj))), label = paste("UP:", up_count), hjust = 1.1, vjust = -0.5, size = 3, color = "#525252", fontface = "italic")

print(plot_volcano_EndCyp26b1_clean)

```

```{r}
# Alternative color scheme
plot_volcano_EndCyp26b1x <- ggplot(data=de, aes(x=avg_log2FC, y=-log10(p_val_adj), color=diffexpressed, label=delabel)) +
  geom_point(size=3, alpha=0.7, aes(shape=diffexpressed)) +
  theme_classic(base_size = 15) +
  geom_text_repel(size=5.5, max.overlaps = 20, box.padding = 0.35, point.padding = 0.3, segment.color = 'grey50') +
  scale_color_manual(values=c("#551153", "black", "#c85b00")) +
  scale_shape_manual(values=c(16, 17, 15)) +
  ggtitle("EndCyp26b1 vs Rest of the Endothelium") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 15, color = "black",face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 9),
    legend.position = "none", 
    axis.line = element_line(size = 0.2)
  ) +
  labs(
    x = "Average Log2 Fold Change",
    y = "-log10(padj)"
  )+
  annotate("text",x = min(de$avg_log2FC), y = max(-log10(na.omit(de$p_val_adj))), label = paste("DOWN:", down_count), hjust = 0.05, vjust =-0, size = 5.5, color = "red", fontface = "italic") +
  annotate("text",x = max(de$avg_log2FC),y = max(-log10(na.omit(de$p_val_adj))), label = paste("UP:", up_count), hjust = 3.75, vjust = -0, size = 5.5, color = "blue",  fontface = "italic")

print(plot_volcano_EndCyp26b1x)
```

# ==============================================================================
# SECTION 19: COMBINED VOLCANO PLOT FIGURE
# ==============================================================================

```{r,fig.width=20, fig.height=5}
# ==============================================================================
# Combine All Three Volcano Plots
# ==============================================================================

# plot_grid(): Combine plots horizontally
pl_volcano_0 <- plot_grid(plot_volcano_baIfit1,plot_volcano_EndRsad2,plot_volcano_EndCyp26b1, ncol=3)
pl_volcano_0
```

```{r}
# Export combined plot
pdf("./plots/volcano_0.pdf",width = 20, height = 5)
pl_volcano_0
dev.off()

png(filename = "./plots/colcano_0.png",width = 20, height = 5,units = "in",res = 300)
pl_volcano_0
dev.off()
```

```{r}
# High-resolution version
png(filename = "./plots/colcano_0_600.png",width = 20, height = 5,units = "in",res = 600)
pl_volcano_0
dev.off()
```

# ==============================================================================
# SECTION 20: SAVE MARKER DATA
# ==============================================================================

```{r}
# ==============================================================================
# Export Marker Results for Each Cluster
# ==============================================================================

# EndRsad2
saveRDS(EndRsad2_cmo_CvsN_3_df, "./data_output/EndRsad2_cmo_CvsN_3_df.rds")
write.table(EndRsad2_cmo_CvsN_3_df, "./data_output/EndRsad2_cmo_CvsN_3_df.txt", sep = '\t',row.names = F)

# EndCyp26b1
saveRDS(EndCyp26b1_cmo_CvsN_3_df, "./data_output/EndCyp26b1_cmo_CvsN_3_df.rds")
write.table(EndCyp26b1_cmo_CvsN_3_df, "./data_output/EndCyp26b1_cmo_CvsN_3_df.txt", sep = '\t',row.names = F)

# baIfit1
saveRDS(baIfit1_cmo_CvsN_3_df, "./data_output/baIfit1_cmo_CvsN_3_df.rds")
write.table(baIfit1_cmo_CvsN_3_df, "./data_output/baIfit1_cmo_CvsN_3_df.txt", sep = '\t',row.names = F)
```

```{r}
# ==============================================================================
# Clean Up Large Objects
# ==============================================================================

rm(s_cmo)
rm(s_cmo_baIfit1)
rm(s_koatf3)
rm(s_ntatf3)
rm(s_ntatf3_baIfit1)
```

```{r}
# ==============================================================================
# Export Intersection Gene Lists
# ==============================================================================

write.table(genes_65,"./data_output/genes_65.txt",row.names = F,col.names = F, quote = F)  # baIfit1 ∩ EndRsad2 only
write.table(genes_1,"./data_output/genes_1.txt",row.names = F,col.names = F, quote = F)    # EndRsad2 ∩ DEGs only
write.table(genes_13,"./data_output/genes_13.txt",row.names = F,col.names = F, quote = F)  # baIfit1 ∩ DEGs only
write.table(genes_51,"./data_output/genes_51.txt",row.names = F,col.names = F, quote = F)  # Triple intersection
```

```{r}
# ==============================================================================
# Save Complete Workspace
# ==============================================================================

save.image("./data_output/rdata.rdata")
```

# ==============================================================================
# SECTION 21: SESSION INFORMATION
# ==============================================================================

```{r}
sessionInfo()
```

# ==============================================================================
# SECTION 22: ALTERNATIVE VOLCANO PLOT LAYOUT
# ==============================================================================

# asdasd

```{r,fig.width=10, fig.height=20}
# ==============================================================================
# Vertical Volcano Plot Arrangement
# ==============================================================================

pl_volcano_0x <- plot_grid(plot_volcano_baIfit1x,plot_volcano_EndRsad2x,plot_volcano_EndCyp26b1x, ncol=1)

```

```{r}
pl_volcano_0x
```

```{r}
# Export vertical layout
ggsave("./plots/volcano_x.png",width = 10, height = 20, units = "in",dpi = 600)
ggsave("./plots/volcano_x.svg",width = 10, height = 20, units = "in",dpi = 600)
ggsave("./plots/volcano_x.pdf",width = 10, height = 20, units = "in",dpi = 600)

# Save plot objects
saveRDS(pl_volcano_0x, "./data_output/pl_volcano_0x.rds")
saveRDS(plot_volcano_baIfit1x, "./data_output/plot_volcano_baIfit1x.rds")
saveRDS(plot_volcano_EndRsad2x, "./data_output/plot_volcano_EndRsad2x.rds")
saveRDS(plot_volcano_EndCyp26b1x, "./data_output/plot_volcano_EndCyp26b1x.rds")


```

```{r}
sessionInfo()
```

# ==============================================================================
# SECTION 23: ADDITIONAL FIGURE DISPLAY
# ==============================================================================

# Figure more

```{r}
# Display additional UpSet plots
plot_Atf3_unique_o05


plot_Atf3_unique_o05_be_deg
```




