---
title: "Differential Abundance Analysis with Speckle"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

DISCLAMIR:
I am summarizing my code here, so there might be typos or copy paste errors. This is more like a detailed index page.

# ==============================================================================
# SECTION 1: COLOR PALETTES FOR VISUALIZATION
# ==============================================================================

```{r eval=FALSE,}
# ==============================================================================
# Define Custom Color Palettes
# ==============================================================================
# r3dcol: A list containing multiple color palette options for differential
# abundance visualization across conditions and cell types

# Initialize empty list to store all color palettes
r3dcol <- list()

# cols_94: Large palette with 94 distinct colors
# Useful for visualizing many subclusters or cell types
r3dcol$cols_94 <- c(
  "#F0E442","#1CE6FF","#FF34FF","#FF4A46","#008941","#006FA6","#A30059","#FFDBE5","#7A4900","#0089A3",
  "#63FFAC","#B79762",
  "#004D43","#8FB0FF","#997D87","#809693","#6A3A4C","#b4b701","#4FC601","#3B5DFF","#FF2F80","#61615A",
  "#BA0900","#6B7900",
  "#00C2A0","#FFAA92","#FF90C9","#B903AA","#D16100","#922329","#7B4F4B","#A1C299","#0AA6D8","#00846F",
  "#FFB500","#C2FFED",
  "#A079BF","#CC0744","#C0B9B2","#C2FF99","#00489C","#6F0062","#0CBD66","#EEC3FF","#456D75","#B77B68",
  "#7A87A1","#788D66",
  "#885578","#FAD09F","#FF8A9A","#D157A0","#BEC459","#456648","#0086ED","#886F4C","#34362D","#B4A8BD",
  "#00A6AA","#452C2C",
  "#636375","#A3C8C9","#FF913F","#938A81","#575329","#00FECF","#B05B6F","#8CD0FF","#3B9700","#04F757",
  "#C8A1A1","#1E6E00",
  "#7900D7","#A77500","#6367A9","#A05837","#6B002C","#772600","#D790FF","#9B9700","#549E79","#FFF69F",
  "#201625","#72418F",
  "#BC23FF","#99ADC0","#3A2465","#DDEFFF","#5B4534","#FDE8DC","#404E55","#CB7E98","#A4E804","#324E72"
)

# cols_46: Medium palette with 46 colors for moderate cluster numbers
r3dcol$cols_46 <- c(
  "#FFB500","#00C2A0","#D157A0","#8CD0FF","#FF4A46","#FDE8DC","#63FFAC","#B79762","#8FB0FF","#997D87",
  "#809693","#b4b701",
  "#4FC601","#FFAA92","#FF90C9","#A1C299","#F0E442","#C2FFED","#A079BF","#C0B9B2","#C2FF99","#EEC3FF",
  "#B77B68","#7A87A1",
  "#788D66","#FAD09F","#FF8A9A","#BEC459","#B4A8BD","#A3C8C9","#FF913F","#938A81","#00FECF","#FF34FF",
  "#1CE6FF","#04F757",
  "#C8A1A1","#D790FF","#9B9700","#549E79","#FFF69F","#99ADC0","#DDEFFF","#FFDBE5","#CB7E98","#A4E804"
)

# cols_pjsala: Palette inspired by Pijuan-Sala et al. mouse gastrulation atlas
r3dcol$cols_pjsala <- c(
  "#005579","#0F4A9C","#139992","#1A1A1A","#354E23","#3F84AA","#532C8A","#635547","#647a4f","#65A83E",
  "#7F6874","#8870ad",
  "#8DB5CE","#8EC792","#989898","#9e6762","#B51D8D","#c19f70","#C3C388","#C594BF","#C72228","#c9a997",
  "#C9EBFB","#cc7818",
  "#CDE088","#DABE99","#DFCDE4","#EF4E22","#EF5A9D","#F397C0","#F6BFCB","#f79083","#f7f79e","#f9decf",
  "#FACB12","#FBBE92",
  "#ff891c"
)

# cols_4x3: 12 colors arranged as 4 color families x 3 shades
r3dcol$cols_4x3 <- c(
  "#551153","#970098","#9c24cd","#412c00","#997600","#c6a000","#1e3604","#355e07","#4f890e","#002f64",
  "#0050ae","#1077f3"
)

# cols_12x1: 12 distinct colors from ColorBrewer paired palette
r3dcol$cols_12x1 <- c(
  '#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a',
  '#ffff99','#b15928'
)

# cols_4x3b: Alternative 12-color palette
r3dcol$cols_4x3b <- c(
  "#551153","#970098","#cc34cd","#c85b00","#f98517","#fdbf6f","#00603d","#008c5c","#33b983","#002f64",
  "#0050ae","#1077f3"
)

# cols_4x1: 4 colors for condition comparisons (KO vs Control)
r3dcol$cols_4x1 <- c(
  "#970098","#997600","#4f890e","#0050ae"
)

# cols_4x1b: Alternative 4-color palette
r3dcol$cols_4x1b <- c(
  "#970098","#f98517","#008c5c","#1077f3"
)

# cols_3a: 3 colors for three-group comparisons
r3dcol$cols_3a <- c(
  "#D157A0","#FFB500","#00C2A0"
)

# cols_3b: Alternative 3-color palette
r3dcol$cols_3b <- c(
  "#D454B5","#c85b00","#008c5c"
)
```

# ==============================================================================
# SECTION 2: ENVIRONMENT SETUP AND PACKAGE LOADING
# ==============================================================================

```{r eval=FALSE}
# ==============================================================================
# Load Required Packages
# ==============================================================================

# Suppress startup messages for cleaner output
suppressPackageStartupMessages({
  library(SingleCellExperiment)  # Bioconductor container for single-cell data
  library(miloR)                 # Differential abundance with neighborhoods (alternative method)
  library(dplyr)                 # Data manipulation
  library(scater)                # Single-cell analysis utilities
  library(speckle)               # Main package for differential abundance analysis
  library(ggplot2)               # Visualization
})
```

# Introduction

Here we will summarize the steps of the Compositional Analysis / Differential Abundance Analysis between KO clones and
NT-Control (NonTargetting Control). We will do it in two different analysis 1) cluster levels with *Speckle* and 2)
k-nearstneighborhood with smaller groups of cells.

# ==============================================================================
# SECTION 3: DATA LOADING AND PREPARATION
# ==============================================================================

# Data Preprocessing

## Loading Data


```{r eval=FALSE}
# ==============================================================================
# Load Main Seurat Object and Create KO-Specific Subsets
# ==============================================================================

# Load the preprocessed Seurat object
cmo <- readRDS("~/cmo_chapter/introduction/rds/cmoy2.RDS")

# Set active identity to condition levels (KO vs Control)
Idents(cmo) <- "Conditions_lv"

# ==============================================================================
# Create Condition-Specific Subsets for Pairwise Comparisons
# ==============================================================================

# subset(): Extract cells belonging to specific conditions
# Each comparison is KO vs Control (not all conditions together)

# Atf3 knockout vs Control comparison
cmo_CA <- subset(cmo, idents = c("Control","Atf3KO"))

# Bcl6b knockout vs Control comparison  
cmo_BA <- subset(cmo, idents = c("Control","Bcl6bKO"))

# Zfp711 knockout vs Control comparison
cmo_ZA <- subset(cmo, idents = c("Control","Zfp711KO"))

# ==============================================================================
# Clean Factor Levels After Subsetting
# ==============================================================================

# After subsetting, factor columns retain all original levels (even unused ones)
# This can cause issues with speckle/statistical tests
# droplevels() removes unused factor levels

# Define helper function to clean metadata factors
clean_metadata <- function(seurat_obj) {
  metadata <- seurat_obj@meta.data
  # Apply droplevels to each factor column
  # lapply: Apply function to each column
  metadata[] <- lapply(metadata, function(column) {
    if (is.factor(column)) {
      return(droplevels(column))  # Remove unused levels
    } else {
      return(column)  # Keep non-factor columns unchanged
    }
  })
  seurat_obj@meta.data <- metadata
  return(seurat_obj)
}

# Apply cleaning to each subset
cmo_CA <- clean_metadata(cmo_CA)
cmo_BA <- clean_metadata(cmo_BA)
cmo_ZA <- clean_metadata(cmo_ZA)
```

I will show only one condition (Atf3-KO versus NT-Control).
We will first follow the classical "speckle" pipeline during the analysis. Later I will subset the samples to equal cell numbers and use the plotCellTypeProps function to compare KO and NT conditions at annotation level. 

# ==============================================================================
# SECTION 4: ATF3 KNOCKOUT ANALYSIS
# ==============================================================================

# Atf3

## Loading the RDS file

```{r}
# ==============================================================================
# Load Atf3 vs Control Subset
# ==============================================================================

# Load the pre-saved subset containing only Atf3KO and Control cells
cmo_CA <- readRDS("../../../RDS/CMO/cmo_CA.RDS")
```

## just to be sure
if you are subsetting the data and loosing some of the levels use droplevels() before running speckle and miloR
e.g.

```{r}
# ==============================================================================
# Ensure Clean Factor Levels
# ==============================================================================

# droplevels(): Critical step before DAA
# Removes factor levels that have 0 cells after subsetting
# Prevents errors in statistical modeling
cmo_CA$Conditions_lv <- droplevels(cmo_CA$Conditions_lv)
```

# ==============================================================================
# SECTION 5: EXTRACT ANNOTATION LEVEL FACTORS
# ==============================================================================

```{r}
# ==============================================================================
# Store Factor Levels for Consistent Ordering
# ==============================================================================

# levels(): Extract the ordered factor levels for each annotation
# These are used to ensure consistent ordering in plots and results

# cl_4: Subcluster + cell cycle combined annotation (finest resolution)
c_ca_4 <- levels(cmo_CA$cl_4)

# cl_3: Subcluster annotation (high resolution)
c_ca_3 <- levels(cmo_CA$cl_3)

# cl_2: Cluster annotation (medium resolution)
c_ca_2 <- levels(cmo_CA$cl_2)

# cl_1: Group annotation (coarsest resolution - e.g., Early, Mesoderm, Hematopoietic)
c_ca_1 <- levels(cmo_CA$cl_1)

# Phasel: Cell cycle phase annotation (G1, S, G2M)
c_ca_cc <- levels(cmo_CA$Phasel)

# Conditions_lv: Experimental conditions (Atf3KO, Control)
c_ca_conditions <- levels(cmo_CA$Conditions_lv)

# Samples_lv: Biological replicates within each condition
c_ca_samples <- levels(cmo_CA$Samples_lv)
```

```{r}
# ==============================================================================
# Set Active Identity for Analysis
# ==============================================================================

# Set identity to subcluster level for downstream analysis
Idents(cmo_CA) <- "cl_3"
```

# ==============================================================================
# SECTION 6: PROPELLER DIFFERENTIAL ABUNDANCE TESTING
# ==============================================================================

# Propeller

we will do DAA the test now in 5 levels (cl_1, ...) in both "asin" and "logit" tranformation. You can find the results in the supplementary table.


### Asin&logit



```{r}
# ==============================================================================
# Run Propeller DAA at All Annotation Levels
# ==============================================================================

# propeller(): Main function from speckle package for differential abundance
# Statistical test for changes in cell type proportions between conditions

# Parameters:
# - clusters: Cell type/cluster assignments for each cell
# - sample: Sample/replicate identifier for each cell
# - group: Condition/group identifier (what we're comparing)
# - transform: Variance stabilizing transformation ("asin" or "logit")

# "asin": Arcsine square root transformation - better for proportions near 0 or 1
# "logit": Logit transformation - log(p/(1-p)) - better for mid-range proportions

# Initialize list to store all results
prop_CA <- list()

# --- Arcsine Transformation Results ---
# S1a: Group level (coarsest)
prop_CA$S1a <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "asin")

# S2a: Cluster level
prop_CA$S2a <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "asin")

# S3a: Subcluster level
prop_CA$S3a <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "asin")

# S4a: Subcluster + cell cycle level (finest)
prop_CA$S4a <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "asin")

# S5a: Cell cycle phase level
prop_CA$S5a <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "asin")

# --- Logit Transformation Results ---
# Same analyses with logit transformation for comparison
prop_CA$S1l <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "logit")

prop_CA$S2l <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "logit")

prop_CA$S3l <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "logit")

prop_CA$S4l <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "logit")

prop_CA$S5l <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, 
                         group = cmo_CA$Conditions_lv, transform = "logit")

```

# ==============================================================================
# SECTION 7: PROPELLER RESULTS - INDIVIDUAL VARIABLES
# ==============================================================================

```{r}
# ==============================================================================
# Run Propeller with Arcsine Transformation (Individual Storage)
# ==============================================================================

# Store results in individual variables for easier export
# Arcsine transformation version

# Group level (e.g., Early, Mesoderm, Hematopoietic)
ca1_asin <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, 
                      group = cmo_CA$Conditions_lv, transform = "asin")

# Cluster level
ca2_asin <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, 
                      group = cmo_CA$Conditions_lv, transform = "asin")

# Subcluster level
ca3_asin <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, 
                      group = cmo_CA$Conditions_lv, transform = "asin")

# Cell cycle level
cacc_asin <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, 
                       group = cmo_CA$Conditions_lv, transform = "asin")
```

```{r}
# ==============================================================================
# Run Propeller with Logit Transformation (Individual Storage)
# ==============================================================================

# Logit transformation version of the same analyses
ca1_logit <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, 
                       group = cmo_CA$Conditions_lv, transform = "logit")

ca2_logit <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, 
                       group = cmo_CA$Conditions_lv, transform = "logit")

ca3_logit <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, 
                       group = cmo_CA$Conditions_lv, transform = "logit")

cacc_logit <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, 
                        group = cmo_CA$Conditions_lv, transform = "logit")
```

# ==============================================================================
# SECTION 8: EXPORT PROPELLER RESULTS
# ==============================================================================

```{r}
# ==============================================================================
# Export Arcsine Transformation Results
# ==============================================================================

# write.table(): Export results as tab-separated text files
# sep = '\t': Tab delimiter
# row.names = F: Don't include row numbers
# col.names = T: Include column headers
# quote = F: Don't quote strings

write.table(ca1_asin, "./Atf3/Atf3_propeller_groups_asin.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_asin, "./Atf3/Atf3_propeller_clusters_asin.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_asin, "./Atf3/Atf3_propeller_subclusters_asin.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_asin, "./Atf3/Atf3_propeller_cellcycle_asin.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
```

```{r}
# ==============================================================================
# Export Logit Transformation Results
# ==============================================================================

write.table(ca1_logit, "./Atf3/Atf3_propeller_groups_logit.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_logit, "./Atf3/Atf3_propeller_clusters_logit.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_logit, "./Atf3/Atf3_propeller_subclusters_logit.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_logit, "./Atf3/Atf3_propeller_cellcycle_logit.txt", 
            sep = '\t', row.names = F, col.names = T, quote = F)
```

# ==============================================================================
# SECTION 9: CELL TYPE PROPORTION VISUALIZATION - GROUP LEVEL
# ==============================================================================

```{r,fig.width=6.5, fig.height=5.5}
# ==============================================================================
# Stacked Bar Plot - Group Level Proportions
# ==============================================================================

# plotCellTypeProps(): From speckle, creates stacked bar plots of cell type proportions
# clusters: Cell type assignments
# sample: What to group bars by (here, conditions)

plot_cell_type_normal_1 <-
  plotCellTypeProps(clusters = cmo_CA$cl_1, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  # scale_x_discrete(): Control x-axis order
  scale_x_discrete(limits = c_ca_conditions) +
  # scale_fill_manual(): Custom colors for groups
  scale_fill_manual(
    "Groups", 
    values = c("#66c2a5","#fc8d62","#8da0cb"),
    guide = guide_legend(ncol = 1)
  ) + 
  # geom_vline(): Add vertical lines to separate replicates
  geom_vline(xintercept = c(3.5, 6.5, 9.5), linewidth = 0.2, color = "black")

print(plot_cell_type_normal_1)
```

# ==============================================================================
# SECTION 10: CELL TYPE PROPORTION VISUALIZATION - CLUSTER LEVEL
# ==============================================================================

```{r,fig.width=6.5, fig.height=5.5}
# ==============================================================================
# Stacked Bar Plot - Cluster Level Proportions
# ==============================================================================

plot_cell_type_normal_2 <-
  plotCellTypeProps(clusters = cmo_CA$cl_2, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Clusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 1)
  ) + 
  geom_vline(xintercept = c(3.5, 6.5, 9.5), linewidth = 0.2, color = "black")

print(plot_cell_type_normal_2)
```

# ==============================================================================
# SECTION 11: CELL TYPE PROPORTION VISUALIZATION - SUBCLUSTER LEVEL
# ==============================================================================

```{r,fig.width=6.5, fig.height=5.5}
# ==============================================================================
# Stacked Bar Plot - Subcluster Level Proportions
# ==============================================================================

plot_cell_type_normal_3 <-
  plotCellTypeProps(clusters = cmo_CA$cl_3, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Subclusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 2)  # Two columns for many subclusters
  ) + 
  geom_vline(xintercept = c(3.5, 6.5, 9.5), linewidth = 0.2, color = "black")

print(plot_cell_type_normal_3)
```

# ==============================================================================
# SECTION 12: CELL TYPE PROPORTION VISUALIZATION - FINEST RESOLUTION
# ==============================================================================

```{r,fig.width=14, fig.height=5.5}
# ==============================================================================
# Stacked Bar Plot - Subcluster + Cell Cycle Level
# ==============================================================================

# cl_4: Finest resolution combining subclusters with cell cycle phases
plot_cell_type_normal_4 <-
  plotCellTypeProps(clusters = cmo_CA$cl_4, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right",
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  # Combine two color palettes for the many categories
  scale_fill_manual(
    "subclusters-cc (A-Z)", 
    values = c(r3dcol$cols_94, r3dcol$cols_46),
    guide = guide_legend(ncol = 6)  # Six columns for very many categories
  ) + 
  geom_vline(xintercept = c(3.5, 6.5, 9.5), linewidth = 0.2, color = "black")

print(plot_cell_type_normal_4)
```

# ==============================================================================
# SECTION 13: MULTI-CONDITION COMPARISON SETUP
# ==============================================================================

# We will create one to one comparison plots

Doing some naming adjustment, (Naive and PGC has same naming in clusters and subclusters)


```{r}
# ==============================================================================
# Load Full Dataset for Multi-Condition Analysis
# ==============================================================================

# Load the complete Seurat object with all conditions
cmo <- readRDS("~/PhD_Projects/CMO/cmoy2_reup.RDS")
```

```{r}
# Display object summary
cmo
```

```{r}
# ==============================================================================
# Extract and Examine Subcluster Levels
# ==============================================================================

# Get current subcluster level names
levels(cmo$cl_3)
lvx <- levels(cmo$cl_3)
```

```{r}
# Set active identity to subclusters
Idents(cmo) <- "cl_3"
```


```{r}
# ==============================================================================
# Rename Overlapping Cluster Names
# ==============================================================================

# Some clusters have the same name at different annotation levels
# Rename to avoid confusion in merged results
# "Naive" -> "Naiv" and "PGC" -> "PGCs" at subcluster level

new.cluster.ids <- c("Naiv", lvx[2:6], "PGCs", lvx[-c(1:7)])
new.cluster.ids

```

```{r}
# ==============================================================================
# Apply New Names to Identities
# ==============================================================================

# names(): Assign names to the ID vector (maps old -> new)
names(new.cluster.ids) <- levels(cmo)

# RenameIdents(): Apply the renaming to the Seurat object
cmo <- RenameIdents(cmo, new.cluster.ids)
```

```{r}
# Verify the renaming worked
table(Idents(cmo))
```

```{r}
# Store renamed identities back to metadata
cmo$cl_3 <- Idents(cmo)
```

# ==============================================================================
# SECTION 14: MULTI-CONDITION DAA ANALYSIS
# ==============================================================================

```{r}
library(speckle)
```

```{r}
# Initialize results list for all conditions
daa <- list()
```


```{r}
# ==============================================================================
# Run Propeller DAA Across All Conditions Simultaneously
# ==============================================================================

# When group has >2 levels, propeller performs ANOVA-like test
# Tests if any condition differs from others

# --- Logit Transformation ---
daa$cmo_groups_logit <- propeller(clusters = cmo$cl_1, sample = cmo$Samples_lv, 
                                  group = cmo$Conditions_lv, transform = "logit")
daa$cmo_clusters_logit <- propeller(clusters = cmo$cl_2, sample = cmo$Samples_lv, 
                                    group = cmo$Conditions_lv, transform = "logit")
daa$cmo_subclusters_logit <- propeller(clusters = cmo$cl_3, sample = cmo$Samples_lv, 
                                       group = cmo$Conditions_lv, transform = "logit")
daa$cmo_cellcycle_logit <- propeller(clusters = cmo$Phasel, sample = cmo$Samples_lv, 
                                     group = cmo$Conditions_lv, transform = "logit")

# --- Arcsine Transformation ---
daa$cmo_groups_asin <- propeller(clusters = cmo$cl_1, sample = cmo$Samples_lv, 
                                 group = cmo$Conditions_lv, transform = "asin")
daa$cmo_clusters_asin <- propeller(clusters = cmo$cl_2, sample = cmo$Samples_lv, 
                                   group = cmo$Conditions_lv, transform = "asin")
daa$cmo_subclusters_asin <- propeller(clusters = cmo$cl_3, sample = cmo$Samples_lv, 
                                      group = cmo$Conditions_lv, transform = "asin")
daa$cmo_cellcycle_asin <- propeller(clusters = cmo$Phasel, sample = cmo$Samples_lv, 
                                    group = cmo$Conditions_lv, transform = "asin")
```

```{r}
# ==============================================================================
# Display Results
# ==============================================================================

# View results for each annotation level
daa$cmo_groups_logit
daa$cmo_groups_asin
daa$cmo_clusters_logit
daa$cmo_clusters_asin
daa$cmo_subclusters_logit
daa$cmo_subclusters_asin
```

# ==============================================================================
# SECTION 15: COMBINE RESULTS FROM ALL ANNOTATION LEVELS
# ==============================================================================

```{r}
# ==============================================================================
# Store Level Orders for Consistent Visualization
# ==============================================================================

lv_subclusters <- levels(cmo$cl_3)  # Subcluster names
lv_clusters <- levels(cmo$cl_2)      # Cluster names
lv_groups <- levels(cmo$cl_1)        # Group names
lv_cellcycle <- levels(cmo$Phasel)   # Cell cycle phases
```

## merging results of different levels to one 

```{r}
# ==============================================================================
# Merge All Results into Single Data Frame
# ==============================================================================

# rbind(): Row-bind data frames from different annotation levels
# Creates unified table for visualization
daa$cmo_all <- rbind(daa$cmo_groups_logit, daa$cmo_clusters_logit, 
                     daa$cmo_subclusters_logit, daa$cmo_cellcycle_logit)
```

```{r}
# View merged results
daa$cmo_all
```

```{r}
# ==============================================================================
# Create Combined Level Order
# ==============================================================================

# Concatenate all level names for consistent plot ordering
lv_all <- c(lv_groups, lv_clusters, lv_subclusters, lv_cellcycle)
```

```{r}
# View combined levels
lv_all
```

```{r}
# Check for any duplicate names across levels
duplicated(lv_all)
```

# ==============================================================================
# SECTION 16: ATF3 KO DIFFERENCE BAR PLOT
# ==============================================================================

```{r,fig.width=20, fig.height=4.8}
# ==============================================================================
# Create Relative Difference Bar Plot - Atf3 KO vs Control
# ==============================================================================

# Calculate normalized proportion differences between conditions
# Positive = higher in KO, Negative = higher in Control

difference_data <- daa$cmo_all %>%
  # Add observation names from row names
  mutate(Observation = rownames(daa$cmo_all)) %>%
  # Select relevant columns for Atf3 comparison
  select(Observation, Control = PropMean.Control, Atf3KO = PropMean.Atf3KO) %>%
  mutate(
    # Calculate total for normalization
    Total = Control + Atf3KO,
    # Normalize proportions to sum to 1
    Control = Control / Total,
    Atf3KO = Atf3KO / Total
  ) %>%
  select(-Total) %>%
  # Reshape for calculation
  pivot_longer(cols = c(Control, Atf3KO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    # Calculate difference (positive = KO higher)
    Difference = Atf3KO - Control,
    # Apply custom ordering
    Observation = factor(Observation, levels = lv_all)
  )

# Define divider positions between annotation levels
divider_positions <- c(3, 18, 58)

# ==============================================================================
# Create Bar Plot
# ==============================================================================

ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  # geom_bar with stat="identity": Use pre-calculated values
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  # Add vertical dividers between annotation levels
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, 
             color = "grey20", size = 0.8) +
  # Colorblind-friendly colors from viridis
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]),
    name = "Comparison",
    labels = c("Control Higher", "Atf3-KO Higher")
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Atf3-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL,
    y = "Difference (Atf3KO - Control)"
  ) +
  theme_classic(base_size = 17) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"),
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12),
    legend.position = c(1, 1.27),
    legend.justification = c("right", "top"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank()
  ) +
  guides(fill = guide_legend(reverse = TRUE))
```

```{r}
# ==============================================================================
# Export Atf3 Difference Plot
# ==============================================================================

# ggsave(): Save plot in multiple formats
# width/height: Plot dimensions in inches
# dpi = 600: High resolution for publication

ggsave("./atf3_new_propeller.png", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./atf3_new_propeller.pdf", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./atf3_new_propeller.svg", width = 20, height = 5, units = "in", dpi = 600)
```

# ==============================================================================
# SECTION 17: ZFP711 KO DIFFERENCE BAR PLOT
# ==============================================================================

```{r,fig.width=20, fig.height=4.8}
# ==============================================================================
# Create Relative Difference Bar Plot - Zfp711 KO vs Control
# ==============================================================================

# Same analysis as Atf3, but for Zfp711 knockout
difference_data <- daa$cmo_all %>%
  mutate(Observation = rownames(daa$cmo_all)) %>%
  # Select Zfp711 columns
  select(Observation, Control = PropMean.Control, Zfp711KO = PropMean.Zfp711KO) %>%
  mutate(
    Total = Control + Zfp711KO,
    Control = Control / Total,
    Zfp711KO = Zfp711KO / Total
  ) %>%
  select(-Total) %>%
  pivot_longer(cols = c(Control, Zfp711KO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    Difference = Zfp711KO - Control,
    Observation = factor(Observation, levels = lv_all)
  )

divider_positions <- c(3, 18, 58)

ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, 
             color = "grey20", size = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]),
    name = "Comparison",
    labels = c("Control Higher", "Zfp711-KO Higher")
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Zfp711-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL,
    y = "Difference (Zfp711KO - Control)"
  ) +
  theme_classic(base_size = 17) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"),
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12),
    legend.position = c(1, 1.27),
    legend.justification = c("right", "top"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank()
  ) +
  guides(fill = guide_legend(reverse = TRUE))
```

```{r}
# View cluster-level results
daa$cmo_clusters_logit
```



```{r}
# ==============================================================================
# Export Zfp711 Difference Plot
# ==============================================================================

ggsave("./zfp711_new_propeller.png", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./zfp711_new_propeller.pdf", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./zfp711_new_propeller.svg", width = 20, height = 5, units = "in", dpi = 600)
```

# ==============================================================================
# SECTION 18: BCL6B KO DIFFERENCE BAR PLOT
# ==============================================================================

```{r,fig.width=20, fig.height=4.8}
# ==============================================================================
# Create Relative Difference Bar Plot - Bcl6b KO vs Control
# ==============================================================================

# Same analysis for Bcl6b knockout
difference_data <- daa$cmo_all %>%
  mutate(Observation = rownames(daa$cmo_all)) %>%
  # Select Bcl6b columns
  select(Observation, Control = PropMean.Control, Bcl6bKO = PropMean.Bcl6bKO) %>%
  mutate(
    Total = Control + Bcl6bKO,
    Control = Control / Total,
    Bcl6bKO = Bcl6bKO / Total
  ) %>%
  select(-Total) %>%
  pivot_longer(cols = c(Control, Bcl6bKO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    Difference = Bcl6bKO - Control,
    Observation = factor(Observation, levels = lv_all)
  )

divider_positions <- c(3, 18, 58)

ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, 
             color = "grey20", size = 0.8) +
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]),
    name = "Comparison",
    labels = c("Control Higher", "Bcl6b-KO Higher")
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Bcl6b-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL,
    y = "Difference (Bcl6bKO - Control)"
  ) +
  theme_classic(base_size = 17) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"),
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12),
    legend.position = c(1, 1.27),
    legend.justification = c("right", "top"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank()
  ) +
  guides(fill = guide_legend(reverse = TRUE))
```

```{r}
# ==============================================================================
# Export Bcl6b Difference Plot
# ==============================================================================

ggsave("./bcl6b_new_propeller.png", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./bcl6b_new_propeller.pdf", width = 20, height = 5, units = "in", dpi = 600)
ggsave("./bcl6b_new_propeller.svg", width = 20, height = 5, units = "in", dpi = 600)
```

# ==============================================================================
# SECTION 19: EXPORT ALL RESULTS
# ==============================================================================

# saving 

```{r}
# ==============================================================================
# Export All DAA Results to Excel
# ==============================================================================

library(writexl)

# write_xlsx(): Export list of data frames to Excel
# Each list element becomes a separate sheet
write_xlsx(daa, path = "./daa_all.xlsx")
```

# ==============================================================================
# SECTION 20: SESSION INFORMATION
# ==============================================================================

```{r}
# ==============================================================================
# Record Session Information for Reproducibility
# ==============================================================================

# sessionInfo(): Displays R version, platform, and all loaded packages
sessionInfo()
```
