---
title: "Differential Abundance Analysis with Speckle"
author: "Ridvan Cetin"
date: "`r Sys.Date()`"
output: html_notebook
---

DISCLAMIR:
I am summarizing my code here, so there might be typos or copy paste errors. This is more like a detailed index page.


```{r eval=FALSE,}
r3dcol <- list()
r3dcol$cols_94 <- c(
  "#F0E442","#1CE6FF","#FF34FF","#FF4A46","#008941","#006FA6","#A30059","#FFDBE5","#7A4900","#0089A3",
  "#63FFAC","#B79762",
  "#004D43","#8FB0FF","#997D87","#809693","#6A3A4C","#b4b701","#4FC601","#3B5DFF","#FF2F80","#61615A",
  "#BA0900","#6B7900",
  "#00C2A0","#FFAA92","#FF90C9","#B903AA","#D16100","#922329","#7B4F4B","#A1C299","#0AA6D8","#00846F",
  "#FFB500","#C2FFED",
  "#A079BF","#CC0744","#C0B9B2","#C2FF99","#00489C","#6F0062","#0CBD66","#EEC3FF","#456D75","#B77B68",
  "#7A87A1","#788D66",
  "#885578","#FAD09F","#FF8A9A","#D157A0","#BEC459","#456648","#0086ED","#886F4C","#34362D","#B4A8BD",
  "#00A6AA","#452C2C",
  "#636375","#A3C8C9","#FF913F","#938A81","#575329","#00FECF","#B05B6F","#8CD0FF","#3B9700","#04F757",
  "#C8A1A1","#1E6E00",
  "#7900D7","#A77500","#6367A9","#A05837","#6B002C","#772600","#D790FF","#9B9700","#549E79","#FFF69F",
  "#201625","#72418F",
  "#BC23FF","#99ADC0","#3A2465","#DDEFFF","#5B4534","#FDE8DC","#404E55","#CB7E98","#A4E804","#324E72"
)
r3dcol$cols_46 <- c(
  "#FFB500","#00C2A0","#D157A0","#8CD0FF","#FF4A46","#FDE8DC","#63FFAC","#B79762","#8FB0FF","#997D87",
  "#809693","#b4b701",
  "#4FC601","#FFAA92","#FF90C9","#A1C299","#F0E442","#C2FFED","#A079BF","#C0B9B2","#C2FF99","#EEC3FF",
  "#B77B68","#7A87A1",
  "#788D66","#FAD09F","#FF8A9A","#BEC459","#B4A8BD","#A3C8C9","#FF913F","#938A81","#00FECF","#FF34FF",
  "#1CE6FF","#04F757",
  "#C8A1A1","#D790FF","#9B9700","#549E79","#FFF69F","#99ADC0","#DDEFFF","#FFDBE5","#CB7E98","#A4E804"
)
r3dcol$cols_pjsala <- c(
  "#005579","#0F4A9C","#139992","#1A1A1A","#354E23","#3F84AA","#532C8A","#635547","#647a4f","#65A83E",
  "#7F6874","#8870ad",
  "#8DB5CE","#8EC792","#989898","#9e6762","#B51D8D","#c19f70","#C3C388","#C594BF","#C72228","#c9a997",
  "#C9EBFB","#cc7818",
  "#CDE088","#DABE99","#DFCDE4","#EF4E22","#EF5A9D","#F397C0","#F6BFCB","#f79083","#f7f79e","#f9decf",
  "#FACB12","#FBBE92",
  "#ff891c"
)
r3dcol$cols_4x3 <- c(
  "#551153","#970098","#9c24cd","#412c00","#997600","#c6a000","#1e3604","#355e07","#4f890e","#002f64",
  "#0050ae","#1077f3"
)
r3dcol$cols_12x1 <- c(
  '#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a',
  '#ffff99','#b15928'
)
r3dcol$cols_4x3b <- c(
  "#551153","#970098","#cc34cd","#c85b00","#f98517","#fdbf6f","#00603d","#008c5c","#33b983","#002f64",
  "#0050ae","#1077f3"
)
r3dcol$cols_4x1 <- c(
  "#970098","#997600","#4f890e","#0050ae"
)
r3dcol$cols_4x1b <- c(
  "#970098","#f98517","#008c5c","#1077f3"
)
r3dcol$cols_3a <- c(
  "#D157A0","#FFB500","#00C2A0"
)
r3dcol$cols_3b <- c(
  "#D454B5","#c85b00","#008c5c"
)
```

```{r eval=FALSE}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(miloR)
  library(dplyr)
  library(scater)
  library(speckle)
  library(ggplot2)
})
```

# Introduction

Here we will summarize the steps of the Compositional Analysis / Differential Abundance Analysis between KO clones and
NT-Control (NonTargetting Control). We will do it in two different analysis 1) cluster levels with *Speckle* and 2)
k-nearstneighborhood with smaller groups of cells.

# Data Preprocessing

## Loading Data


```{r eval=FALSE}
cmo <- readRDS("~/cmo_chapter/introduction/rds/cmoy2.RDS")
Idents(cmo) <- "Conditions_lv"

# subsetting with each KO and Control
cmo_CA <- subset(cmo, idents =c("Control","Atf3KO"))
cmo_BA <- subset(cmo, idents =c("Control","Bcl6bKO"))
cmo_ZA <- subset(cmo, idents =c("Control","Zfp711KO"))

# we have to clean the levels for with following code
clean_metadata <- function(seurat_obj) {
  metadata <- seurat_obj@meta.data
  # Apply droplevels to each factor column
  metadata[] <- lapply(metadata, function(column) {
    if (is.factor(column)) {
      return(droplevels(column))
    } else {
      return(column)
    }
  })
  seurat_obj@meta.data <- metadata
  return(seurat_obj)
}

cmo_CA <- clean_metadata(cmo_CA)
cmo_BA <- clean_metadata(cmo_BA)
cmo_ZA <- clean_metadata(cmo_ZA)
```

I will show only one condition (Atf3-KO versus NT-Control).
We will first follow the classical "speckle" pipeline during the analysis. Later I will subset the samples to equal cell numbers and use the plotCellTypeProps function to compare KO and NT conditions at annotation level. 


# Atf3

## Loading the RDS file

```{r}
cmo_CA <- readRDS("../../../RDS/CMO/cmo_CA.RDS")
```

## just to be sure
if you are subsetting the data and loosing some of the levels use droplevels() before running speckle and miloR
e.g.
```{r}
cmo_CA$Conditions_lv <- droplevels(cmo_CA$Conditions_lv)
```



```{r}
c_ca_4 <- levels(cmo_CA$cl_4) # ordered subcluster-cell cycle annotation (jus t for fun)
c_ca_3 <- levels(cmo_CA$cl_3) # ordered subcluster annotation
c_ca_2 <- levels(cmo_CA$cl_2) # ordered cluster annotation
c_ca_1 <- levels(cmo_CA$cl_1) #  ordered group annotation
c_ca_cc <- levels(cmo_CA$Phasel) # ordered cell cycle annotation
c_ca_conditions <- levels(cmo_CA$Conditions_lv)  # Conditions (Atf3-KO and NT-Control)
c_ca_samples <- levels(cmo_CA$Samples_lv) # Replicates
```

```{r}
Idents(cmo_CA) <- "cl_3"
```

# Propeller

we will do DAA the test now in 5 levels (cl_1, ...) in both "asin" and "logit" tranformation. You can find the results in the supplementary table.


### Asin&logit



```{r}
prop_CA <- list()
prop_CA$S1a <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S2a <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S3a <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S4a <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
prop_CA$S5a<- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")

prop_CA$S1l <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S2l <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S3l <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S4l <- propeller(clusters = cmo_CA$cl_4, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
prop_CA$S5l<- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")

```


```{r}
ca1_asin <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
ca2_asin <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
ca3_asin <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
cacc_asin <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "asin")
```

```{r}
ca1_logit <- propeller(clusters = cmo_CA$cl_1, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
ca2_logit <- propeller(clusters = cmo_CA$cl_2, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
ca3_logit <- propeller(clusters = cmo_CA$cl_3, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
cacc_logit <- propeller(clusters = cmo_CA$Phasel, sample = cmo_CA$Samples_lv, group = cmo_CA$Conditions_lv,transform = "logit")
```


```{r}
write.table(ca1_asin, "./Atf3/Atf3_propeller_groups_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_asin, "./Atf3/Atf3_propeller_clusters_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_asin, "./Atf3/Atf3_propeller_subclusters_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_asin, "./Atf3/Atf3_propeller_cellcycle_asin.txt",sep = '\t', row.names = F, col.names = T, quote = F)
```

```{r}
write.table(ca1_logit, "./Atf3/Atf3_propeller_groups_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca2_logit, "./Atf3/Atf3_propeller_clusters_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(ca3_logit, "./Atf3/Atf3_propeller_subclusters_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
write.table(cacc_logit, "./Atf3/Atf3_propeller_cellcycle_logit.txt",sep = '\t', row.names = F, col.names = T, quote = F)
```


```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_1 <-
  plotCellTypeProps(clusters = cmo_CA$cl_1, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Groups", 
    values = c("#66c2a5","#fc8d62","#8da0cb"),
    guide = guide_legend(ncol = 1)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_1)
```

```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_2 <-
  plotCellTypeProps(clusters = cmo_CA$cl_2, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Clusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 1)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_2)
```

```{r,fig.width=6.5, fig.height=5.5}
plot_cell_type_normal_3 <-
  plotCellTypeProps(clusters = cmo_CA$cl_3, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "Subclusters (A-Z)", 
    values = r3dcol$cols_46,
    guide = guide_legend(ncol = 2)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_3)
```

```{r,fig.width=14, fig.height=5.5}
plot_cell_type_normal_4 <-
  plotCellTypeProps(clusters = cmo_CA$cl_4, sample = cmo_CA$Conditions_lv) +
  theme_classic() + 
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_blank(),  # Removing x-axis title since xlab is set to ""
    axis.title.y = element_blank(),  # Removing y-axis title since ylab is set to ""
    legend.text = element_text(size = 9, margin = margin(l=1)),
    legend.title = element_text(size = 10, hjust = 0.5),
    legend.position = "right"  ,
    legend.box.margin = margin(l=-10, unit = "pt")
  ) + 
  scale_x_discrete(limits = c_ca_conditions) +
  scale_fill_manual(
    "subclusters-cc (A-Z)", 
    values = c(r3dcol$cols_94,r3dcol$cols_46),
    guide = guide_legend(ncol = 6)  # Ensure the legend is in one column
  )+ geom_vline(xintercept = c(3.5,6.5,9.5), linewidth=0.2,color="black")

print(plot_cell_type_normal_4)
```
# We will create one to one comparison plots

Doing some naming adjustment, (Naive and PGC has same naming in clusters and subclusters)


```{r}
cmo <- readRDS("~/PhD_Projects/CMO/cmoy2_reup.RDS")
```

```{r}
cmo
```
```{r}
levels(cmo$cl_3)
lvx <- levels(cmo$cl_3)
```

```{r}
Idents(cmo) <- "cl_3"
```


```{r}
new.cluster.ids <- c("Naiv", lvx[2:6],"PGCs",lvx[-c(1:7)])
new.cluster.ids

```

```{r}
names(new.cluster.ids) <- levels(cmo)
cmo <- RenameIdents(cmo, new.cluster.ids)
```
```{r}
table(Idents(cmo))
```
```{r}
cmo$cl_3 <- Idents(cmo)
```


```{r}
library(speckle)
```
```{r}
daa <- list()
```


```{r}
daa$cmo_groups_logit <- propeller(clusters=cmo$cl_1, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "logit")
daa$cmo_groups_asin <- propeller(clusters=cmo$cl_1, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "asin")
daa$cmo_clusters_logit <- propeller(clusters=cmo$cl_2, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "logit")
daa$cmo_clusters_asin <- propeller(clusters=cmo$cl_2, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "asin")
daa$cmo_subclusters_logit <- propeller(clusters=cmo$cl_3, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "logit")
daa$cmo_subclusters_asin <- propeller(clusters=cmo$cl_3, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "asin")
daa$cmo_cellcycle_logit <- propeller(clusters=cmo$Phasel, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "logit")
daa$cmo_cellcycle_asin <- propeller(clusters=cmo$Phasel, sample=cmo$Samples_lv, group=cmo$Conditions_lv, transform = "asin")
```

```{r}
daa$cmo_groups_logit
daa$cmo_groups_asin
daa$cmo_clusters_logit
daa$cmo_clusters_asin
daa$cmo_subclusters_logit
daa$cmo_subclusters_asin
```

```{r}
lv_subclusters <- levels(cmo$cl_3)
lv_clusters <- levels(cmo$cl_2)
lv_groups <- levels(cmo$cl_1)
lv_cellcycle <- levels(cmo$Phasel)
```

## merging results of different levels to one 

```{r}
daa$cmo_all <- rbind(daa$cmo_groups_logit, daa$cmo_clusters_logit,daa$cmo_subclusters_logit, daa$cmo_cellcycle_logit)
```

```{r}
daa$cmo_all
```

```{r}
lv_all <- c(lv_groups,lv_clusters, lv_subclusters, lv_cellcycle)
```
```{r}
lv_all
```

```{r}
duplicated(lv_all)
```


```{r,fig.width=20, fig.height=4.8}
# Start with the initial dataframe and add observation labels from row names
difference_data <- daa$cmo_all %>%
  mutate(Observation = rownames(daa$cmo_all)) %>%
  select(Observation, Control = PropMean.Control, Atf3KO = PropMean.Atf3KO) %>% # Ensure correct column selection
  mutate(
    Total = Control + Atf3KO,           # Calculate total proportion per observation
    Control = Control / Total,          # Normalize Control
    Atf3KO = Atf3KO / Total             # Normalize Atf3KO
  ) %>%
  select(-Total) %>%
  pivot_longer(cols = c(Control, Atf3KO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    Difference = Atf3KO - Control,      # Calculate difference as Atf3KO - Control
    Observation = factor(Observation, levels = lv_all) # Apply custom order
  )

# Positions for dividers (as indices in lv_all)
divider_positions <- c(3, 18, 58)

# Create the bar plot with proper ordering, dividers, and legend inside
ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") + # Add black border for clarity
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, color = "grey20", size = 0.8) + # Add dividers
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]), # Colorblind-friendly colors
    name = "Comparison",
    labels = c("Control Higher", "Atf3-KO Higher") # Adjusted labels
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Atf3-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL, # Remove the x-axis title
    y = "Difference (Atf3KO - Control)"
  ) +
  theme_classic(base_size = 17) + # Increase base font size for better readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"), # Rotate and bold labels
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16), # Center the title
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12), # Add subtitle
    legend.position = c(1, 1.27), # Top-right inside the plot area
    legend.justification = c("right", "top"), # Align legend to the top-right corner
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank() # Add white background and border
  ) +
  guides(fill = guide_legend(reverse = TRUE)) # Reverse legend order
```
```{r}
ggsave("./atf3_new_propeller.png",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./atf3_new_propeller.pdf",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./atf3_new_propeller.svg",width = 20, height = 5,units = "in",dpi = 600)
```



```{r,fig.width=20, fig.height=4.8}
# Start with the initial dataframe and add observation labels from row names
difference_data <- daa$cmo_all %>%
  mutate(Observation = rownames(daa$cmo_all)) %>%
  select(Observation, Control = PropMean.Control, Zfp711KO = PropMean.Zfp711KO) %>% # Ensure correct column selection
  mutate(
    Total = Control + Zfp711KO,           # Calculate total proportion per observation
    Control = Control / Total,          # Normalize Control
    Zfp711KO = Zfp711KO / Total             # Normalize Zfp711KO
  ) %>%
  select(-Total) %>%
  pivot_longer(cols = c(Control, Zfp711KO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    Difference = Zfp711KO - Control,      # Calculate difference as Zfp711KO - Control
    Observation = factor(Observation, levels = lv_all) # Apply custom order
  )

# Positions for dividers (as indices in lv_all)
divider_positions <- c(3, 18, 58)

# Create the bar plot with proper ordering, dividers, and legend inside
ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") + # Add black border for clarity
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, color = "grey20", size = 0.8) + # Add dividers
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]), # Colorblind-friendly colors
    name = "Comparison",
    labels = c("Control Higher", "Zfp711-KO Higher") # Adjusted labels
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Zfp711-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL, # Remove the x-axis title
    y = "Difference (Zfp711KO - Control)"
  ) +
  theme_classic(base_size = 17) + # Increase base font size for better readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"), # Rotate and bold labels
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16), # Center the title
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12), # Add subtitle
    legend.position = c(1, 1.27), # Top-right inside the plot area
    legend.justification = c("right", "top"), # Align legend to the top-right corner
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank() # Add white background and border
  ) +
  guides(fill = guide_legend(reverse = TRUE)) # Reverse legend order
```

```{r}
daa$cmo_clusters_logit
```



```{r}
ggsave("./zfp711_new_propeller.png",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./zfp711_new_propeller.pdf",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./zfp711_new_propeller.svg",width = 20, height = 5,units = "in",dpi = 600)
```



```{r,fig.width=20, fig.height=4.8}
# Start with the initial dataframe and add observation labels from row names
difference_data <- daa$cmo_all %>%
  mutate(Observation = rownames(daa$cmo_all)) %>%
  select(Observation, Control = PropMean.Control, Bcl6bKO = PropMean.Bcl6bKO) %>% # Ensure correct column selection
  mutate(
    Total = Control + Bcl6bKO,           # Calculate total proportion per observation
    Control = Control / Total,          # Normalize Control
    Bcl6bKO = Bcl6bKO / Total             # Normalize Bcl6bKO
  ) %>%
  select(-Total) %>%
  pivot_longer(cols = c(Control, Bcl6bKO), 
               names_to = "Group", 
               values_to = "Proportion") %>%
  pivot_wider(names_from = Group, values_from = Proportion) %>%
  mutate(
    Difference = Bcl6bKO - Control,      # Calculate difference as Bcl6bKO - Control
    Observation = factor(Observation, levels = lv_all) # Apply custom order
  )

# Positions for dividers (as indices in lv_all)
divider_positions <- c(3, 18, 58)

# Create the bar plot with proper ordering, dividers, and legend inside
ggplot(difference_data, aes(x = Observation, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") + # Add black border for clarity
  geom_vline(xintercept = divider_positions + 0.5, linetype = 1, color = "grey20", size = 0.8) + # Add dividers
  scale_fill_manual(
    values = c("TRUE" = viridis(5)[5], "FALSE" = viridis(5)[1]), # Colorblind-friendly colors
    name = "Comparison",
    labels = c("Control Higher", "Bcl6b-KO Higher") # Adjusted labels
  ) +
  labs(
    title = "Relative Difference Between Cell Type Abundance of Bcl6b-KO and Control",
    subtitle = "Normalized proportions across cell type annotations",
    x = NULL, # Remove the x-axis title
    y = "Difference (Bcl6bKO - Control)"
  ) +
  theme_classic(base_size = 17) + # Increase base font size for better readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "plain"), # Rotate and bold labels
    axis.text.y = element_text(face = "plain"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16), # Center the title
    plot.subtitle = element_text(hjust = 0.5, face = "italic", size = 12), # Add subtitle
    legend.position = c(1, 1.27), # Top-right inside the plot area
    legend.justification = c("right", "top"), # Align legend to the top-right corner
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.background = element_blank() # Add white background and border
  ) +
  guides(fill = guide_legend(reverse = TRUE)) # Reverse legend order
```

```{r}
ggsave("./bcl6b_new_propeller.png",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./bcl6b_new_propeller.pdf",width = 20, height = 5,units = "in",dpi = 600)
ggsave("./bcl6b_new_propeller.svg",width = 20, height = 5,units = "in",dpi = 600)
```

# saving 

```{r}
library(writexl)
# Save to Excel
write_xlsx(daa, path = "./daa_all.xlsx")
```

```{r}
sessionInfo()
```













